---
title: "m6A-mediated decay: analsis of human data"
editor: visual
execute:
  eval: false
format:
  html:
    code-fold: true
    code-tools: true
    embed-resources: true
toc: true
---

# Prerequisites

Set up the workflow.

```{r setup}

############
analysisID = "m6A_decay_tRNA"
############

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(Hmisc)
  library(readxl)
  library(doParallel)
  library(stringr)
  library(zoo)
  library(yaml)
  library(purrr)
  library(digest)
  library(dineq)
  
  library(rtracklayer)
  library(seqinr)
  library(biomaRt)
  library(Rsamtools)
  library(Biostrings)
  library(GenomicFeatures)
  library(GenomicAlignments)
  library(hiAnnotator)
  library(Rsubread)
  library(ProliferativeIndex)
  library(msigdbr)

  
  #library(colorspace)
  library(RColorBrewer)
  library(ggpubr)
  library(ggrepel)
  library(ggnewscale)
  library(ggbeeswarm)
  library(ggpattern)
  #library(scales)
  #library(GGally)
  library(ggh4x)
  library(ggpmisc)

})

for(file in list.files("../R_codebase/", "*.R$", full.names = TRUE)){source(file)}

###
### genome setup
###

genomeDir = "~/genomes/homo_sapiens_GRCh38_v35"
fastaFile = file.path(genomeDir, "GRCh38.primary_assembly.genome.fa.gz")
txFile = file.path(genomeDir, "gencode.v35.annotation.gtf.gz")

if(!file.exists(fastaFile)){
  dir.create(dirname(fastaFile), recursive = T)
  download.file(url = "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_35/GRCh38.primary_assembly.genome.fa.gz", destfile = fastaFile, method = "wget")
}

if(!file.exists(txFile)){
  dir.create(dirname(txFile), recursive = T)
  download.file(url = "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_35/gencode.v35.annotation.gtf.gz", destfile = txFile)
}

###
### output setup
###

tmpDir = file.path("/scratch/linder", analysisID)
if(!dir.exists(tmpDir)){dir.create(tmpDir, recursive = T)}

```

```{r parameters}

exlenbins = 17
decay_shift_ntiles = 100/5

```

```{r definitions}

drach_codons <- c("AAA", "GAA", "TAA", "AGA", "GGA", "TGA", "GAC", "AAC", "ACA", "ACC", "ACT")
stop_codons <- c("TGA", "TAA", "TAG")
mcm5s2U_codons   = c("CAA", "AAA", "AGA", "GAA")
elongator_codons = c("TTA", "GTA", "TCA", "CCA", "ACA", "GCA", "CAA", "AAA", "GAA", "AGA", "GGA") # from Johansson, Byström (2018). BBA 1861, 401–408.
queuosine_codons = c("GAC", "AAC", "TAC", "CAC", "GAT", "AAT", "TAT", "CAT")

```

```{r palettes}

palval_m6A_region = c(none = "grey", utr5 = "#FFCC00", cds = "#669966", utr3 = "#993366", both = "grey80") 

palval_m6A_region_fill         = c(none = "grey", utr5 = "#FFCC00", cds = "#669966", utr3 = "#993366", cdsutr3 = "#669966")
palval_m6A_region_fill_pattern = c(none = NA, utr5 = NA, cds = NA, utr3 = NA, cdsutr3 = "#993366")

palval_exbin  = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(exlenbins) ## diverging palette

palval_drach_status = c("all" = "black", "other" = "grey", "no DRACH" = "#4575B4", "DRACH" = "orange", "m6A" = "#D73027")


palval_reporter = c("none" = "grey40", "C1" = "#9A3367", "C2" = "#669965", "C3" = "#00a19a", "C4" = "#36a9e1", "C5" = "#fcdd2e", "C6" = "#00a19a", "C7" = "#6691cb" )
palval_reporter_codons = palval_reporter
names(palval_reporter_codons) = sedit(names(palval_reporter_codons), c("C2", "C5"), c("AGA", "GGA"))

palval_genotype <-
  c(  "WT"      = "grey40"
    , "ELP1 KO" = "#8DCA46"   # "#4b8bc0"
    , "CTU2 KO" = "#BC6CA7"   # "#842976"   #"#ffd95c"   # "#FF206E" 
    ) 

palval_m6A_inh = c("mock" = "grey40", "STM" = "orange")

palval_transl_inh = c("mock" = "grey40", "EM" = "#00cccc", "CHX" = "#00afcc")

palval_sets = c(m6A     = "#FFDE0D",
                m6A_neg = "grey80",
                mcm5    = "grey60",
                s2U     = "#6E94CD",
                ngd     = "blue",
                yth     = "orange",
                proliferation = "#F8766D",
                ribosomal = "#F8766D"
                )

palval_queosine <-
  c("WT" = "grey40",
    "QTRT1 KO" = "#e3e3ff",
    "QTRT2 KO" = "#c2c2ff",
    "QTGAL KO" = "#ffc7c7",
    "QTMAN KO" = "#cbe5c3")

```

```{r labellers}

labeller_reporter = c("none" = "untransfected", "C4" = "no~m^6*A", "C7" = "CAA", "C2" = "CDS", "C6" = "GAA", "C5" = "GGA", "C1" = "`3'UTR`" ) ## "#00a19a" 

labeller_reporter_codons = labeller_reporter
labeller_reporter_codons["C2"] = "AGA"

```

```{r functions}

get_codon_strength <-
  function(codon){
    ## from: Grosjean & Westhof 2016
    strength = rep(NA, length(codon))
    strength = ifelse(grepl("[GC][GC].", codon), "strong", strength)
    strength = ifelse(grepl("([GC][AT].)|([AT][GC].)", codon), "intermediate", strength)
    strength = ifelse(grepl("[AT][AT].", codon), "weak", strength)
    strength = factor(strength, levels = c("weak", "intermediate", "strong"))
    return(strength)
  }

get_codon_tRNA_mod <-
  function(codon){
    mod = rep("other", length(codon))
    mod[codon %in% c("ACA", "CCA", "GCA", "GGA", "GTA", "TCA", "TTA")] = "mcm5U"
    mod[codon %in% c("AAA", "GAA", "AGA", "CAA")] = "mcm5s2U"
    mod[codon %in% c("TAA", "TAG", "TGA")] = "stop"
    mod = factor(mod, levels = c("other", "mcm5U", "mcm5s2U", "stop"))
    return(mod)
  }

saf_from_grl <-
  function(grl){
    gr = unlist(grl)
    saf = data.table(GeneID = as.character(names(gr)),
                     Chr = as.character(seqnames(gr)),
                     Start = as.integer(start(gr)),
                     End = as.integer(end(gr)),
                     Strand = as.character(strand(gr)) )
    return(saf)
    
  }


exbinMarkerBar <-
  function(m6A_fill_levels = 17){
    list(
    new_scale_fill(),
    geom_rect(aes(xmin = as.integer(exbin)-0.5,
                  xmax = as.integer(exbin)+0.5,
                  ymin = min(..y..) - ((max(..y..)-min(..y..))*0.05)*2.5,
                  ymax = min(..y..) - (max(..y..)-min(..y..))*0.05,
                  fill = exbin),
              col = NA, show.legend = FALSE),
    scale_fill_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(m6A_fill_levels))
    )
  }

geom_decay_line <-
  function(by = NULL) {
    geom_line(data = function(x){
      x[, predict_decay(fit_decay(
        .SD, method = "nls", time_col = "time_num")$coef,
        timepoints = seq(0,4,0.2), time_col = "time_num"),
        by = by]})
  }

geom_decay_text <-
  function(by = NULL, vjust_var = NULL, size = 4) {
    geom_text(data = function(x){
      x[, fit_decay(
        .SD, method = "nls", time_col = "time_num")$coef[, t12 := log(2)/k],
        by = by][, vj := if(is.null(vjust_var)){1}else{as.integer(factor(get(vjust_var)))}]},
      aes(label = paste0(#"t[1/2]==", 
                         round(t12, 1), "*h"), vjust = vj*1.05 ), x = Inf, y = Inf, hjust = 1, parse = TRUE, show.legend = FALSE, size = size)
  }

```

# Transcript model

Generate the transcipt model for convenient analysis.

```{r canonical_transcripts}

canonical_transcripts <- 
  getBM(attributes = c("ensembl_gene_id", "ensembl_transcript_id", "transcript_is_canonical", "hgnc_symbol"),
        filters = "transcript_is_canonical",
        values = TRUE,
        mart = useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", version = 113, verbose = TRUE))

canonical_transcripts = as.data.table(canonical_transcripts)
setnames(canonical_transcripts, "hgnc_symbol", "gene")

head(canonical_transcripts)

```

```{r}
# getBM(attributes = c("ensembl_gene_id", "ensembl_transcript_id", "transcript_is_canonical"),
#         filters = "transcript_is_canonical",
#         values = TRUE,
#         mart = useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl"))
```

```{r gtf}

gtf <-
  import(con = txFile)

gtf_dt <-
  as.data.table(mcols(gtf))[, c("chrom", "start", "end") := .(as.character(seqnames(gtf)), start(gtf), end(gtf)) ]

gtf_dt[, ensembl_transcript_id := gsub("\\..*$", "", transcript_id)]
gtf_dt[, ensembl_gene_id := gsub("\\..*$", "", gene_id)]
gtf_dt[, gene := gene_name]

head(gtf_dt)

```

```{r BSgenome}

bsg  <-
  create_BSgenome(fastaFile = fastaFile, organism = "Homo sapiens", genome = "hg38")

bsg

```

```{r txdb}

txdbFile = file.path(dirname(txFile), paste0(basename(txFile), ".txdb.RData"))
# file.remove(txdbFile)
if (!file.exists(txdbFile)) {
  txdb <-
   makeTxDbFromGFF(txFile, format = "gtf")
  saveDb(x = txdb, file = txdbFile)
} 

txdb <- loadDb(txdbFile)

txdb
```

```{r txmodel}
 
txdatfile = file.path(dirname(txFile), paste0(basename(txFile),".txdat.RData"))
# file.remove(txdatfile)
if(!file.exists(txdatfile)){
cat("Generating txmodel...")
  txdat <-
    get_txdat_from_gtf(txFile)
    save(txdat, file = txdatfile )
    cat(" ...done.\n")
  }else{cat("Loading txmodel..."); load(txdatfile);cat(" ...done.\n") }

txmodel = txdat$txmodel

txmodel[, is_canonical := gsub("\\..*$", "", tx_name) %in% canonical_transcripts$ensembl_transcript_id]

txmodel[! is.na(cds_len) & is_canonical, is_longest_cds := {
    ## pick the longest CDS (in case of equal length, we simply pick the first)
    x = cds_len == max(cds_len, na.rm = TRUE)
    hit = min(which(x))
    x = rep(FALSE, length(x))
    x[hit] = TRUE
    x
  }, by = .(gene_id) ]

txmodel[, is_protein_coding := gene_id %in% gtf_dt[gene_type == "protein_coding", unique(gene_id)] ]

txmodel[, is_QC_pass := utr5_len > 10 & cds_len  > 60 & cds_len < 10000 & utr3_len > 60]

txmodel <-
  txmodel[is_canonical & is_protein_coding & is_longest_cds & is_QC_pass]

cat(
  paste(
    "Filtering for valid transcripts leaves", nrow(txmodel), "out of", nrow(txdat$txmodel), "transcripts.\n"
  )
)

stopifnot(txmodel[, length(gene_id) == length(unique(gene_id))])

txmodel[, ensembl_gene_id := gsub("\\..*$", "", gene_id)]
txmodel[, gene := setkeyv(gtf_dt[, .(ensembl_gene_id, gene)][, unique(.SD)], "ensembl_gene_id")[ensembl_gene_id, gene]]
txmodel[, exlen := unlist(lapply(cds_exon_len, mean, na.rm=TRUE))] ## get mean CDS exon lenth per transcript
txmodel[, exlen_max := unlist(lapply(cds_exon_len, max, na.rm=TRUE))] ## get mean CDS exon lenth per transcript

head(txmodel)

```

```{r txmodel gini}
require(dineq)

txmodel[, cds_exon_length_gini := unlist(txmodel[, lapply(cds_exon_len, gini.wtd)])]

head(txmodel[, .(cds_exon_length_gini = median(cds_exon_length_gini)), by = "gene_id"])

```

```{r ex_by_tx}

ex_by_tx <-
  txdat$ex_by_tx[names(txdat$ex_by_tx) %in% as.character(txmodel$tx_name)]

ex_by_tx

```

```{r exlenTabGene}

exlenTabGene <-
  txmodel[, .(exlen = mean(exlen, na.rm=T),      ## get mean CDS exon length per gene
              exlen_max = max(exlen_max, na.rm = TRUE), 
              tx_len = mean(tx_len, na.rm=T),  ## get mean tx length per gene
              utr5_len = mean(utr5_len, na.rm=T),  ## get mean utr5 length per gene
              cds_len = mean(cds_len, na.rm=T),  ## get mean CDS length per gene
              utr3_len = mean(utr3_len, na.rm=T),  ## get mean utr3 length per gene 
              nexon = mean(nexon, na.rm=T),       ## get mean number of exons per gene
              cds_exon_length_gini = mean(cds_exon_length_gini, na.rm = TRUE)
              ), by=c("gene_id", "ensembl_gene_id", "gene")] 


## remove olfactory receptor genes
exlenTabGene = exlenTabGene[!grepl("^OR[0-9]+", gene)]

exlenTabGene[, exbin := cut2((exlen), g = exlenbins, digits = 1)]
exlenTabGene[, exrank := rank(exlen, ties.method = "random")]
exlenTabGene[, longer_than_median := exlen > median(exlen) ]

ggplot(exlenTabGene, aes(x = exlen)) +
  # geom_density(alpha = 0.3, aes(fill = exbin), col = NA) +
  geom_histogram(bins = 200, alpha = 1, aes(fill = exbin, group=exbin), col = NA) +
  
  geom_vline(xintercept = median(exlenTabGene$exlen, na.rm=T), linetype=2, col = "blue") +
  geom_text(aes(median(exlenTabGene$exlen, na.rm=T), Inf, label = round(median(exlenTabGene$exlen, na.rm=T), digits = 1)), col = "blue", vjust=1, check_overlap = T) +
  
  geom_vline(xintercept = mean(exlenTabGene$exlen, na.rm=T), linetype=2, col = "red") +
  geom_text(aes(mean(exlenTabGene$exlen, na.rm=T), Inf, label = round(mean(exlenTabGene$exlen, na.rm=T), digits = 1)), col = "red", vjust=1, check_overlap = T) +
  
  scale_x_continuous(trans='log10', limits= c(50, 5000)) +
  annotation_logticks(base = 10, sides = "b") +
  
  labs( fill = "CDS exon length bin", y = "Gene count", x = "Gene average CDS exon length [nt]") +
  scale_fill_manual(values = palval_exbin) +
  theme_pubr() +
  guides(fill=guide_legend(ncol=3)) +
  theme(plot.title = element_text(hjust = 0.5), legend.position=c(0.7, 0.6)) +
  NULL

 fwrite(exlenTabGene, "/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/exlen_human.tsv", sep = "\t")

```

```{r eejs.tx}
## Get EEJ coordinates. We use the first nucleotide of all exons that are not the first exon.
ex <-
  unlist(ex_by_tx, use.names = F)

eejs <-
  unique(resize(ex[ex$exon_rank != 1], 1, "start"))

eejs.tx <-
  mapToTranscripts(x = eejs, transcripts = ex_by_tx)
strand(eejs.tx) = "+"

eejs.tx$exon_name <-
  eejs[eejs.tx$xHits]$exon_name

eejs.tx$exlen  <-
  width(ex)[match(eejs.tx$exon_name, ex$exon_name)]

eejs.tx$frame <-
  get_frame(eejs.tx, txmodel)

eejs.tx$relpos.cds <-
  relpos.cds.gr(eejs.tx)

eejs.tx$region <-
  get_region(eejs.tx, txmodel)

eejs.tx

# for QC in IGV
# export(object = eejs, con = gzfile("~/eejs.bed.gz"),format = "bed")
# export(object = ex_by_tx, con = gzfile("~/exbytx.bed.gz"),format = "bed")

#ggplot(as.data.frame(mcols(eejs.tx)), aes(x=frame, fill = factor(exlen %% 3) )) + geom_bar(position="dodge")

```

```{r ex_on_tx}
require(Hmisc)

## Adjust exons to CDS start and stop.
## Generate GRanges of CDS exons in tx space. The start and end coordinates of exons spanning the start and stop codons are adjusted to match those features.

## GRanges of individual exons
ex.gr <-
  unlist(ex_by_tx)
## we only need each exon once and remove exons w/o name (e.g spike-ins)
ex.gr <-
  ex.gr[ ! duplicated(ex.gr$exon_name) & ! is.na(ex.gr$exon_name)] 
## map exons to transcript space
ex_on_tx <-
  mapToTranscripts(ex.gr, ex_by_tx) 
## annotate
mcols(ex_on_tx) <-
  mcols(ex.gr[ex_on_tx$xHits]) 
## test for exons that have not been mapped on transcripts
stopifnot(sum(!ex.gr$exon_name %in% unique(ex_on_tx$exon_name)) == 0)
## as we use txmodel to annotate, remove tx not in txmodel
ex_on_tx <-
  ex_on_tx[seqnames(ex_on_tx) %in% txmodel$tx_name]
## remove exons in UTRs
ex_on_tx <-
  ex_on_tx[ ! start(ex_on_tx) >= txmodel$utr3.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)]]
ex_on_tx <-
  ex_on_tx[ ! end(ex_on_tx) <= txmodel$cds.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)]]

## restrict exon boundaries to CDS
start(ex_on_tx) <-
  ifelse(test = start(ex_on_tx) < txmodel$cds.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)] &  ## on exons spanning the start codon
                  end(ex_on_tx) > txmodel$cds.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)],  
         yes  = txmodel$cds.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)], ## replace start with first nucleotide of CDS
         no   = start(ex_on_tx)) ## else leave exon start

end(ex_on_tx) <-
  ifelse(test =   end(ex_on_tx) > txmodel$utr3.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)] &  ## on exons spanning the stop codon
                start(ex_on_tx) < txmodel$utr3.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)],  
         yes  = txmodel$utr3.start[match(as.character(seqnames(ex_on_tx)), txmodel$tx_name)] -1 , ## replace end with last nucleotide of CDS
         no   = end(ex_on_tx)) ## else leave exon end

## exon length bins
ex_on_tx$exlen <-
  width(ex_on_tx)
ex_on_tx$exbin <-
  cut2(ex_on_tx$exlen, g = exlenbins)

ex_on_tx

## map back to genome space
ex_on_gn <-
  mapFromTranscripts(ex_on_tx, ex_by_tx)
mcols(ex_on_gn) <-
  mcols(ex_on_tx[ex_on_gn$xHits])

```

```{r drach.tx}

## get reference sequence
seqlevelsStyle(bsg) = seqlevelsStyle(ex_by_tx)
tx.seqs = extractTranscriptSeqs(bsg, keepStandardChromosomes(ex_by_tx, pruning.mode = "coarse"))

## get drach motifs
drach.tx <-
  GRanges(vmatchPattern("DRACH", tx.seqs, fixed = F)) - 2 # center on A

## annotate
drach.tx$kmer <-
  getSeq(tx.seqs, drach.tx + 2)
drach.tx$relpos.cds <-
  relpos.cds.gr(drach.tx)
drach.tx$region <-
  get_region(drach.tx, txmodel)
drach.tx$region = factor(drach.tx$region, levels = c("utr5", "cds", "utr3") )
drach.tx$frame <-
  get_frame(drach.tx, txmodel = txmodel)
drach.tx$codon = substr(drach.tx$kmer, 3 - as.numeric(as.character(drach.tx$frame)) , 3 - as.numeric(as.character(drach.tx$frame)) + 2 )

drach.tx

```

```{r cods.gr}

codsID <-
  digest(c(get_codon_gr, txmodel, tx.seqs, eejs.tx, drach.tx, ex_on_tx))

codsFile = file.path(tmpDir, paste0("codon_GRanges.", codsID, ".RData") )

require(doParallel)
registerDoParallel(cores = 6)
system.time(
if(!file.exists(codsFile)){
  cat("Generating cods.gr...")
  cods.gr <- 
    get_codon_gr(txmodel = txmodel, tx.seqs = tx.seqs, eejs.tx = eejs.tx, drach.tx = drach.tx, ex_on_tx = ex_on_tx)
  saveRDS(cods.gr, codsFile)
  cat(" ...done.\n")
}else{cat("Loading cods.gr..."); cods.gr <- readRDS(codsFile); cat(" ...done.\n")}
)
registerDoParallel(cores = 1)

cods.gr$codon_rank_from_stop = as.data.table(cods.gr)[, codon_rank - max(codon_rank), by = .(seqnames)][[2]]

cods.gr$codon_strength <- get_codon_strength(cods.gr$codon)
cods.gr$codon_tRNA_mod <- get_codon_tRNA_mod(cods.gr$codon)

cods.gr

```

```{r}
# saveRDS(cods.gr, "~/exports/data/codon_GRanges.RDS")
# saveRDS(drach.tx, "~/exports/data/DRACH_GRanges.RDS")
```

```{r geneset}

geneset <-
  get_geneset(dict = unique(canonical_transcripts$gene))

geneset

```

```{r msig}

msigDT <-
  as.data.table(msigdbr(species = "Homo sapiens"))
msigDT[, gene := gene_symbol]

msigList <-
  split(x = msigDT$gene, f = msigDT$gs_name)

msigList[["CUSTOM_DIFFERENTIATION"]] <-
  sort(unique(unlist(msigList[grepl("DIFFERENTIAT", names(msigList))])))
  
msigList[["CUSTOM_PROLIFERATION"]] <-
  sort(unique(unlist(msigList[grepl("PROLIFERAT", names(msigList))])))

msigList[["CUSTOM_METABOLISM"]] <-
  sort(unique(unlist(msigList[grepl("METABOLI", names(msigList))])))

## create unambiguous subsets of genes not in the other groups
for(gc in grep("CUSTOM", names(msigList), value = TRUE)) {
  msigList[[gc]] <-
    setdiff(msigList[[gc]], unlist(msigList[setdiff(grep("CUSTOM", names(msigList), value = TRUE), gc) ]))
  }

lapply(msigList[grep("CUSTOM|GOCC_DNA_PACKAGING_COMPLEX", names(msigList))], length)


## geneset categories
gs_signaling = msigDT[grepl("HALLMARK|KEGG", gs_name) & grepl("SIGNALING|RESPONSE", gs_name) & !grepl("_UP$|_DN$|UNFOLDED", gs_name), unique(gs_name) ]
gs_metabolic = msigDT[grepl("HALLMARK|KEGG", gs_name) & grepl("METABOLISM|GLYCOLYSIS|SECRETION|OXYGEN|OXIDATIVE|CHOLESTEROL", gs_name), unique(gs_name) ]
gs_select = grep("TGF|TNF|NOTCH|MYC|OXIDATIVE|SECRETION|FATTY|INFLAMMATORY_RESPONSE", c(gs_signaling, gs_metabolic), value = TRUE)

genes_histones = grep("^H", msigList$GOCC_DNA_PACKAGING_COMPLEX, value = TRUE)

```

# Data

Collect data from various sources.
## --- m6A sites --- 
### miCLIP Read in the BED file of DRACH sites called by the miCLIP pipeline in analysis_miCLIP.Rmd

```{r read c2t}

## read c2t
miCLIPfile = "~/exports/data/miclip_c2t.bed"

c2t <-
  import.bed(miCLIPfile)

c2t$siteID = 1:length(c2t)

c2t$call <-
  gsub("_.*$", "", c2t$name)

c2t$is_m6A <-
  c2t$call == "methylated"

c2t$cvg <-
  as.integer(gsub("cvg","",do.call(rbind, strsplit(c2t$name, split ="_"))[,2]))
  
c2t$nC <-
  as.integer(gsub("c","",do.call(rbind, strsplit(c2t$name, split ="_"))[,3]))

c2t$nT <-
  as.integer(gsub("t","",do.call(rbind, strsplit(c2t$name, split ="_"))[,4]))

c2t$c2t <- c2t$nT / c2t$nC

## map to transcripts
c2t.tx <-
  mapToTranscripts(x = c2t, transcripts =  ex_by_tx)
strand(c2t.tx) = "+"

mcols(c2t.tx) <-
  cbind(mcols(c2t.tx), mcols(c2t)[c2t.tx$xHits,])

c2t.tx$relpos.cds <-
  relpos.cds.gr(c2t.tx)

c2t.tx$frame <-
  get_frame(c2t.tx, txmodel = txmodel)

c2t.tx$region <- 
  factor(get_region(c2t.tx, txmodel), levels = c("utr5", "cds", "utr3"))
  # merge(data.table( tx_name = as.character(seqnames(c2t.tx)), start = start(c2t.tx)), txmodel, "tx_name", all.x = TRUE)[, region := ifelse(test = start < cds.start, yes = "utr5", no = ifelse(test = start >= cds.start & start < utr3.start, yes = "cds", no = ifelse(test = start >= utr3.start, yes = "utr3", no = NA )))]$region

# c2t.tx$framelabel <- 
#   get_framelabel(frame = c2t.tx$frame)

c2t.dt <-
  cbind(data.table("tx_name" = as.character(seqnames(c2t.tx))), as.data.table(mcols(c2t.tx)))

c2t.dt[, ensembl_transcript_id := gsub("\\..*$", "", tx_name)]

c2t.dt[, ensembl_gene_id := setkeyv(gtf_dt[!is.na(ensembl_transcript_id), unique(.SD)], "ensembl_gene_id")[ensembl_transcript_id, ensembl_gene_id]]

head(c2t.dt)

```

```{r}
c2t.dt[(is_m6A), .N, by = "region"]
```

```{r define miCLIP m6A codons}
### annotate with called m6A from integrated miCLIP
cods.gr$is_m6A <-
  suppressWarnings(overlapsAny(cods.gr, c2t.tx[c2t.tx$is_m6A]))

### codon-matched control sites in same CDSs as m6A
cods.gr$matched_status <-
  get_codon_matched_sites(cods.gr = cods.gr)

### status of all DRACH codons in transcriptome (is_m6A, is_drach or not DRACH)
cods.gr$drach_status <-
  get_drach_status(cods.gr = cods.gr, m6A_var = "is_m6A")

cat("miCLIP m6A sites:")
table(cods.gr$is_m6A)
cat("CDS matched status:")
table(cods.gr$matched_status)
cat("DRACH status:")
table(cods.gr$drach_status)

```

```{r lift c2t to codons}

### annotate with miCLIP info
cods.c2t.hits <-
  suppressWarnings(findOverlaps(query = cods.gr, subject = c2t.tx ))
cods.gr$c2t <- as.numeric(NA)
cods.gr[queryHits(cods.c2t.hits)]$c2t <-
  c2t.tx[subjectHits(cods.c2t.hits)]$c2t

cods.gr$cvg <- as.integer(NA)
cods.gr[queryHits(cods.c2t.hits)]$cvg <-
  c2t.tx[subjectHits(cods.c2t.hits)]$cvg

cods.gr$nC<- as.integer(NA)
cods.gr[queryHits(cods.c2t.hits)]$nC <-
  c2t.tx[subjectHits(cods.c2t.hits)]$nC

cods.gr$nT<- as.integer(NA)
cods.gr[queryHits(cods.c2t.hits)]$nT <-
  c2t.tx[subjectHits(cods.c2t.hits)]$nT

head(cods.gr,3)

```

```{r lift c2t to drach}

drach.tx$is_m6A_miCLIP <-
  suppressWarnings(overlapsAny(drach.tx, c2t.tx[c2t.tx$is_m6A]))

### annotate with miCLIP info
drach.c2t.hits <-
  suppressWarnings(findOverlaps(query = drach.tx, subject = c2t.tx ))
drach.tx$c2t <- as.numeric(NA)
drach.tx[queryHits(drach.c2t.hits)]$c2t <-
  c2t.tx[subjectHits(drach.c2t.hits)]$c2t


head(drach.tx,3)

```

## --- m6A levels ---

### GLORI

Liu, C., Sun, H., Yi, Y., Shen, W., Li, K., Xiao, Y., Li, F., Li, Y., Hou, Y., Lu, B., et al. (2022). Absolute quantification of single-base m6A methylation in the mammalian transcriptome using GLORI. Nat Biotechnol, 1–12. 10.1038/s41587-022-01487-9.

```{r}
glori  <-
  within(list(), {
    dataURL = "https://static-content.springer.com/esm/art%3A10.1038%2Fs41587-022-01487-9/MediaObjects/41587_2022_1487_MOESM3_ESM.xlsx"
    dataFile = file.path(tmpDir, basename(dataURL))
    if(!file.exists(dataFile)){download.file(dataURL, dataFile)}
    dt = as.data.table(read_xlsx(dataFile))
    dt[, m6A_lvl := rowMeans(cbind(m6A_level_rep1, m6A_level_rep2))]
    gr = dt[, GRanges(seqnames = Chr, ranges = IRanges(start = Sites, width = 1), strand = Strand, m6A_lvl, score = m6A_lvl)]
    
    ## map to transcripts
    tx <-
      mapToTranscripts(x = gr, transcripts =  ex_by_tx)
    strand(tx) = "+"
    mcols(tx) <-
      cbind(mcols(tx), mcols(gr)[tx$xHits,])
    tx$relpos.cds <-
      relpos.cds.gr(tx)
    tx$frame <-
      get_frame(gr = tx, txmodel = txmodel)
    tx$region <-
      factor(get_region(tx, txmodel), levels = c("utr5", "cds", "utr3"))
    tx$kmer <-
      # get_kmer(gr = trim(tx + 2), tx.seqs)
      getSeq(tx.seqs, trim(tx + 2))
    
  })

glori$dt
glori$gr
glori$tx


#export(glori$gr, "~/exports/data/glori.bed", format = "bed")
#pushFileToOwncube("~/exports/data/glori.bed", "exports/data/glori.bed")

```

```{r lift GLORI to codons}

## define position of m6A in codons to use with GRanges resize()
codon_m6A_pos <-
  c(
    ## DRACH codons
    "AAA" = "end",
    "AGA" = "end",
    "GAA" = "end",
    "GGA" = "end",
    "AAC" = "center",
    "GAC" = "center",
    "ACA" = "start",
    "ACC" = "start",
    "ACT" = "start",
    ## non-DRACH codons
    "ACG" = "start",
    "ATT" = "start",
    "GTA" = "end",
    "CAA" = "end",
    ## stop codons
    "TGA" = "end",
    "TAG" = "center" 
    )

## findOverlaps on codons resized to their m6A position (codons that are not a DRACH codon get resized to width 0)

cods.gr.glori.hits <-
  findOverlaps(query = resize(cods.gr,
                              ifelse(cods.gr$codon %in% names(codon_m6A_pos), 1, 0),
                              ifelse(cods.gr$codon %in% names(codon_m6A_pos), codon_m6A_pos[cods.gr$codon], "start") ),
             subject = glori$tx )
## assign metadata columns
cods.gr$glori_m6A_lvl = as.numeric(0)
cods.gr[queryHits(cods.gr.glori.hits)]$glori_m6A_lvl <-
  glori$tx[subjectHits(cods.gr.glori.hits), ]$m6A_lvl

cods.gr$gloribin = relevel(factor(ifelse(cods.gr$glori_m6A_lvl == 0, "[0]", as.character(cut2(cods.gr$glori_m6A_lvl,cuts = seq(0,1,0.1)))  ) ), "[0]")

#cods.gr$glori_m6A_lvl_stm = as.numeric(NA)
# cods.gr[queryHits(cods.gr.glori.hits)]$glori_m6A_lvl_stm <-
#   glori_stm$tx[subjectHits(cods.gr.glori.hits), ]$m6A_lvl_stm


```

```{r check glori DRACH, fig.height = 6, fig.width = 4}

as.data.table(glori$tx)[
    , is_DRACH := grepl("[AGT][AG]AC[ACT]", kmer)
  ][
    , {cbind(N=.N, mean_se(m6A_lvl))}
    , by = .(is_DRACH)
  ]

as.data.table(glori$tx)[
    , is_DRACH := grepl("[AGT][AG]AC[ACT]", kmer)
  ][
    , {cbind(N=.N, mean_sd(m6A_lvl*100))}
    , by = .(kmer, is_DRACH = ifelse(is_DRACH, "DRACH", "other") )
  ][
    order(-N)
    , 
    ][
      N > 100 & nchar(kmer) == 5
    ][
      ,
      ggarrange(
      ggplot(.SD,
             aes(x = factor(kmer, levels = unique(kmer[order(N)])),
                 y = N,
                 fill = is_DRACH)) +
        geom_col() +
        coord_flip() +
        scale_y_continuous(trans = "log10") + annotation_logticks(sides = "b") +
        theme_bw() +
        theme(axis.text.y = element_text(family = "Courier",size = 7, hjust =0.5)) +
        labs(x = "Motif", y = "Sites [N]", fill = "")
      ,
      ggplot(.SD,
             aes(x = factor(kmer, levels = unique(kmer[order(N)])),
                 y = y,
                 col = is_DRACH)) +
        geom_errorbar(aes(ymin=ymin, ymax=ymax), width = 0) +
        geom_point() +
        coord_flip() +
        theme_bw() +
        theme(axis.text.y = element_blank(), axis.title.y = element_blank()) +
        labs(y = "Avg. m6A level [%]"),
      ncol = 2, common.legend = TRUE
      )
    ]

```

```{r lift glori to DRACH}

drach.tx.glori.hits <-
  suppressWarnings(findOverlaps(query = drach.tx, subject = glori$tx))

drach.tx$glori_m6A_lvl <- as.numeric(0)

drach.tx[queryHits(drach.tx.glori.hits)]$glori_m6A_lvl <-
  glori$tx[subjectHits(drach.tx.glori.hits)]$m6A_lvl

```

```{r gloribins}

gloribins <-
  glori$dt[Gene != "ELSE", .(glori_m6A_lvl = sum(m6A_lvl)), by = .(gene = Gene)][
    , gloribin := cut2(glori_m6A_lvl, g= 50)
  ][
  ]



gloribins[
  , merge(.SD, exlenTabGene, "gene")
][
  , ggplot(.SD, aes(x = gloribin, y = glori_m6A_lvl)) +
    geom_boxplot()
]

gloribins[
  , merge(.SD, exlenTabGene, "gene")
][
  , ggplot(.SD, aes(x = gloribin, y = exlen)) +
    geom_boxplot() +
    coord_cartesian(ylim = c(NA, 2000))
]

```

## --- decay rates ---

### Collect data

#### mcm5s2U SLAM-Seq

```{r mcm5s2U_SLAM}

mcm5s2U_SLAM <-
  within(list(), {
    ## initial setup
    gse = NA
    experiment_name = "mcm5s2U_quantseq"
    samples = fread("../samples.mcm5s2U_quantseq.tsv")
    
    config  = yaml::read_yaml("../config.mcm5s2U_quantseq.yaml")
    genome  = "homo_sapiens_GRCh38_v35_ERCC92"
    
    ## sample metadata
    meta = samples[, c("run", "title")]
    
    meta[, label_type := "4sU"]
    meta[, label_conc := "100"]
    meta[, label_time := str_extract(title, "[0-9]+h")]
    meta[grepl("0IAA", run), label_time := "0h"]
    meta[grepl("100IAA", run), label_time := "4h"]
    meta[, label_time  := factor(label_time, levels = gtools::mixedsort(unique(label_time))) ]
    meta[, label_time_num := as.numeric(gsub("h", "", label_time))]
    meta[, label_treat := "IAA"]
    
    meta[, genotype := str_extract(title, "WT|ELP|CTU2|FTSJ1")]
    meta[, genotype := sedit(genotype, c("ELP", "CTU2", "FTSJ1"), c("ELP1 KO", "CTU2 KO", "FTSJ1 KO")  )]
    meta[, genotype := factor(genotype, levels = c("WT", "ELP1 KO", "CTU2 KO", "FTSJ1 KO"))]
    
    meta[, replicate := ifelse(grepl("IAA", title), "pilot", paste0("rep", str_extract(title, "[A-Z]$"))) ]
    
    meta
    
    ## grandslam data
    slam <- 
      read_grandslam_introns("../results/mcm5s2U_quantseq/gedi/homo_sapiens_GRCh38_v35_ERCC92/slam.tsv")
      
    ## calculate new and old from min2 (if there is no read with min2 transitions, we set new to 0)
    slam[, new := ifelse(setkeyv(meta, "run")[run, label_time_num] == 0, 0,  ifelse(min2 > 0, RPM * (min2 / Readcount), NA)) ] 
    slam[, old := RPM - new ]
    
    ## calculate NTRs
    slam[, NTR_grandslam := MAP]
    slam[, NTR_min2 := min2/Readcount]
    slam[, NTR_conv := (Conversions / Coverage) / 0.15 ]
    
    ## calculate decay rate
    slam[, decay_rate_ntr  := decay_from_NTR(label_time = setkeyv(meta, "run")[run, label_time_num], NTR = NTR_grandslam)  ]
    slam[, decay_rate_min2 := decay_from_NTR(label_time = setkeyv(meta, "run")[run, label_time_num], NTR = NTR_min2)]
    slam[, decay_rate_conv := decay_from_NTR(label_time = setkeyv(meta, "run")[run, label_time_num], NTR = NTR_conv)]
    head(slam)
    
    ## define robustly expressed genes
    expressed_in_all <-
      slam[region == "exonic",
              .(is_expressed = Readcount > 0 ),
              by = c("gene_id", "run")][,
                                        .(is_expressed_in_all = sum(is_expressed) == length(is_expressed)),
                                        by = "gene_id"][(is_expressed_in_all),
                                                        gene_id]
    cat(paste("Of", length(unique(slam$gene_id)), "genes,", sum(unique(slam$gene_id) %in% expressed_in_all), "are expressed in all runs."))
    

    
    ## define outlier runs
    # outliers = meta[title == "WT_200IAA", run]
    outliers = c("CTU2_2h_B")
    # outliers = c("")
    
    tpm <-
      dcast(slam[region == "exonic",], gene_id ~ run, value.var = "TPM")
    
    cts <-
      dcast(slam[region == "exonic",], gene_id ~ run, value.var = "Readcount")

    dtm0 <-
      merge(slam, meta, "run")[, .(time = label_time_num, 
               gene_id,
               region,
               genotype, 
               replicate, 
               expression = ifelse(label_time_num == 0, 1, 1 - NTR) )]
  })

```

```{r}

registerDoParallel(cores = 4)
system.time( 
mcm5s2U_SLAM <-
  within(mcm5s2U_SLAM, {
    decay_coef_file = file.path(tmpDir, paste0("decay_coef_", digest::digest(dtm0), ".tsv"))
    if(file.exists(decay_coef_file)){
      decay_coef <- 
        fread(decay_coef_file)
      } else {
        decay_coef <-
          foreach(data = c("dtm0"), .combine = rbind) %:% foreach(method = c("nls", "lm"), .combine = rbind) %dopar% {
            get(data)[gene_id %in% exlenTabGene$gene_id,  fit_decay(.SD, method = method)$coef
                      ,  by = c("gene_id", "region", "genotype", "replicate")][, data := data][, method := method]
            }
        fwrite(decay_coef, decay_coef_file)
      }
    
  })
)

head(mcm5s2U_SLAM$decay_coef)

```

#### Lugowski DRUID

Half-life data from: Lugowski, A., Nicholson, B., and Rissland, O.S. (2018). DRUID: a pipeline for transcriptome-wide measurements of mRNA stability. RNA 24, 623–632. We then back-calculate decay rates to be consistent with other datasets.

```{r druid}

dataURLs = c("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE99nnn/GSE99517/suppl/GSE99517%5FHEK%2E4SU%2EHL%2EDRUID%2Eone%2Ecsv%2Egz", 
             "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE99nnn/GSE99517/suppl/GSE99517%5FHEK%2E4SU%2EHL%2EDRUID%2Etwo%2Ecsv%2Egz")
dataFiles = file.path(tmpDir, basename(sapply(dataURLs, URLdecode )))

foreach(dataURL = dataURLs, dataFile = dataFiles) %do% {
  if(!file.exists(dataFile))download.file(dataURL, dataFile)
}

druid <-
  foreach(dataFile = dataFiles, .combine = rbind) %do% {
    fread(dataFile)[, replicate := sedit(str_extract(dataFile, "one|two"), c("one", "two"), c("rep1", "rep2") )]
  }
setnames(druid, "V1", "gene")

druid <-
  dcast(druid, gene ~ replicate, value.var = names(druid[, !c("gene", "replicate")]))

## back-calculate decay rates
decay_druid <-
  druid[, lapply(.SD, function(x){log(2)/x} ), by = "gene"]
setnames(decay_druid, c(names(decay_druid[, 1]), paste0("druid_", names(druid[, -1]))))

head(decay_druid)

```

#### Bazzini lab

1.  Wu, Q., Medina, S.G., Kushawah, G., DeVore, M.L., Castellano, L.A., Hand, J.M., Wright, M., and Bazzini, A.A. (2019). Translation affects mRNA stability in a codon-dependent manner in human cells. eLife 8, e45396. https://doi.org/10.7554/eLife.45396.

```{r wu_2019}
require(org.Hs.eg.db)


wu_2019 <-
  within(list(), {
    dataURL = "https://elifesciences.org/download/aHR0cHM6Ly9jZG4uZWxpZmVzY2llbmNlcy5vcmcvYXJ0aWNsZXMvNDUzOTYvZWxpZmUtNDUzOTYtZmlnMS1kYXRhMS12Mi54bHN4/elife-45396-fig1-data1-v2.xlsx?_hash=nQA8gIYVIymZnU1QP4ErD2TtiD9Qtm%2B5sBoO2wJmujs%3D"
    dataFile = file.path(tmpDir, basename(dataURL))
if(!file.exists(dataFile))download.file(url = dataURL, destfile = dataFile)
    
    data <-
  rbindlist(foreach(i = 1:6) %do% {as.data.table(read_xlsx(dataFile, sheet = i))}, fill = TRUE)
setnames(data, "Name", "ensembl_gene_id")

## remove ORFome data
data = data[datatype != "orf-ome"]
## some datasets have the sign included; for these datasets, we need to flip the sign
data[term == "time", decay_rate := - decay_rate]
    
  })


head(wu_2019$data)


## read data


```

#### Emetine SLAM-Seq

1.  Rosa-Mercado, N.A., Buskirk, A.R., and Green, R. (2024). Translation elongation inhibitors stabilize select short-lived transcripts. RNA, rna.080138.124. https://doi.org/10.1261/rna.080138.124.

```{r}
require(bakR)

emetine <-
  within(list(), {
    
    ## low dose data
    dataURL = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE264661&format=file&file=GSE264661%5FLD%5Flog2fc%5FJuly2023%2Ecsv%2Egz"
    dataFile = file.path(tmpDir, basename(URLdecode(dataURL)) )
    # file.remove(dataFile)
    if(!file.exists(dataFile)){download.file(dataURL, dataFile)}
    
    # ld <-
    #   setnames(fread(dataFile), "V1", "gene")
    
    ## high dose data
    dataURL = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE264661&format=file&file=GSE264661%5FHD%5Flog2fc%5FJuly2023%2Ecsv%2Egz"
    dataFile = file.path(tmpDir, basename(URLdecode(dataURL)) )
    # file.remove(dataFile)
    if(!file.exists(dataFile)){download.file(dataURL, dataFile)}
    
    # hd <-
    #   setnames(fread(dataFile), "V1", "gene")
    
    
    
    decayFile = file.path(tmpDir, "emetine_Fit.rds" )
    
    if(!file.exists(decayFile)){
      
      ## cB
      dataURL = "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE264661&format=file&file=GSE264661%5FcB%2Ecsv%2Egz"
      
      dataFile = file.path(tmpDir, basename(URLdecode(dataURL)) )
      # file.remove(dataFile)
      if(!file.exists(dataFile)){download.file(dataURL, dataFile)}
      cB <-
        fread(dataFile)
      
      metadf <-
        data.frame(row.names = unique(cB$sample),
                   tl = ifelse(grepl("ctl", unique(cB$sample)), 0, as.numeric(str_extract(unique(cB$sample), "[0-9]$"))),
                   Exp_ID = ifelse(grepl("UT", unique(cB$sample)), 1, 2) )
      
      bakRData <-
        bakRData(cB, metadf)
      
      Fit <-
        bakRFit(bakRData)
      
      saveRDS(object = Fit, file = decayFile)
      cB = NULL
      gc()
    }else{
      Fit <-
        readRDS(decayFile)
    }
    
  })

head(emetine$Fit$Fast_Fit$Fn_Estimates)

```

#### mcm5s2U STM2457

```{r mcm5s2U_STM}

mcm5s2U_STM <-
  readRDS("~/exports/m6A_decay_ms_share/data/mcm5s2U_actD_tpm_noryeast_new.RDS")

mcm5s2U_STM <-
  mcm5s2U_STM[ c("meta", "tpm", "outliers")]

setnames(mcm5s2U_STM$tpm, "gene_id", "ensembl_gene_id")

mcm5s2U_STM$outliers

mcm5s2U_STM$meta

head(mcm5s2U_STM$tpm)

```

```{r}

mcm5s2U_STM$log2FC_STM <-
  mcm5s2U_STM$tpm[
    , melt(.SD, id.vars = "ensembl_gene_id", variable.name = "run")
  ][
    , merge(.SD, mcm5s2U_STM$meta[!run %in% mcm5s2U_STM$outliers], "run")
  ][
    , dcast(.SD, ensembl_gene_id + genotype + time ~ treatment, value.var = "value", fun.aggregate = mean)
  ][
    , log2FC := log2((STM+1)/(mock+1))
  ][
    is.finite(log2FC)
  ][
    
  ]

head(mcm5s2U_STM$log2FC_STM)

```

### Consolidate decay data

Here we create a single table with decay rates. Note that if decay rates were calculated with different methods, we can choose which to use here.

```{r ++ decayData}

## a table of decay rates
decayData <-
  Reduce(
    f = function(x, y) {
      merge(x, y, "gene", all.x = TRUE)
    },
    init = exlenTabGene[, .(gene)],
    x = list(
      wu_2019$data[
        , merge(.SD, exlenTabGene, "ensembl_gene_id")
      ][
        is.finite(decay_rate), dcast(.SD, gene ~ datatype + cell_type + term, value.var = "decay_rate", fun.aggregate = mean)
      ]
      ,
     druid[, .(gene, std_rep1, std_rep2)]
    ))

decayData <-
  decayData[apply(decayData[, -1], 1, function(x){!all(is.na(x))}), ]

head(decayData)

```

```{r adjust decayData, fig.width = 15, fig.height = 15}

decayDataCorrected <-
  decayData[, c(.SD[, 1], lapply(.SD[, -1], shift_ntile_zero, ntile = 0.01)) ]

```

```{r}

ggplot(decayData, aes(x = value, col = normalization)) +
  stat_density(data = function(x)melt(x, id.vars = "gene")[, normalization := "original"], geom="line", linetype = 2) +
  stat_density(data = function(x)melt(x[, c(.SD[, 1], lapply(.SD[, -1], shift_ntile_zero, ntile = 0.01)) ], id.vars = "gene")[, normalization := "ntile_zero"], geom="line", linetype = 2) +
  stat_density(data = function(x)melt(x[,  as.data.table(limma::normalizeQuantiles(as.matrix(.SD[, -1], rownames = .SD[[1]])), keep.rownames = names(.SD[1])) ], id.vars = "gene")[, normalization := "quantile"], geom="line", linetype = 2) +
  coord_cartesian(xlim = c(-5, 10)) +
  #xlim(-5,10) +
  facet_wrap(~ variable, scales = "free_y") +
  NULL

ggplot(decayData, aes(x = value, col = variable)) +
  stat_density(data = function(x)melt(copy(decayData), id.vars = "gene")[, normalization := "original"], geom="line", linetype = 2) +
  stat_density(data = function(x)melt(copy(decayDataCorrected), id.vars = "gene")[, normalization := "ntile_zero"], geom="line", linetype = 2) +
  coord_cartesian(xlim = c(-5, 10)) +
  #xlim(-5,10) +
  facet_wrap(~ normalization) +
  NULL

```

```{r decaybins}

decaybins <-
  decayDataCorrected[
    , .(gene, decay_rate = endogenous_293t_time)
  ][
    , t12 := t12_from_decay(decay_rate)
  ][
   ! ((t12 < 0) | (t12 > 48)) , t12_bin := cut2(t12, g = 17, digits = 2)
  ][
    
  ]

decaybins[
  ,
  ][
    , ggplot(.SD, aes( x= t12, fill = t12_bin)) +
      geom_histogram(bins = 100) +
      scale_x_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      geom_vline(xintercept = 5.5, linetype = 2, col = "blue") +
      scale_fill_manual("t12 bins", values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      NULL
  ]

```

## --- reporter qPCR ---

```{r qpcr_data_complete, fig.width = 24, fig.height = 4}

qpcr_data_combined = fread("~/exports/m6A_decay_ms_share/data/reporter_qPCR_Ct_values.tsv", sep ="\t")

qpcr_data_combined[
  , experiment := as.character(experiment)
][
  , reporter := factor(reporter, levels = names(palval_reporter))
][
  , txn_inh := factor(txn_inh, levels = c("ActD", "TLD"))
][
  , m6A_inh := factor(m6A_inh, levels = c("mock", "STM")) 
][
  , transl_inh := factor(gsub("none", "mock", str_split_i(sample , "_", 5)), levels = c("mock", "EM", "CHX"))
][
  , genotype := factor(gsub("KO", " KO", toupper(str_split_i(sample , "_", 6))), levels = names(palval_genotype))
]

## for experiments where the same untreated sample is used as t0, we need to fill in this sample for the conditions with translation inhibitor
qpcr_data_filled <- 
  qpcr_data_combined[
    , {
      ## duplicate t0 values
      orig_data <- copy(.SD)
      
      orig_data[, condition_wo_transl_inh := paste(experiment, genotype, reporter, txn_inh, m6A_inh, target, replicate)]
      
      missing_CHX_t0 <-
        orig_data[transl_inh == "CHX", dcast(.SD,  condition_wo_transl_inh ~ time, value.var = "Ct")][is.na(`0h`) & (!is.na(`1h`)|!is.na(`2h`)|!is.na(`4h`)) , condition_wo_transl_inh]
      
      missing_EM_t0 <-
        orig_data[transl_inh == "EM" , dcast(.SD,  condition_wo_transl_inh ~ time, value.var = "Ct")][is.na(`0h`) & (!is.na(`1h`)|!is.na(`2h`)|!is.na(`4h`)) , condition_wo_transl_inh]

      orig_data[, is_filled := FALSE]

      filled_data <-
        rbindlist(
          list(
            orig_data,
            orig_data[time == "0h" & transl_inh == "mock" & condition_wo_transl_inh %in% missing_CHX_t0, ][, transl_inh := "CHX"][, is_filled := TRUE],
            orig_data[time == "0h" & transl_inh == "mock" & condition_wo_transl_inh %in% missing_EM_t0,  ][, transl_inh := "EM" ][, is_filled := TRUE]
          )
        )

      filled_data
      
    }
    #, by = .(experiment, genotype, reporter, txn_inh, m6A_inh, target, replicate)
  ]

qpcr_data_normalized_spikein <-
  qpcr_data_filled[
    , {
      ## normalize each sample to the TDH1 / yeast spike-in to control for cellular RNA amount / cell viability.
      norm_data <- .SD[target == "TDH1" , .(norm_factor = mean(expression)), by = .(sample, experiment, replicate)]
      data = merge(.SD, norm_data, by = c("sample", "experiment", "replicate"))
      data[, expression := expression / norm_factor]
      data[, norm_factor := NULL]
    }
  ]

qpcr_data_normalized_mCherry <- 
  qpcr_data_normalized_spikein[
    , {
      ## normalize the mNeonGreen2 in each sample to mCherry to control for transfection efficiency
      ## we need a helper object "has_plasmid" to prevent dropping out of samples w/o plasmid
      data = copy(.SD)[, has_plasmid := reporter != "none"]
      norm_data <- data[, .(norm_factor = mean(expression[grepl("mCherry|NeoR", target)])), by = .(sample, experiment, replicate, has_plasmid)]
      norm_data[!(has_plasmid), norm_factor := 1]
      data = merge(data, norm_data, by = c("sample", "experiment", "replicate", "has_plasmid"))
      data[grepl("mCherry|NeoR|Neon2", target), expression := expression / norm_factor] ## we only normalize the reporter mRNAs to mCherry
      data[, norm_factor := NULL]
      data[, has_plasmid := NULL]
    }
  ]

qpcr_data_normalized_t0 <-
    qpcr_data_normalized_mCherry[
      , {
        ## finally, we normalize each time series to t0. This is optional and should not change the half-lives.
        norm_data <- .SD[time == "0h" , .(norm_factor = mean(expression)), by = .(target, reporter, txn_inh, m6A_inh, transl_inh, genotype, experiment)]
        data = merge(.SD, norm_data, by = c("target", "reporter", "txn_inh", "m6A_inh", "transl_inh", "genotype", "experiment"))
        data[, expression := expression / norm_factor]
        data[, norm_factor := NULL][]
        data
      }
    ]

qpcr_data_complete <-
  qpcr_data_normalized_t0[
    , {
      ## filter for complete timecourse
      data = .SD
      condition_table = data[, .(has_all_timepoints = all((c("0h") %in% as.character(time) & any(c("1h", "2h", "4h") %in% as.character(time)) )  ),
                                 timepoints_missing = paste(setdiff(c("0h", "1h", "2h", "4h"), unique(as.character(time))), collapse = " ") )
                             , by = .(experiment, genotype, reporter, txn_inh, transl_inh, m6A_inh, target)
      ]
      condition_table = unique(condition_table)
      data[paste(experiment, genotype, reporter, txn_inh, transl_inh, m6A_inh, target) %in%
             condition_table[(has_all_timepoints), paste(experiment, genotype, reporter, txn_inh, transl_inh, m6A_inh, target)] ]
    }
  ]

head(qpcr_data_complete)

```

```{r  fig.width = 24, fig.height = 4}

qpcr_data_filled[
  ! target %in% c("TDH1", "mCherry")
][
  , ggplot(.SD,
           aes(x = time_num, y = expression)) +
    geom_hline(yintercept = 1, linetype = 2) +
    geom_quasirandom(aes(shape= paste(experiment, replicate), col = is_filled), width = 0.2) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE, col = "black", linewidth = 0.5) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), aes(label = paste0(round(after_stat(log(2)/k_estimate), digits = 1), "h") )) +
    scale_shape_manual(values= c(1:3,5:22)) +
    facet_nested(target ~ genotype + reporter + txn_inh + transl_inh + m6A_inh, labeller = labeller(reporter = labeller_reporter_codons), scales = "free_y")  +
    theme_bw() +
    labs(title = paste(unique(.SD$experiment), collapse = " + ") ) +
    NULL
]

```

```{r  fig.width = 24, fig.height = 4}

qpcr_data_complete[
  ! target %in% c("TDH1", "mCherry")
][
  , ggplot(.SD,
           aes(x = time_num, y = expression)) +
    geom_hline(yintercept = 1, linetype = 2) +
    geom_quasirandom(aes(shape= paste(experiment, replicate), col = experiment), width = 0.2) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE, col = "black", linewidth = 0.5) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), aes(label = paste0(round(after_stat(log(2)/k_estimate), digits = 1), "h") )) +
    scale_shape_manual(values= c(1:3,5:22)) +
    facet_nested(target ~ genotype + reporter + txn_inh + transl_inh + m6A_inh, labeller = labeller(reporter = labeller_reporter_codons), scales = "free_y")  +
    theme_bw() +
    labs(title = paste(unique(.SD$experiment), collapse = " + ") ) +
    NULL
]

```

## --- Ribosome profiling ---

```{r}
## Set the method to merge genes with ribosome profiling coverage within experiment. Use either "union" or "intersect".
gene_coverag_merge_mode = "union" 
```

#### mcm5s2U riboseq

```{r mcm5s2U_riboseq}

## function for reading Jie's files
read_tx_coverage <-
  function(coverage_file, file_suffix){
    data <-
      fread(coverage_file, sep = "\t", sep2 = ",", header = FALSE)[
        , .(codon_rank = as.integer(1:(V2+1)), coverage = as.numeric(str_split_1(V3, ",")) )
        , by = .(ensembl_transcript_id = str_split_i(V1, "\\|", 2), cds_len_cods = V2 )]
    setnames(data, "coverage", gsub(file_suffix, "", basename(coverage_file)))
    setnames(data, gsub("^PSLib[0-9]+_RPF_", "", names(data)))
  }

## wrapper to get coverage objects
get_tx_coverage <-
  function(experiment_name, file_path, file_suffix, combine = "intersect"){
    
    coverage_files = list.files(path = file_path, pattern = file_suffix, full.names = TRUE)
    if(length(coverage_files) == 0){stop("No files found!")}else(cat(paste("Reading", length(coverage_files), "files.\n")))
    
    ## define combine method
    if(combine == "union"){
      cat("Using union for gene merging.\n")
      use_all = FALSE
    }else if(combine == "intersect"){
      cat("Using intersection for gene merging.\n")
      use_all = TRUE
    }else{
      cat("Error: Undefined method for gene merging. Use 'union' or 'intersect'. \n")
      stop()
    }
    
    ## read in data
    coverage <-
      foreach(coverage_file = coverage_files,
              .combine = function(x,y){merge(x,y, c("ensembl_transcript_id", "cds_len_cods", "codon_rank"), all = use_all)}) %do% {
                read_tx_coverage(coverage_file, file_suffix)
              }
    
    meta <-
      data.table(experiment = experiment_name, run = setdiff(names(coverage), c("ensembl_transcript_id", "cds_len_cods", "codon_rank")) )
    
    
    return(list(file_suffix = file_suffix, coverage = coverage, meta = meta))
    
  }


# mcm5s2U_riboseq experiment 1
## read in data
mcm5s2U_riboseq_1 <-
  get_tx_coverage(experiment_name = "exp1",
                  file_path = "/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/RP_coverage/old_experiment/RPM_coverage/",
                  file_suffix = "_clipped_tr_nohrRNA_un_codon_Asite_offset_RPM_coverage.txt",
                  combine = gene_coverag_merge_mode)
## normalize in place
with(mcm5s2U_riboseq_1, coverage[,meta$run := lapply(.SD, function(x){x/(sum(x[is.finite(x)])/cds_len_cods) }), .SDcols = meta$run, by = .(ensembl_transcript_id, cds_len_cods)])

## metadata
mcm5s2U_riboseq_1$meta[, title := run]
mcm5s2U_riboseq_1$meta[, genotype  := factor(sedit(str_extract(title, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO")) ]
mcm5s2U_riboseq_1$meta[, treatment := factor("CTRL", levels = c("CTRL", "STM")) ]
mcm5s2U_riboseq_1$meta[, replicate := str_extract(title, "[0-9]$") ]


# mcm5s2U_riboseq experiment 2 (STM treatment)
## read in data
mcm5s2U_riboseq_2 <-
  get_tx_coverage(experiment_name = "exp2",
                  file_path = "/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/RP_coverage/new_experiment/RPM_coverage/",
                  file_suffix = "_Mono_clpd_tr_no_r_t_sno_sn_RNA_codon_Asite_offset_RPM_coverage.txt",
                  combine = gene_coverag_merge_mode)
## normalize in place
with(mcm5s2U_riboseq_2, coverage[,meta$run := lapply(.SD, function(x){x/(sum(x[is.finite(x)])/cds_len_cods) }), .SDcols = meta$run, by = .(ensembl_transcript_id, cds_len_cods)])

## metadata
mcm5s2U_riboseq_2$meta[, title := run]
mcm5s2U_riboseq_2$meta[, genotype  := factor(sedit(str_extract(title, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO")) ]
mcm5s2U_riboseq_2$meta[, treatment := factor(str_extract(title, "CTRL|STM"), levels = c("CTRL", "STM")) ]
mcm5s2U_riboseq_2$meta[, replicate := str_extract(title, "[0-9]$") ]


#
# combine experiments
#

mcm5s2U_riboseq <-
  list(
    coverage = merge(mcm5s2U_riboseq_1$coverage, mcm5s2U_riboseq_2$coverage, c("ensembl_transcript_id", "cds_len_cods", "codon_rank"), all = TRUE),
    meta = rbind(mcm5s2U_riboseq_1$meta, mcm5s2U_riboseq_2$meta)
  )

setkeyv(mcm5s2U_riboseq$meta, "run")

rm("mcm5s2U_riboseq_1", "mcm5s2U_riboseq_2")
gc()

head(mcm5s2U_riboseq$coverage)

mcm5s2U_riboseq$meta

```

```{r lift mcm5s2U coverage to codons}

mcols(cods.gr) = mcols(cods.gr)[! names(mcols(cods.gr)) %in% mcm5s2U_riboseq$meta$run]
gc()

mcols(cods.gr) <-
  cbind(mcols(cods.gr), as.data.table(cods.gr)[
    , codonID := paste0(gsub("\\..*$", "", seqnames), "_",codon_rank)
  ][][
    , merge(.SD, mcm5s2U_riboseq$coverage[, codonID := paste0(ensembl_transcript_id, "_", codon_rank)][
      , c("codonID", mcm5s2U_riboseq$meta$run)
      , with = FALSE
    ], "codonID", all.x = TRUE, sort = FALSE) 
  ][
    , mcm5s2U_riboseq$meta$run
    , with = FALSE
  ])

## set NA coverage to 0
for(run in mcm5s2U_riboseq$meta$run) {
  mcols(cods.gr)[run][which(is.na(mcols(cods.gr)[run])),] = 0
}

cods.gr

## remove coverage data to free up RAM
mcm5s2U_riboseq$coverage = NULL
gc()

```

#### Queuosine riboseq

From: 1. Zhao, X., Ma, D., Ishiguro, K., Saito, H., Akichika, S., Matsuzawa, I., Mito, M., Irie, T., Ishibashi, K., Wakabayashi, K., et al. (2023). Glycosylated queuosines in tRNAs optimize translational rate and post-embryonic growth. Cell 186, 5517–5535. https://doi.org/10.1016/j.cell.2023.10.026.

```{r queuosine_riboseq}

## read in data
queuosine_riboseq <-
  get_tx_coverage(experiment_name = "zhao",
                  file_path = "/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/RP_coverage/queuosine/normalized/",
                  file_suffix = "_coverage.txt",
                  combine = gene_coverag_merge_mode)

## metadata
queuosine_riboseq$meta[, title := run]
queuosine_riboseq$meta[, genotype  := factor(sedit(str_extract(title, "WT|QTRT1|QTRT2"), from = c("QTRT1", "QTRT2"), to = c("QTRT1 KO", "QTRT2 KO")), levels = c("WT", "QTRT1 KO", "QTRT2 KO")) ]
queuosine_riboseq$meta[, replicate := str_extract(title, "[0-9]$") ]

queuosine_riboseq

```

```{r lift queuosine coverage to codons}

mcols(cods.gr) = mcols(cods.gr)[! names(mcols(cods.gr)) %in% queuosine_riboseq$meta$run]
gc()

mcols(cods.gr) <-
  cbind(mcols(cods.gr), as.data.table(cods.gr)[
    , codonID := paste0(gsub("\\..*$", "", seqnames), "_",codon_rank)
  ][][
    , merge(.SD, queuosine_riboseq$coverage[, codonID := paste0(ensembl_transcript_id, "_", codon_rank)][
      , c("codonID", queuosine_riboseq$meta$run)
      , with = FALSE
    ], "codonID", all.x = TRUE, sort = FALSE) 
  ][
    , queuosine_riboseq$meta$run
    , with = FALSE
  ])

## set NA coverage to 0
for(run in queuosine_riboseq$meta$run) {
  mcols(cods.gr)[run][which(is.na(mcols(cods.gr)[run])),] = 0
}

cods.gr

## remove coverage data to free up RAM
queuosine_riboseq$coverage = NULL
gc()

```

# Paper figures

```{r}
#stop()
```


## m6A position effect

```{r m6A_region_miCLIP}

m6A_region_miCLIP <-
  setnames(as.data.table(drach.tx), "seqnames", "tx_name")[
    , merge(.SD, txmodel[, .(tx_name, ensembl_gene_id)], "tx_name")
  ][
    ( is_m6A_miCLIP )
  ][
    , .(N_m6A = .N)
    , by = .(ensembl_gene_id, region)
  ][
    , merge(exlenTabGene, .SD, all.x = TRUE) ## we need all.x = TRUE to include genes without m6A
  ][
    , .(gene, region, N_m6A)
  ][
    , region := factor(region, levels = c("utr5", "cds", "utr3"))
  ][
    , dcast(.SD, gene ~ region, value.var = "N_m6A", fun.aggregate = sum, fill = 0)
  ][
    !is.na(gene) , !"NA"
  ][
    , m6A_region := ifelse(utr5+cds+utr3 == 0,
                           "none",
                           paste0(
                             ifelse(utr5 > 0, "utr5", ""),
                             ifelse(cds  > 0, "cds",  ""),
                             ifelse(utr3 > 0, "utr3", "")
                           ))
  ][
    , m6A_region := factor(m6A_region, levels = c("none", "utr5", "utr5cds", "utr5cdsutr3", "cds", "cdsutr3", "utr3", "utr5utr3")) 
  ][
    
  ]



m6A_region_miCLIP[, .N, by = .(m6A_region)]

head(m6A_region_miCLIP)

```

```{r m6A_region_GLORI}

m6A_region_GLORI <-
  setnames(as.data.table(drach.tx), "seqnames", "tx_name")[
    , merge(.SD, txmodel[, .(tx_name, ensembl_gene_id)], "tx_name")
  ][
    , is_m6A_GLORI := glori_m6A_lvl > 0.1
  ][
    ( is_m6A_GLORI )
  ][
    , .(N_m6A = .N)
    , by = .(ensembl_gene_id, region)
  ][
    , merge(exlenTabGene, .SD, all.x = TRUE) ## we need all.x = TRUE to include genes without m6A
  ][
    , .(gene, region, N_m6A)
  ][
    , region := factor(region, levels = c("utr5", "cds", "utr3"))
  ][
    , dcast(.SD, gene ~ region, value.var = "N_m6A", fun.aggregate = sum, fill = 0)
  ][
    !is.na(gene) , !"NA"
  ][
    , m6A_region := ifelse(utr5+cds+utr3 == 0,
                           "none",
                           paste0(
                             ifelse(utr5 > 0, "utr5", ""),
                             ifelse(cds  > 0, "cds",  ""),
                             ifelse(utr3 > 0, "utr3", "")
                           ))
  ][
    , m6A_region := factor(m6A_region, levels = c("none", "utr5", "utr5cds", "utr5cdsutr3", "cds", "cdsutr3", "utr3", "utr5utr3")) 
  ][
    
  ]

m6A_region_GLORI[, .N, by = .(m6A_region)]

head(m6A_region_GLORI)

```

```{r region_metagene_miCLIP, fig.width = 2, fig.height = 1}

metagene <-
  ggplot(as.data.table(drach.tx)[
    ( is_m6A_miCLIP )
  ]
  , aes(
    x = relpos.cds
    , fill = region
  )) +
  geom_histogram(bins = 200) +
  geom_vline(xintercept = c(1,2)
             , col ="blue"
             , linetype=2) +
  scale_fill_manual(values = palval_m6A_region
                    , labels = c("utr5" = "5'UTR", "cds" = "CDS", "utr3" = "3'UTR")) +
  scale_x_continuous(expand = expansion(mult = 0)) +
  scale_y_continuous(expand = expansion(c(0, 0))) +
  labs(y = expression("m"^6*"A sites [N]")
       , fill = "Transcript region") +
  theme_pubr(base_size = 8) +
  #theme(legend.position=c(0.35, 0.8), legend.background = element_rect(fill = NA)) +
  theme(legend.position="none") +
  theme(axis.text.x = element_blank()
        , axis.title.x = element_blank()
        , axis.ticks.x = element_blank()
  ) +
  guides(fill=guide_legend(ncol=3)) +
  theme(legend.position = "none"
        , panel.grid = element_blank()
        , #axis.line = element_blank()
        , axis.title.x = element_blank()
        , axis.text = element_blank()
        , axis.ticks = element_blank()
        , panel.background = element_blank()
  )

gridmin = layer_scales(metagene)$y$range$range[1]
gridmax = layer_scales(metagene)$y$range$range[2]
gridrng = gridmax - gridmin
gridmid = gridrng/ 2
sizefct = 2 ## adjust y size (higher gives thicker UTRs)

txlegend <-
  ggplot() +
  geom_blank() +
  geom_rect(aes(xmin = -Inf, xmax = 1,    ymin = gridmin+gridrng/(2*sizefct) , ymax = gridmax-gridrng/(2*sizefct) ), fill = palval_m6A_region["utr5"]) +
  geom_rect(aes(xmin = 1,    xmax = 2,    ymin = gridmin, ymax = gridmax ), fill = palval_m6A_region["cds"]) +
  geom_rect(aes(xmin = 2,    xmax = Inf,  ymin = gridmin+gridrng/(2*sizefct) , ymax = gridmax-gridrng/(2*sizefct) ), fill = palval_m6A_region["utr3"]) +
  geom_vline(xintercept = c(1,2), col ="blue", linetype=2) +
  scale_x_continuous(expand = expansion(mult = 0), limits = layer_scales(metagene)$x$range$range) +
  scale_y_continuous(expand = expansion(mult = 0), limits = layer_scales(metagene)$y$range$range) +
  theme(legend.position = "none",
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank()) +
  NULL

p <-
  ggarrange(
    metagene + theme(plot.margin=unit(c(0,0,0,2), "mm"))
    , txlegend + theme(plot.margin=unit(c(0,0,0,2), "mm"))
    , ncol = 1
    , heights = c(10, 1)
    , align = "v")

p

ggsave("~/exports/m6A_decay_ms_share/region_metagene_miCLIP.pdf", width = 2, height = 1)

```

```{r region_decay_boxplot_miCLIP, fig.width = 2, fig.height = 3}

m6A_region_miCLIP[
  , merge(.SD, decayDataCorrected, "gene")
][
  gene %in% canonical_transcripts$gene
][
  , {
    xlabs = c("none" = expression("no"~m^6*"A"),  "cds" = "CDS", "utr3" = "3'UTR", "cdsutr3" = "CDS + 3'UTR")
    breaks = c("none", "cds", "utr3", "cdsutr3")
    labels = xlabs[breaks]
    #labels = setkeyv(.SD[, .N, by = "m6A_region"], "m6A_region")[breaks, paste0(xlabs[breaks], "\n(N=", N, ")") ]
    
    ggplot(.SD[m6A_region %in% names(xlabs)],
           aes(x = factor(m6A_region, levels = names(xlabs)),
               y = endogenous_293t_time
           )) +
      #geom_blank() +
      geom_boxplot_pattern(aes(fill = m6A_region, pattern_fill = m6A_region )
                           , notch = T, outlier.size = 0.5
                           , fatten = 1 ## controls median line segment
                           , pattern_colour = NA
                           , pattern_density = 0.5
                           , outlier.color = "grey50"
                           , outlier.alpha = 0.3
                           ) +
            stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      stat_compare_means(comparisons = list(c("none", "cds"), c("utr3", "cds"), c("utr3", "cdsutr3"))
                         , label = "p.signif"
                         , label.y =  0.2 + c(0:3)*0.006,
                         , tip.length = 0) +
            scale_fill_manual(values = palval_m6A_region_fill ) +
      scale_pattern_fill_manual(values = palval_m6A_region_fill_pattern) +
      scale_x_discrete(breaks = breaks, labels = labels) +
      scale_y_continuous(name =  expression("Decay rate [h"^-1*"]") ) +
      theme_bw() +
      theme(legend.position = "none") +
      theme(axis.title.x = element_blank()) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      #theme(axis.text.x = element_text(size = 5)) +
      coord_cartesian(ylim = c(0, 0.3)) +
      NULL
  }
]

ggsave("~/exports/m6A_decay_ms_share/region_decay_boxplot_miCLIP.pdf", width = 1.5, height = 4)

```

```{r}

m6A_region_miCLIP[
  , merge(.SD, decayDataCorrected, "gene")
][
  gene %in% canonical_transcripts$gene
][
  m6A_region %in% names(c("none" = expression("no"~m^6*"A"),  "cds" = "CDS", "utr3" = "3'UTR", "cdsutr3" = "CDS + 3'UTR"))
][
  , .N, by = "m6A_region"
]

```

```{r region_decay_boxplot_GLORI, fig.width = 2, fig.height = 3}

m6A_region_GLORI[
  , merge(.SD, decayDataCorrected, "gene")
][
  gene %in% canonical_transcripts$gene
][
  endogenous_293t_time < 3
][
  , {
    xlabs = c("none" = expression("no"~m^6*"A"),  "cds" = "CDS", "utr3" = "3'UTR", "cdsutr3" = "CDS + 3'UTR")
    breaks = c("none", "cds", "utr3", "cdsutr3")
    labels = xlabs[breaks]
    #labels = setkeyv(.SD[, .N, by = "m6A_region"], "m6A_region")[breaks, paste0(xlabs[breaks], "\n(N=", N, ")") ]
    
    ggplot(.SD[m6A_region %in% names(xlabs)],
           aes(x = factor(m6A_region, levels = names(xlabs)),
               y = endogenous_293t_time
           )) +
      geom_blank() +
      geom_boxplot_pattern(aes(fill = m6A_region, pattern_fill = m6A_region )
                           , notch = T, outlier.size = 0.5
                           , fatten = 1 ## controls median line segment
                           , pattern_colour = NA
                           , pattern_density = 0.5 
                           , outlier.color = "grey50"
                           , outlier.alpha = 0.3
                           ) +
            stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      stat_compare_means(comparisons = list(c("none", "cds"), c("utr3", "cds"), c("utr3", "cdsutr3"))
                         , label = "p.signif"
                         , label.y =  0.2 + c(0:3)*0.006
                         , tip.length = 0) +
      scale_fill_manual(values = palval_m6A_region_fill ) +
      scale_pattern_fill_manual(values = palval_m6A_region_fill_pattern) +
      scale_x_discrete(breaks = breaks, labels = labels) +
      scale_y_continuous(name =  expression("Decay rate [h"^-1*"]") ) +#
      theme_bw() +
      theme(legend.position = "none") +
      theme(axis.title.x = element_blank()) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      #theme(axis.text.x = element_text(size = 5)) +
      coord_cartesian(ylim = c(0, 0.3)) +
      NULL
  }
]

ggsave("~/exports/m6A_decay_ms_share/region_decay_boxplot_GLORI.pdf", width = 1.5, height = 4)

```

```{r}

m6A_region_GLORI[
  , merge(.SD, decayDataCorrected, "gene")
][
  gene %in% canonical_transcripts$gene
][
  m6A_region %in% names(c("none" = expression("no"~m^6*"A"),  "cds" = "CDS", "utr3" = "3'UTR", "cdsutr3" = "CDS + 3'UTR"))
][
  , .N, by = "m6A_region"
]

```

```{r}

m6A_region_miCLIP[
  , merge(.SD, decayDataCorrected, "gene")
][
  gene %in% mcm5s2U_SLAM$tpm[, .(expression = mean(WT_0h_A, WT_0h_B)) , by = .(gene_id) ][expression > 0.1, mapIds(x = org.Hs.eg.db, keys = gsub("\\..*$", "", gene_id), column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")]
][
  endogenous_293t_time < 0.8
][
  , {
    xlabs = c("none" = "no\nm6A",  "utr5" = "5'UTR")
    
    region_labeller <-
      function(breaks){
        #setkeyv(.SD[, .N, by = "m6A_region"], "m6A_region")[breaks, paste0(xlabs[breaks], "\n(N=", N, ")") ]
        xlabs[breaks]
      }
    
    ggplot(.SD[m6A_region %in% names(xlabs)],
           aes(x = factor(m6A_region, levels = names(xlabs)),
               y = endogenous_293t_time
           )) +
      geom_blank() +
      geom_boxplot_pattern(aes(fill = m6A_region, pattern_fill = m6A_region )
                           , notch = T, outlier.size = 0.5
                           , fatten = 1 ## controls median line segment
                           , pattern_colour = NA
                           , pattern_density = 0.5 
                           , outlier.color = "grey50"
                           , outlier.alpha = 0.3
                           ) +
      stat_compare_means(comparisons = list(c("none", "utr5"))
                         , label = "p.signif"
                         , label.y = 0.3 + c(0.03*0:2)
                         , tip.length = 0) + 
      scale_fill_manual(values = palval_m6A_region_fill ) +
      scale_pattern_fill_manual(values = palval_m6A_region_fill_pattern) +
      scale_x_discrete(labels = region_labeller) +
      scale_y_continuous(name =  expression("Decay rate [h"^-1*"]") ) +
      theme_pubr(base_size = 8) +
      theme(legend.position = "none") +
      theme(axis.title.x = element_blank()) +
      #theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) +
      #theme(axis.text.x = element_text(size = 8)) +
      coord_cartesian(ylim = c(-0.2, 0.5)) +
      NULL
  }
]

ggsave("~/exports/m6A_decay_ms_share/region_decay_boxplot_miCLIP_utr5.pdf", width = 1.5, height = 3)

```

```{r glori_decay_metagene, fig.width = 8, fig.height = 4}

p.glorimeta <-
  as.data.table(drach.tx)[
    , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
  ][
    ! ( (region == "utr5" & relpos.cds >= 1) | (region == "utr3" & relpos.cds <= 2) ) , ## exclude codon / nucleotide region discrepancies
  ][
    , merge(.SD, canonical_transcripts, "ensembl_transcript_id")
  ][
   # , merge(.SD, exlenTabGene, "ensembl_gene_id")
  ][
    , merge(.SD, decayDataCorrected, "gene")
  ][
    , relposbin := cut2(relpos.cds, cuts = seq(0, 3, 0.01))
  ][
    , decayrate := endogenous_293t_time
  ][
    , merge(.SD, decaybins, "gene", all.x = TRUE)
  ][
    ! is.na(t12_bin)
  ][
    , c("t12_bin_mean", "t12_bin_N_genes" ) := .(mean(t12), length(unique(ensembl_gene_id)))
    , by = .(t12_bin)
  ][
    , .(glori_m6A_sum = sum(glori_m6A_lvl), glori_m6A_mean = mean(glori_m6A_lvl)  )
    , by = .(relposbin, region, t12_bin, t12_bin_mean, t12_bin_N_genes)
  ][
    , t12_label := as.character(round(t12_bin_mean, 1))
  ][
    , t12_label := factor(t12_label, levels = gtools::mixedsort(unique(t12_label))  )
  ][
    ,
    ggplot(.SD,
           aes( x = as.integer(relposbin), y = glori_m6A_sum )) +
      geom_vline(data = function(x)x[region == "cds"], aes(xintercept = min(as.integer(relposbin))), linetype = 2, col = "blue" ) +
      geom_vline(data = function(x)x[region == "cds"], aes(xintercept = max(as.integer(relposbin))), linetype = 2, col = "blue" ) +
      ## points with continuous color scale
      geom_point(aes(col = t12_bin_mean), shape = 16, alpha = 0.5) +
      scale_color_gradientn(name = "Half-life [h]", colours = (brewer.pal(n = 7, name = "RdYlBu")), guide = "colorbar" ) +
      #guides(color = guide_legend(title.position = "top")) +
      ## smoothing splines with discrete color scale
      new_scale_color() +
      geom_smooth(aes(col = t12_label), method = "lm", formula = y ~ splines::bs(x, df = 40), se = F, show.legend = FALSE) +
      scale_color_manual("t12 bins", values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      guides(color = guide_legend()) +
      ## style
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(name = expression("GLORI m"^6*"A level"), expand = c(0, NA)) +
      theme_pubr(base_size = 8) +
      theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
      theme(legend.position  = c(0.3, 0.7), legend.direction = "horizontal", legend.background = element_blank(), legend.title = element_blank() ) +
      NULL
  ]

p.glorimeta

ggsave("~/exports/m6A_decay_ms_share/glori_decay_metagene.pdf", width = 4, height = 2)

```

```{r decaybins_expression}

decaybins[
  , merge(.SD, exlenTabGene)
][
  , merge(.SD, mcm5s2U_STM$tpm, "ensembl_gene_id")
][
  , merge(.SD, glori$dt[Gene != "ELSE", .(m6A_lvl = sum(m6A_lvl)), by = .(gene = Gene) ], "gene")
][
  !is.na(t12_bin)
][
  , 
  
  ggarrange(
    ggplot(.SD, aes(x = t12_bin, y = m6A_lvl, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 20)) +
      labs(x="", y = expression(m^6*A~level), fill = "Half-life bin") +
      NULL
    ,
    ggplot(.SD, aes(x = t12_bin, y = WT_Ctrl_0h_rep1, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 200)) +
      labs(x="Half-life bin [h]", y = "Expression level") +
      NULL
    , 
    ncol = 1, align = "v", common.legend = TRUE, legend = "right"
  )
  
]

ggsave("~/exports/m6A_decay_ms_share/decaybins_expression.pdf", width = 4.5, height = 4.5)

```

```{r , fig.height = 8, fig.width = 4}

decaybins[
  , merge(.SD, exlenTabGene)
][
  , merge(.SD, mcm5s2U_STM$tpm, "ensembl_gene_id")
][
  , merge(.SD, glori$dt[Gene != "ELSE", .(m6A_sites = length(m6A_lvl), m6A_lvl = sum(m6A_lvl)), by = .(gene = Gene) ], "gene")
][
  !is.na(t12_bin)
][
  ,  {
    p.legend.dummy <-
      ggplot(.SD, aes(x = t12_bin, y = m6A_lvl, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 20)) +
      labs(fill = "Half-life bin", x="Half-life bin [h]", y = expression("Sum"~m^6*A~"lvl")) +
      theme(axis.title.x = element_blank()) +
      NULL
    
    
    ggarrange(
    p.glorimeta + labs(y = expression("Sum"~m^6*A~"lvl"))
    ,
    ggplot(.SD, aes(x = t12_bin, y = m6A_lvl, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 20)) +
      labs(fill = "Half-life bin", x="Half-life bin [h]", y = expression("Sum"~m^6*A~"lvl")) +
      theme(axis.title.x = element_blank()) +
      NULL
    ,
    ggplot(.SD, aes(x = t12_bin, y = m6A_sites, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 40)) +
      labs(fill = "Half-life bin", x="Half-life bin [h]", y = expression("N"~m^6*A~"sites") ) +
      theme(axis.title.x = element_blank()) +
      NULL
    ,
    ggplot(.SD, aes(x = t12_bin, y = m6A_lvl/ m6A_sites, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 1)) +
      scale_y_continuous(breaks = c(0, 0.5, 1)) +
      labs(fill = "Half-life bin", x="Half-life bin [h]", y = expression("Site-avg."~m^6*A) ) +
      theme(axis.title.x = element_blank()) +
      NULL
    ,
    ggplot(.SD, aes(x = t12_bin, y = WT_Ctrl_0h_rep1, fill = t12_bin )) +
      geom_boxplot() +
      stat_summary(fun.data = mean_se, geom = "point", shape = 3) +
      scale_fill_manual(values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
      #scale_y_continuous(trans = "log10", breaks = c(3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 3), limits = c(1, 96) )  +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      theme(axis.text.x = element_blank()) +
      #theme(legend.position = "none" ) +
      coord_cartesian(ylim = c(0, 250)) +
      labs(fill = "Half-life bin", x = "Half-life bin [h]", y = expression(~"Expression level") ) +
      NULL
    , 
    ncol = 1, align = "hv", common.legend = TRUE, legend = "right"  , legend.grob = get_legend(p.legend.dummy)
  )
    
  }
  
  
  
 
]


 

```

```{r}

as.data.table(drach.tx)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  ! ( (region == "utr5" & relpos.cds >= 1) | (region == "utr3" & relpos.cds <= 2) ) , ## exclude codon / nucleotide region discrepancies
][
  , merge(.SD, canonical_transcripts, "ensembl_transcript_id")
][
  , merge(.SD, decayDataCorrected, "gene")
][
  , relposbin := cut2(relpos.cds, cuts = seq(0, 3, 0.01))
][
  , decayrate := endogenous_293t_time
][
  , t12 := t12_from_decay(decayrate)
][
 #  t12 > 0 & t12 < 96
][
 # , merge(.SD, decaybins[, .(ensembl_gene_id, t12_bin)], "ensembl_gene_id", all.x = TRUE) ## all.x = TRUE
][
  , t12_bin := cut2(t12, g = 17)
][
  , c("t12_bin_mean", "t12_bin_N_genes" ) := .(mean(t12), length(unique(ensembl_gene_id)))
  , by = .(t12_bin)
][
  , .(glori_m6A_sum = sum(glori_m6A_lvl)  )
  , by = .(relposbin, region, t12_bin, t12_bin_mean, t12_bin_N_genes)
][
  , t12_label := as.character(round(t12_bin_mean, 1))
][
  # , t12_label := paste0(t12_label, "h; N=", t12_bin_N_genes ), by = .(t12_bin)
][
  , t12_label := factor(t12_label, levels = gtools::mixedsort(unique(t12_label))  )
][
  
][
  ,
  ggplot(.SD,
         aes( x = as.integer(relposbin), y = glori_m6A_sum )) +
    geom_vline(data = function(x)x[region == "cds"], aes(xintercept = min(as.integer(relposbin))), linetype = 2, col = "blue" ) +
    geom_vline(data = function(x)x[region == "cds"], aes(xintercept = max(as.integer(relposbin))), linetype = 2, col = "blue" ) +
    ## points with continuous color scale
    geom_point(aes(col = t12_bin_mean), shape = 16, alpha = 0.5) +
    scale_color_gradientn(name = "Half-life [h]", colours = (brewer.pal(n = 7, name = "RdYlBu")), guide = "colorbar" ) +
    #guides(color = guide_legend(title.position = "top")) +
    ## smoothing splines with discrete color scale
    new_scale_color() +
    geom_smooth(aes(col = t12_label), method = "lm", formula = y ~ splines::bs(x, df = 40), se = F, show.legend = FALSE) +
    scale_color_manual("t12 bins", values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
    guides(color = guide_legend()) +
    ## style
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(name = "GLORI m6A level", expand = c(0, NA)) +
    theme_pubr(base_size = 8) +
    theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
    theme(legend.position  = c(0.3, 0.7), legend.direction = "horizontal", legend.background = element_blank(), legend.title = element_blank()) +
    NULL
]

```

```{r region_decay_scatterplot}

as.data.table(drach.tx)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
 # ! ( (region == "utr5" & relpos.cds >= 1) | (region == "utr3" & relpos.cds <= 2) ) , ## exclude codon / nucleotide region discrepancies
][
  , merge(.SD, canonical_transcripts, "ensembl_transcript_id")
][
  , merge(.SD, exlenTabGene, "gene")
][
  , .(glori_m6A_lvl = sum(glori_m6A_lvl))
  , by = .(gene, region)
][
  , merge(.SD, decayDataCorrected, "gene")
][
  , merge(.SD, decaybins, "gene")
][
  region != "utr5"
][ 
  , ggplot(.SD, aes( x= glori_m6A_lvl , y = endogenous_293t_time)) +
    geom_point(alpha = 0.5, aes(col = t12_bin), shape = 16) +
    geom_smooth(method = "lm") +
    stat_cor() +
    ylim(0, 0.3) +
    scale_color_manual("t12 bins", values = colorRampPalette((brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$t12_bin))) ) +
    scale_x_continuous(trans = "log10") +
    annotation_logticks(sides = "b") +
    theme_bw() +
    theme(legend.position = "none") + 
    facet_grid(region~., labeller = labeller(region = c("cds" = "CDS", "utr3" = "3'UTR"))) +
    labs(x = expression("m"^6*"A level"), y = expression("Decay rate [h"^-1*"]")) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/region_decay_scatterplot_GLORI.pdf", width = 3, height = 4.5)

```

```{r gloribin_t12_emetine_scatter}

## adapted from bakR::plotMA()

percent_change_threshold = 0.2

p.scatter <-
as.data.table(emetine$Fit$Fast_Fit$Effects_df)[
  , Avg_reads_natural := rowMeans(emetine$Fit$Data_lists$Stan_data$Avg_Reads_natural)
][
  , setnames(.SD, "XF", "gene")
][
  , merge(.SD, gloribins, "gene")
][
  , ggplot(.SD, aes( x = glori_m6A_lvl, y = L2FC_kdeg )) +
    geom_hline(yintercept = 0) +
    geom_point(aes(color = gloribin), alpha = 0.3) +
    
    scale_x_continuous(trans = "log10") +
    annotation_logticks(sides = "b")  +
    geom_smooth(method = "lm") +
    stat_cor(label.y = 1, label.x.npc = "right", hjust = 1) +
    
    geom_hline(yintercept = log(1-0.2, 2), linetype = 2) +
    geom_text(y = log(1-0.2, 2), x = Inf, label = paste("Stabilized by less than 20%:", .SD[, sum(L2FC_kdeg >=log(1-0.2, 2))], "genes"), check_overlap = TRUE, hjust = 1, vjust = -0.3 ) +
    geom_text(y = log(1-0.2, 2), x = Inf, label = paste("Stabilized by more than 20%:", .SD[, sum(L2FC_kdeg < log(1-0.2, 2))], "genes"), check_overlap = TRUE, hjust = 1, vjust =  1.3 ) +
    
    
    
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    theme_bw() +
    theme(legend.position = "none") +
    coord_cartesian(ylim = c(-1.5, 1)) +
    xlab(expression("m"^6*"A level")) + 
    ylab(bquote(L2FC(k[deg]))) + 
    ylab(expression("Log"[2]~"fold-change [decay rate]")) +
    NULL
]

p.density <-
as.data.table(emetine$Fit$Fast_Fit$Effects_df)[
  , Avg_reads_natural := rowMeans(emetine$Fit$Data_lists$Stan_data$Avg_Reads_natural)
][
  , setnames(.SD, "XF", "gene")
][
  , merge(.SD, gloribins, "gene")
][
  , ggplot(.SD, aes( x = L2FC_kdeg )) +
    geom_vline(xintercept = 0) +
    geom_density(aes(color = gloribin), alpha = 0.3)+
    
    geom_vline(xintercept = log(1-0.2, 2), linetype = 2) +

    #scale_x_continuous(trans = "log10") +
    #annotation_logticks(sides = "b")  +
    #geom_smooth(method = "lm") +
    #stat_cor(label.y = 1, label.x.npc = "right", hjust = 1) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    theme_bw() +
    theme(legend.position = "none") +
    #coord_cartesian(ylim = c(-1.5, 1)) +
    #xlab("m6A level") + 
    #ylab(bquote(L2FC(k[deg]))) + 
    coord_flip(xlim = c(-1.5, 1)) +
    NULL
]

ggarrange(p.scatter, p.density + theme(axis.text.y = element_blank(), axis.title.y = element_blank()), nrow = 1, widths = c(4,1))

ggsave("~/exports/m6A_decay_ms_share/region_emetine_scatterplot_GLORI.png", width = 5, height = 4)

```

```{r gloribin_t12_emetine}

plotdat <-
  as.data.table(emetine$Fit$Fast_Fit$Fn_Estimates)[
    , setnames(.SD, "XF", "gene")
  ][
    , merge(.SD, gloribins, "gene")
  ][
    , treatment := factor(sedit(str_extract(sample, "UT|LDE"), c("UT", "LDE"), c("mock", "emetine") ), levels = c("mock", "emetine") )
  ][
    , .(decay_rate = mean(kdeg))
    , by = .(treatment, gloribin)
  ][
    , dcast(.SD, gloribin ~ treatment, value.var = "decay_rate")
  ]

pl.scatter <-
  plotdat[
    , 
    ggplot(.SD,
           aes(x =  as.integer(gloribin), y = t12_from_decay(`mock`), col = (gloribin) )) +
      ## correlation
      #geom_smooth(method = "lm") +
      #stat_cor(label.y = Inf, vjust = 1.3*2) +
      ## scatter
      geom_point(size = 0.75) +
      #geom_point(aes(y=t12_from_decay(`mock`)), size = 2, col = palval_genotype[["mock"]]) +
      #geom_point(aes(y=t12_from_decay(`emetine`)), size = 2, col = palval_genotype[["emetine"]]) +
      
      geom_segment(aes(xend = as.integer(gloribin), y = t12_from_decay(`mock`), yend = t12_from_decay(`emetine`)), arrow = arrow(length = unit(2, "mm")) ) +
      
      #stat_poly_line(aes(y= t12_from_decay(`mock`)), col = palval_genotype[["mock"]], se = FALSE, linetype = 2) +
      #stat_poly_line(aes(y= t12_from_decay(`emetine`)), col = palval_genotype[["emetine"]], se = FALSE, linetype = 2) +
      geom_label_npc(label = "Emetine", npcx = 0.95, npcy = 0.95) +
      scale_y_continuous(trans = "log10", breaks = c(0.75, 1.5, 3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
      ## style
     # coord_cartesian(ylim = c(6, 60)) +
      scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(x = expression(m^6*A~"level [A.U.]"),
           y = "Half-life [h]"
      )
    ]

pl.bar <-
  plotdat[
    , 
    ggplot(.SD,
           aes(x =  as.integer(gloribin), y = log2(t12_from_decay(`emetine`) / t12_from_decay(`mock`)) )) +
      
      ## bar
      geom_point(aes(col = (gloribin) , fill = (gloribin))) +
      
      ## correlation
      stat_poly_line() +
      #stat_poly_eq(use_label("R2"), size = 4, coef.digits = 3, label.y = "bottom", label.x = "right", parse = TRUE) +
      
      ## scatter
      #geom_point(aes(col = (gloribin) , fill = (gloribin)), size = 0.75) +
      
      
      ## style
      #coord_cartesian(ylim = c(6, 60)) +
      scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
      scale_fill_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(x = expression(m^6*A~"level [A.U.]"),
           y = expression(Log[2]*FC)
      )
    ]


ggarrange(pl.scatter +
            theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
            theme(aspect.ratio=1) +
            NULL
          , pl.bar
          , ncol = 1, align = "v", heights = c(2,1)
          )

ggsave("~/exports/m6A_decay_ms_share/gloribin_t12_emetine.pdf", width = 3, height = 4)

```

```{r hallmark_t12_emetine}

plotdat <-
  as.data.table(emetine$Fit$Fast_Fit$Fn_Estimates)[
    , setnames(.SD, "XF", "gene")
  ][
    , treatment := factor(sedit(str_extract(sample, "UT|LDE"), c("UT", "LDE"), c("mock", "emetine") ), levels = c("mock", "emetine") )
  ][ 
    , merge(.SD
            , exlenTabGene
            , "gene")
  ][
    , merge(.SD
            , msigDT[grepl("HALLMARK_", gs_name), .(gene, gs_name)]
            , "gene"
            , allow.cartesian = TRUE
    )
  ][
    , .(decay_rate = mean(kdeg)
        , exlen = mean(exlen)
        
    )
    , by = .(gs_name, treatment)
  ][
    , exbin := cut2(exlen, g = exlenbins)
  ][
    , dcast(.SD, gs_name + exbin ~ treatment, value.var = "decay_rate")
  ][
         , merge(.SD
                 , merge(
                   glori$dt[, .(m6Alvl = sum(m6A_lvl)), by = .(gene = Gene)]
                   , msigDT, "gene")[, .(m6Alvl = mean(m6Alvl, na.rm = TRUE)), by = .(gs_name)]
                 
                 , "gs_name"
         )
       ]


pl.scatter <-
  plotdat[
    , 
    ggplot(.SD,
           aes(x =  m6Alvl, y = t12_from_decay(`mock`), col = (exbin) )) +
      ## correlation
      #geom_smooth(method = "lm") +
      #stat_cor(label.y = Inf, vjust = 1.3*2) +
      ## scatter
      geom_point(size = 0.75) +
      #geom_point(aes(y=t12_from_decay(`mock`)), size = 2, col = palval_genotype[["mock"]]) +
      #geom_point(aes(y=t12_from_decay(`emetine`)), size = 2, col = palval_genotype[["emetine"]]) +
      
      geom_segment(aes(xend = m6Alvl, y = t12_from_decay(`mock`), yend = t12_from_decay(`emetine`)), arrow = arrow(length = unit(2, "mm")) ) +
      
              geom_text_repel(data = function(x){x[gs_name %in% gs_select, ]},
                        aes(
                          label = gsub(
                            "HALLMARK_",
                            "",
                            gs_name
                          )
                          , col = exbin)
                        #,  position = position_nudge_center(y = -0.1, center_y = 18)
                        #, min.segment.length = 0
                        , size = 3
        ) +
      
      #stat_poly_line(aes(y= t12_from_decay(`mock`)), col = palval_genotype[["mock"]], se = FALSE, linetype = 2) +
      #stat_poly_line(aes(y= t12_from_decay(`emetine`)), col = palval_genotype[["emetine"]], se = FALSE, linetype = 2) +
      geom_label_npc(label = "Emetine", npcx = 0.95, npcy = 0.95) +
      scale_y_continuous(trans = "log10", breaks = c(0.75, 1.5, 3, 6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
      ## style
     # coord_cartesian(ylim = c(6, 60)) +
      scale_color_manual(values = palval_exbin) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(x = expression(m^6*A~"level [A.U.]"),
           y = "Half-life [h]"
      )
    ]

pl.bar <-
  plotdat[
    , 
    ggplot(.SD,
           aes(x =  m6Alvl, y = log2(t12_from_decay(`emetine`) / t12_from_decay(`mock`)) )) +
      
      ## bar
      geom_point(aes(col = (exbin) , fill = (exbin))) +
      
      ## correlation
      stat_poly_line() +
      #stat_poly_eq(use_label("R2"), size = 4, coef.digits = 3, label.y = "bottom", label.x = "right", parse = TRUE) +
      
      ## scatter
      #geom_point(aes(col = (gloribin) , fill = (gloribin)), size = 0.75) +
      
      
      ## style
      #coord_cartesian(ylim = c(6, 60)) +
      scale_color_manual(values = palval_exbin) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(x = expression(m^6*A~"level [A.U.]"),
           y = expression(Log[2]*FC)
      )
    ]


ggarrange(pl.scatter +
            theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
            theme(aspect.ratio=1) +
            NULL
          , pl.bar
          , ncol = 1, align = "v", heights = c(2,1)
          )

ggsave("~/exports/m6A_decay_ms_share/hallmark_t12_emetine.pdf", width = 3, height = 4)

```

## Ribosome occupancy

### miCLIP bins

```{r, fig.width = 3, fig.height = 3}

as.data.table(cods.gr)[
  ! is.na(matched_status) &
  ! codon %in% stop_codons &
    codon %in% drach_codons &
   rep(T,.N)
][
  , is_expressed := sum(cvg, na.rm = TRUE) > 10
  , by = .(seqnames)
][
  (is_expressed) 
][
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
  , melt(.SD, id.vars=c("matched_status", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(matched_status, run)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
 # , treatment := factor(treatment, levels = c("mock", "STM"), labels = c("mock", "STM2457"))
][
  experiment == "exp2" & genotype == "WT"  & occupancy > 0
][
  , ggplot(.SD, aes(x = matched_status, y = occupancy)) +
    #geom_point(aes(col = matched_status, shape = replicate)) +
   # geom_boxplot(aes(col = matched_status), notch = TRUE, fill = NA) +
    stat_summary(fun.data=mean_se, aes(col = matched_status), geom  = "pointrange") +
    scale_color_manual(values = palval_drach_status ) +
    #scale_fill_manual(values = palval_drach_status ) +
    scale_x_discrete(labels = c("no DRACH" = "no DRACH", "DRACH" = "DRACH", "m6A" = expression(m^6*A)) ) +
    facet_nested(~ treatment,  labeller = labeller(treatment = c("CTRL" = "mock", "STM" = "STM2457")) ) +
    #facet_nested(~ codon, scales ="free_y") +
    
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1) ) +
    #theme(axis.text.x = element_blank()) +
    #theme(axis.title.x = element_blank()) +
    labs(x = expression("Codon"~m^6*A~"status"), y="A-site occupancy", col = expression(m^6*A~"status") ) +
    theme(legend.position =  "none") +
    #coord_cartesian(ylim = c(NA, 50)) +
    NULL
]


ggsave("~/exports/m6A_decay_ms_share/miCLIP_occupancy.pdf", width = 3, height = 3)

```

miCLIP bins

```{r, fig.width = 6, fig.height = 6}

as.data.table(cods.gr)[
  ! is.na(matched_status)
][
  , melt(.SD, id.vars=c("codon", "matched_status", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(codon, matched_status, run)
][
  ! codon %in% stop_codons &
    codon %in% drach_codons &
   rep(T,.N)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  experiment == "exp1" & genotype == "WT" & treatment == "CTRL"
][
  , codon_strength := get_codon_strength(codon)
][
  , ggplot(.SD, aes(x = matched_status, y = occupancy)) +
    #geom_point(aes(col = matched_status, shape = replicate)) +
    stat_summary(fun.data=mean_se, aes(col = matched_status), size = 0.25) +
    
    scale_color_manual(values = palval_drach_status ) +
    # facet_nested_wrap(~ codon, scales ="free_y", nrow = 1) +
    facet_nested( ~ codon_strength + codon, scales ="free_y") +
    
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) ) +
    theme(axis.text.x = element_blank()) +
    labs(x = expression("Codon"~m^6*A~"status"), y="A-site occupancy", col = expression(m^6*A~"status") ) +
    theme(legend.position =  "none") +
    NULL
]


ggsave("~/exports/m6A_decay_ms_share/miCLIP_codons_occupancy.pdf", width = 8, height = 3)

```

### GLORI bins

```{r, fig.width = 3, fig.height = 3}

as.data.table(cods.gr)[
  ! is.na(matched_status)
][
 # codon %in% setdiff(drach_codons, stop_codons)
][
 # , is_expressed := sum(cvg, na.rm = TRUE) > 1
 # , by = .(seqnames)
][
 # (is_expressed) 
][
 # , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
 # , merge(.SD, canonical_transcripts )
][
  , melt(.SD, id.vars=c("gloribin", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(gloribin, run)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
 genotype == "WT" 
][
  , ggplot(.SD, aes(x = glori_m6A_lvl, y = occupancy, col = treatment)) +
    stat_summary(fun.data = mean_se) +
    stat_poly_line(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ] ) +
    stat_poly_eq(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ],
                 aes(label = after_stat(
                   ifelse(
                     r.squared < 0.1,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2<0.01"
                     )
                     ,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2~`=`~", round(r.squared, 2)
                     ))
                 )),
                 output.type = "numeric",
                 parse = TRUE,
                 vstep = 0.1
    ) +
    scale_color_manual(values = c("CTRL" = "grey40", "STM" = "orange") ) +
    # facet_nested( ~ experiment ) +
    
    theme_bw() +
    labs(x = expression("Codon"~ m^6*A~"stoichiometry"), y="A-site occupancy") +
    theme(legend.position =  "none") +
    scale_x_continuous(limits = c(0,1)) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/GLORI_occupancy_quant.pdf", width = 3, height = 3)

```

```{r, fig.width = 9, fig.height = 3}

as.data.table(cods.gr)[
  codon %in% setdiff(drach_codons, stop_codons)
][
#   , is_expressed := sum(cvg, na.rm = TRUE) > 1
#   , by = .(seqnames)
# ][
#   (is_expressed) 
# ][
#   , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
# ][
#   , merge(.SD, canonical_transcripts )
][
  , melt(.SD, id.vars=c("codon_strength", "gloribin", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(codon_strength, gloribin, run)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  genotype == "WT" 
][
  , ggplot(.SD, aes(x = glori_m6A_lvl, y = occupancy, col = treatment)) +
    
    stat_summary(fun.data = mean_se) +
    stat_poly_line(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ] ) +
    stat_poly_eq(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ],
                 aes(label = after_stat(
                   ifelse(
                     r.squared < 0.1,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2<0.01"
                     )
                     ,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2~`=`~", round(r.squared, 2)
                     ))
                 )),
                 output.type = "numeric",
                 parse = TRUE,
                 vstep = 0.1
    ) +
     
    scale_color_manual(values = c("CTRL" = "grey40", "STM" = "orange") ) +
    
    facet_nested(~codon_strength ) +
    
    theme_bw() +
    labs(x = expression("Codon"~ m^6*A~"stoichiometry"), y="A-site occupancy") +
    theme(legend.position =  "none") +
    scale_x_continuous(limits = c(0,1)) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/GLORI_occupancy_quant_codon_strength.pdf", width = 8, height = 3)

```

```{r, fig.width = 3, fig.height = 6}

as.data.table(cods.gr)[
  codon %in% setdiff(drach_codons, stop_codons) &
    codon_strength == "intermediate"
# ][
#   , is_expressed := sum(cvg, na.rm = TRUE) > 1
#   , by = .(seqnames)
# ][
#   (is_expressed) 
# ][
#   , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
# ][
#   , merge(.SD, canonical_transcripts )
][
  , m6A_position := factor(ifelse(codon %in% c("AAA", "AGA", "GAA", "GGA"), "Wobble", "Pos. 1 or 2"), levels = c("Wobble", "Pos. 1 or 2") )
][
  , gloribin := relevel(factor(ifelse(glori_m6A_lvl == 0, "[0]", as.character(cut2(glori_m6A_lvl,cuts = seq(0,1,0.1)))  ) ), "[0]")
][
  , melt(.SD, id.vars=c("m6A_position", "gloribin", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(m6A_position, gloribin, run)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  genotype == "WT"  
][
  , ggplot(.SD, aes(x = glori_m6A_lvl, y = occupancy, col = treatment)) +
    
    stat_summary(fun.data = mean_se) +
    stat_poly_line(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ] ) +
    stat_poly_eq(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ],
                 aes(label = after_stat(
                   ifelse(
                     r.squared < 0.1,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2<0.01"
                     )
                     ,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2~`=`~", round(r.squared, 2)
                     ))
                 )),
                 output.type = "numeric",
                 parse = TRUE,
                 vstep = 0.1
    ) +
     
    scale_color_manual(values = c("CTRL" = "grey40", "STM" = "orange") ) +
    
    facet_nested_wrap(~m6A_position, ncol = 1 ) +
    
    theme_bw() +
    labs(x = expression("Codon"~ m^6*A~"stoichiometry"), y="A-site occupancy") +
    theme(legend.position =  "none") +
    scale_x_continuous(limits = c(0,1)) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/GLORI_occupancy_quant_codon_strength_wobble.pdf", width = 3, height = 6)

```

```{r, fig.width = 6, fig.height = 9}

as.data.table(cods.gr)[
  codon %in% setdiff(drach_codons, stop_codons)
# ][
#   , is_expressed := sum(cvg, na.rm = TRUE) > 1
#   , by = .(seqnames)
# ][
#   (is_expressed) 
# ][
#   , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
# ][
#   , merge(.SD, canonical_transcripts )
][
  , melt(.SD, id.vars=c("codon", "codon_tRNA_mod", "gloribin", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(codon, codon_tRNA_mod, gloribin, run)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  codon %in% c("AAA", "AGA", "GAA", "CAA")
][
  , ggplot(.SD, aes(x = glori_m6A_lvl, y = occupancy, col = genotype)) +
    
    stat_summary(fun.data = mean_se) +
    stat_poly_line(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ] ) +
    stat_poly_eq(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ],
                 aes(label = after_stat(
                   ifelse(
                     r.squared < 0.1,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2<0.01"
                     )
                     ,
                     paste0(
                     "m~`=`~", round(b_1, 2)
                     #, "*`,`~italic(R)^2~`=`~", round(r.squared, 2)
                     ))
                 )),
                 output.type = "numeric",
                 parse = TRUE,
                 vstep = 0.1
    ) +
    scale_color_manual(values = palval_genotype ) +
    
    facet_nested(codon ~  treatment, scale = "free_y",  labeller = labeller(treatment = c("CTRL" = "mock", "STM" = "STM2457"))) +
    
    theme_bw() +
    labs(x = expression("Codon"~ m^6*A~"stoichiometry"), y="A-site occupancy") +
    theme(legend.position =  "none") +
    scale_x_continuous(limits = c(0,1)) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/GLORI_occupancy_quant_codon_strength_mcm5s2U.pdf", width = 4.5, height = 6)

```

```{r, fig.width = 4.5, fig.height = 3}

as.data.table(cods.gr)[
  codon %in% c("AGA", "GGA") 
][
#   , is_expressed := sum(cvg, na.rm = TRUE) > 1
#   , by = .(seqnames)
# ][
#   (is_expressed) 
# ][
#   , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
# ][
#   , merge(.SD, canonical_transcripts )
][
  , melt(.SD, id.vars=c("codon", "codon_tRNA_mod", "gloribin", "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(codon, codon_tRNA_mod, gloribin, run)
][
 , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
   genotype == "WT"
][
  , ggplot(.SD, aes(x = glori_m6A_lvl, y = occupancy, col = codon)) +
    
    stat_summary(fun.data = mean_se) +
    stat_poly_line(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ] ) +
    stat_poly_eq(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ],
                 aes(label = after_stat(
                   ifelse(
                     r.squared < 0.1,
                     paste0(
                     "m~`=`~", round(b_1, 2),
                     "*`,`~italic(R)^2<0.01")
                     ,
                     paste0(
                     "m~`=`~", round(b_1, 2),
                     "*`,`~italic(R)^2~`=`~", round(r.squared, 2)))
                 )),
                 output.type = "numeric",
                 parse = TRUE,
                 vstep = 0.1
    ) +
    scale_color_manual(values = palval_reporter_codons ) +
    
    facet_nested( ~  treatment, scale = "free_y",  labeller = labeller(treatment = c("CTRL" = "mock", "STM" = "STM2457"))) +
    
    theme_bw() +
    labs(x = expression("Codon"~ m^6*A~"stoichiometry"), y="A-site occupancy") +
    theme(legend.position =  "none") +
    scale_x_continuous(limits = c(0,1)) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/GLORI_occupancy_quant_codon_strength_mcm5s2U_GGA.pdf", width = 4.5, height = 3)

```

### Codons general

```{r codons_occupancy_KO, fig.width = 18, fig.height = 3}

p <-
  as.data.table(cods.gr)[
   codon_tRNA_mod != "stop"
  ][
    , melt(.SD, id.vars= c( c("codon", "codon_tRNA_mod"), "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
  ][
    , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
    , by = .(codon, codon_tRNA_mod, run)
  ][
    , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
  ][
    experiment == "exp1" & treatment == "CTRL"
  ][
    , ggplot(.SD, aes(x = codon, y = occupancy, col = genotype)) +
      #geom_point(aes(col = gloribin, shape = replicate)) +
      #stat_summary(fun.data = mean_se, size = 0.25, position = position_dodge(width = 0.5)) +
      
      scale_color_manual(values = palval_genotype) +
      
      #geom_text(y = Inf, aes(label = ifelse(N>1e4,"N>10k",paste0("N=",N)) ), angle = 90, hjust = 1, size = 2,check_overlap = TRUE) +
      
      
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
      #theme(axis.text.x = element_blank()) +
      theme(legend.position =  "none") +
      
      labs(x = "Codon", y="A-site occupancy", col = "Genotype") +
      
      NULL
  ]

## scaled facets from: https://stackoverflow.com/a/61127086
p + 
  facet_nested(~ codon_tRNA_mod , scales ="free_x", space = "free") +
  stat_summary(fun.data = mean_se, size = 0.25, position = position_dodge(width = 0.75)) + 
   ggforce::facet_row( ~ codon_tRNA_mod  , scales = "free", space = "free") +
   theme(panel.spacing.x = unit( 0 , units = "lines") ) +
  NULL

ggsave("~/exports/m6A_decay_ms_share/codons_occupancy_KO.pdf", width = 9, height = 2)

```

## Reporter

```{r DRACH_reporter_mock, fig.width = 2, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
        txn_inh == "ActD" & 
    transl_inh == "mock" &
    m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = reporter)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.11) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_reporter) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.2)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    NULL
]
  
ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_mock.pdf"), width = 2, height = 2, units = "in")

```

```{r DRACH_reporter_mock_TLD, fig.width = 2, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
        txn_inh == "TLD" & 
    transl_inh == "mock" &
    m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = reporter)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.11) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_reporter) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.2)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    NULL
]
  
ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_mock_TLD.pdf"), width = 2, height = 2, units = "in")

```

```{r DRACH_reporter_mock2, fig.width = 4, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
   #     txn_inh == "ActD" & 
    transl_inh == "mock" &
    m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = reporter)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.11) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_reporter) +
    facet_nested( ~ txn_inh , scales = "free_y", labeller = labeller(txn_inh = c("ActD" = "Actinomycin D", "TLD"="Triptolide") ) )  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.3)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    NULL
]
  
ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_mock2.pdf"), width = 2.5, height = 2, units = "in")

```

```{r DRACH_reporter_STM2, fig.width = 4, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
    txn_inh == "ActD" & 
    transl_inh == "mock" &
    #m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , reporter := factor(reporter, levels = names(labeller_reporter), labels = labeller_reporter)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = m6A_inh)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.12) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_m6A_inh) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    #facet_nested(~reporter, strip = ) +
    facet_nested(~ reporter,
                 strip = strip_nested(background_x = elem_list_rect(fill= palval_reporter[c("C4", "C2", "C1")] ) , by_layer_x = FALSE),
                 labeller = label_parsed ) +
    NULL
]

ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_STM2.pdf"), width = 4, height = 2, units = "in")

```

```{r DRACH_reporter_TLD, fig.width = 4, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
    txn_inh == "TLD" & 
    transl_inh == "mock" &
    #m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , reporter := factor(reporter, levels = names(labeller_reporter), labels = labeller_reporter)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = m6A_inh)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.12) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_m6A_inh) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    #facet_nested(~reporter, strip = ) +
    facet_nested(~ reporter,
                 strip = strip_nested(background_x = elem_list_rect(fill= palval_reporter[c("C4", "C2", "C1")] ) , by_layer_x = FALSE),
                 labeller = label_parsed ) +
    NULL
]

ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_TLD.pdf"), width = 4, height = 2, units = "in")

```

```{r , fig.width = 4, fig.height = 4}

qpcr_data_complete[
  genotype == "WT" &
   # txn_inh == "TLD" & 
    transl_inh == "mock" &
    #m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , reporter := factor(reporter, levels = names(labeller_reporter), labels = labeller_reporter)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = m6A_inh)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.11) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_m6A_inh) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.3)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    #facet_nested(~reporter, strip = ) +
    facet_nested(txn_inh ~ reporter,
                 strip = strip_nested(background_x = elem_list_rect(fill= palval_reporter[c("C4", "C2", "C1")] ) , by_layer_x = FALSE),
                 labeller = labeller(txn_inh = c("ActD" = "Actinomycin D", "TLD"="Triptolide"), reporter = label_parsed )  ) +
    NULL
]

```

```{r, fig.width = 4, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
    transl_inh == "mock" &
    #m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C3", "C4") &
    rep(TRUE, .N)
][
  , reporter := factor(reporter, levels = names(labeller_reporter), labels = labeller_reporter)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = m6A_inh)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.12) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_m6A_inh) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    #facet_nested(~reporter, strip = ) +
    facet_nested(~ reporter,
                 strip = strip_nested(background_x = elem_list_rect(fill= palval_reporter[c("C4", "C2", "C1")] ) , by_layer_x = FALSE),
                 labeller = label_parsed ) +
    NULL
]

```

```{r DRACH_reporter_EM2, fig.width = 3, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
    txn_inh == "ActD" & 
    transl_inh != "CHX" &
    m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C1", "C2", "C4") & 
    rep(TRUE, .N)
][
  , reporter := factor(reporter, levels = names(labeller_reporter), labels = labeller_reporter)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = transl_inh)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.1 ) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_transl_inh) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    facet_nested(~reporter, strip = ) +
    facet_nested(~ reporter,
                 strip = strip_nested(background_x = elem_list_rect(fill= palval_reporter[c("C4", "C2", "C1")] ) , by_layer_x = FALSE),
                 labeller = label_parsed ) +
    NULL
]

ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_EM2.pdf"), width = 3, height = 2.5, units = "in")

```

```{r DRACH_reporter_CHX2, fig.width = 3, fig.height = 2}

qpcr_data_complete[
  genotype == "WT" &
    txn_inh == "ActD" & 
    transl_inh != "EM" &
    m6A_inh == "mock" &
    target == "Neon2" &
   reporter %in% c("C1", "C2", "C4") & 
    rep(TRUE, .N)
][
  #, reporter := factor(reporter, levels = names(labeller_reporter))
][
  , reporter := factor(reporter, levels = names(labeller_reporter), labels = labeller_reporter)
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = transl_inh)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression)) ), label.x = "right",
                  aes(label = paste0(signif(after_stat(log(2)/k_estimate), digits = 2), "h") ), vstep = 0.1 ) +
    scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_transl_inh) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    #facet_nested(m6A_inh ~ transl_inh , scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(0, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    facet_nested(~reporter, strip = ) +
    facet_nested(~ reporter,
                 strip = strip_nested(background_x = elem_list_rect(fill= palval_reporter[c("C4", "C2", "C1")] ) , by_layer_x = FALSE),
                 labeller = label_parsed ) +
    NULL
]

ggsave(paste0("/mnt/c/Users/BastianLinder/Nextcloud/inbox/exports/m6A_decay_ms_share/DRACH_reporter_CHX2.pdf"), width = 3, height = 2.5, units = "in")

```

```{r DRACH_reporter_AGA, fig.width = 6, fig.height = 3}

qpcr_data_complete[
 # genotype == "WT" &
  txn_inh == "ActD" & 
    transl_inh == "mock" &
    # m6A_inh == "mock" &
    target == "Neon2" &
    reporter %in% c("C2") &
    #  ! grepl("ELP", genotype) &
    rep(TRUE, .N)
][
  , m6A_inh := factor(m6A_inh, levels = c("mock", "STM"), labels = c("mock", "STM2457"))
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = genotype)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    geom_decay_line(by = c("target", "reporter", "transl_inh",  "m6A_inh", "genotype")) +
    geom_decay_text(by = c("target", "reporter", "transl_inh",  "m6A_inh", "genotype"), vjust_var = "genotype") +
    #stat_poly_line(formula = y ~ poly(x, 2, raw = TRUE) ) +
    #scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_genotype) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    facet_nested(reporter ~ m6A_inh , labeller = labeller(reporter = labeller_reporter_codons)
                 #, strip = strip_nested(background_x = elem_list_rect(fill= palval_m6A_inh[c("mock", "STM")] ))
                 ,  scales = "free_y")  +
    theme_bw() +
    #coord_cartesian(ylim = c(0, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/DRACH_reporter_AGA.pdf", width = 4.5, height = 2.5)

```

```{r DRACH_reporter_GGA, fig.width = 4.5, fig.height = 4}

qpcr_data_complete[
 # genotype == "WT" &
  txn_inh == "ActD" & 
    transl_inh == "mock" &
    # m6A_inh == "mock" &
    target == "Neon2" &
     reporter %in% c("C2" ,"C5") &
    #  ! grepl("ELP", genotype) &
    rep(TRUE, .N)
][
  , m6A_inh := factor(m6A_inh, levels = c("mock", "STM"), labels = c("mock", "STM2457"))
][
  genotype == "WT"
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = reporter)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    
    
   # geom_decay_line(by = c("target", "reporter", "transl_inh",  "m6A_inh", "genotype")) +
  #  geom_decay_text(by = c("target", "reporter", "transl_inh",  "m6A_inh", "genotype"), vjust_var = "genotype") +
    
    geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
    stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression))  ), aes(label = paste0(round(after_stat(log(2)/k_estimate), digits = 1), "h") )) +

    
    #scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_reporter ) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    facet_nested( ~ m6A_inh , labeller = labeller(reporter = labeller_reporter_codons), scales = "free_y")  +
    theme_bw() +
    coord_cartesian(ylim = c(NA, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/DRACH_reporter_GGA.pdf", width = 4.5, height = 2.5)

```

```{r DRACH_reporter_GGA_mcm5s2U, fig.width = 4.5, fig.height = 4}

qpcr_data_complete[
 # genotype == "WT" &
  txn_inh == "ActD" & 
    transl_inh == "mock" &
     m6A_inh == "mock" &
    target == "Neon2" &
     reporter %in% c("C2" ,"C5") &
    #  ! grepl("ELP", genotype) &
    rep(TRUE, .N)
][
 # genotype == "WT"
][
  , ggplot(.SD,
           aes(x = time_num, y = expression, col = genotype)) +
    geom_hline(yintercept = 1, linetype = 2) +
    #geom_quasirandom(aes(shape= paste(experiment, replicate)), width = 0.2) +
    stat_summary(fun.data = mean_sd) +
    
    
    geom_decay_line(by = c("target", "reporter", "transl_inh",  "m6A_inh", "genotype")) +
    geom_decay_text(by = c("target", "reporter", "transl_inh",  "m6A_inh", "genotype"), vjust_var = "genotype") +
    
  #  geom_smooth(method = "nls", formula = y ~ I(a*exp(-k*x)), method.args = list(  start = c(k = 0, a = max(expression)) ), se = FALSE) +
   # stat_fit_tidy(method = "nls", method.args = list(formula = y ~ I(a*exp(-k*x)), start = c(k = 0, a = max(expression))  ), aes(label = paste0(round(after_stat(log(2)/k_estimate), digits = 1), "h") )) +

    
    #scale_shape_manual(values= c(1:3,5:20)) +
    scale_color_manual(values = palval_genotype ) +
    scale_y_continuous(breaks = c(0, 0.5, 1)) +
    facet_nested( ~ reporter, labeller = labeller(reporter = labeller_reporter_codons), scales = "free_y")  +
    theme_bw() +
    #coord_cartesian(ylim = c(NA, 1.4)) +
    theme(legend.position = "none") +
    labs(x = "Time [h]", y = "Relative expression") +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/DRACH_reporter_GGA_mcm5s2U.pdf", width = 4.5, height = 2.5)

```

## endogenous mRNAs

### GLORI bin

```{r, fig.width = 3, fig.height = 3}

mcm5s2U_SLAM$slam[
  region == "exonic"
][
  , ensembl_gene_id := gsub("\\..*$","",gene_id)
][
  , merge(.SD, exlenTabGene, "ensembl_gene_id")
][
  , merge(.SD, gloribins, "gene")
][
  , .(Readcount = sum(Readcount) , min2 = sum(min2))
  , by = .(gloribin, run)
][
  , NTR := min2 / Readcount
][
  , merge(.SD, mcm5s2U_SLAM$meta, "run")
][
  , expression := ifelse(label_time_num == 0, 1, 1 - NTR)
][
  , {
    fit = lm(formula = log2(expression) ~ label_time_num, data = .SD)
    coeff = summary(fit)$coefficients
    data.table(decay_rate = -1 * coeff["label_time_num", "Estimate"] )
  }
  , by = .(gloribin, genotype )
][
  , dcast(.SD, gloribin ~ genotype, value.var = "decay_rate")
][
  , ggplot(.SD,
           aes(x =  as.integer(gloribin), y = t12_from_decay(`WT`), col = (gloribin) )) +
    ## correlation
    #geom_smooth(method = "lm") +
    #stat_cor(label.y = Inf, vjust = 1.3*2) +
    ## scatter
    geom_point(size = 0.75) +
    #geom_point(aes(y=t12_from_decay(`WT`)), size = 2, col = palval_genotype[["WT"]]) +
    #geom_point(aes(y=t12_from_decay(`CTU2 KO`)), size = 2, col = palval_genotype[["CTU2 KO"]]) +
    
    geom_segment(aes(xend = as.integer(gloribin), y = t12_from_decay(`WT`), yend = t12_from_decay(`CTU2 KO`)), arrow = arrow(length = unit(2, "mm")) ) +
    geom_label_npc(label = "CTU2 KO", npcx = 0.95, npcy = 0.95) +
    #stat_poly_line(aes(y= t12_from_decay(`WT`)), col = palval_genotype[["WT"]], se = FALSE, linetype = 2) +
    #stat_poly_line(aes(y= t12_from_decay(`CTU2 KO`)), col = palval_genotype[["CTU2 KO"]], se = FALSE, linetype = 2) +
    scale_y_continuous(trans = "log10", breaks = c(6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
    ## style
    #coord_cartesian(ylim = c(6, 60)) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = expression(m^6*A~"level [A.U.]"),
         y = "Half-life [h]"
    ) 
]

#ggsave("~/exports/m6A_decay_ms_share/gloribin_t12_CTU2.pdf", width = 3, height = 3)

```

```{r gloribin_t12_CTU2, fig.width = 3, fig.height = 4}

plotdat <-
  mcm5s2U_SLAM$slam[
    region == "exonic"
  ][
    , ensembl_gene_id := gsub("\\..*$","",gene_id)
  ][
    , merge(.SD, exlenTabGene, "ensembl_gene_id")
  ][
  , merge(.SD, gloribins, "gene")
  ][
    , .(Readcount = sum(Readcount)
        , min2 = sum(min2)
    )
    , by = .(gloribin, run)
  ][
    , NTR := min2 / Readcount
  ][
    , merge(.SD, mcm5s2U_SLAM$meta, "run")
  ][
    , expression := ifelse(label_time_num == 0, 1, 1 - NTR)
  ][
    , {
      fit = lm(formula = log2(expression) ~ label_time_num, data = .SD)
      coeff = summary(fit)$coefficients
      data.table(decay_rate = -1 * coeff["label_time_num", "Estimate"] )
    }
    , by = .(gloribin, genotype )
  ][
    , dcast(.SD, gloribin ~ genotype, value.var = "decay_rate")
  ]

pl.scatter <-
  plotdat[
    , 
    ggplot(.SD,
           aes(x =  as.integer(gloribin), y = t12_from_decay(`WT`), col = (gloribin) )) +
      ## correlation
      #geom_smooth(method = "lm") +
      #stat_cor(label.y = Inf, vjust = 1.3*2) +
      ## scatter
      geom_point(size = 0.75) +
      #geom_point(aes(y=t12_from_decay(`WT`)), size = 2, col = palval_genotype[["WT"]]) +
      #geom_point(aes(y=t12_from_decay(`CTU2 KO`)), size = 2, col = palval_genotype[["CTU2 KO"]]) +
      
      geom_segment(aes(xend = as.integer(gloribin), y = t12_from_decay(`WT`), yend = t12_from_decay(`CTU2 KO`)), arrow = arrow(length = unit(2, "mm")) ) +
      
      #stat_poly_line(aes(y= t12_from_decay(`WT`)), col = palval_genotype[["WT"]], se = FALSE, linetype = 2) +
      #stat_poly_line(aes(y= t12_from_decay(`CTU2 KO`)), col = palval_genotype[["CTU2 KO"]], se = FALSE, linetype = 2) +
      geom_label_npc(label = "CTU2 KO", npcx = 0.98, npcy = 0.98) +
      scale_y_continuous(trans = "log10", breaks = c(6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
      ## style
      coord_cartesian(ylim = c(6, 60)) +
      scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(x = expression(m^6*A~"level [A.U.]"),
           y = "Half-life [h]"
      )
    ]

pl.bar <-
  plotdat[
    , 
    ggplot(.SD,
           aes(x =  as.integer(gloribin), y = log2(t12_from_decay(`CTU2 KO`) / t12_from_decay(`WT`)) )) +
      
      ## bar
      geom_point(aes(col = (gloribin) , fill = (gloribin))) +
      
      ## correlation
      stat_poly_line() +
     # stat_poly_eq(use_label("R2"), size = 4, coef.digits = 3, label.y = "top", label.x = "right", parse = TRUE) +
      
      ## scatter
      #geom_point(aes(col = (gloribin) , fill = (gloribin)), size = 0.75) +
      
      
      ## style
      #coord_cartesian(ylim = c(6, 60)) +
      scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
      scale_fill_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
      theme_bw() +
      theme(legend.position = "none") +
      labs(x = expression(m^6*A~"level [A.U.]"),
           y = expression(Log[2]*FC)
      )
    ]


ggarrange(pl.scatter +
            theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
            theme(aspect.ratio=1) +
            NULL
          , pl.bar
          , ncol = 1, align = "v", heights = c(2,1)
          )

ggsave("~/exports/m6A_decay_ms_share/gloribin_t12_CTU2.pdf", width = 3, height = 4)

```

```{r gloribin_t12_ELP1, fig.width = 3, fig.height = 4}


plotdat <-
mcm5s2U_SLAM$slam[
  region == "exonic"
][
  , ensembl_gene_id := gsub("\\..*$","",gene_id)
][
  , merge(.SD, exlenTabGene, "ensembl_gene_id")
][
  , merge(.SD, gloribins, "gene")
][
  , .(Readcount = sum(Readcount)
      , min2 = sum(min2)
  )
  , by = .(gloribin, run)
][
       , NTR := min2 / Readcount
     ][
       , merge(.SD, mcm5s2U_SLAM$meta, "run")
     ][
       , expression := ifelse(label_time_num == 0, 1, 1 - NTR)
     ][
       , {
         fit = lm(formula = log2(expression) ~ label_time_num, data = .SD)
         coeff = summary(fit)$coefficients
         data.table(decay_rate = -1 * coeff["label_time_num", "Estimate"] )
       }
       , by = .(gloribin, genotype )
     ][
       , dcast(.SD, gloribin ~ genotype, value.var = "decay_rate")
     ]



pl.scatter <-
plotdat[
       , 
       ggplot(.SD,
                aes(x =  as.integer(gloribin), y = t12_from_decay(`WT`), col = (gloribin) )) +
         ## correlation
         #geom_smooth(method = "lm") +
         #stat_cor(label.y = Inf, vjust = 1.3*2) +
         ## scatter
         geom_point(size = 0.75) +
         #geom_point(aes(y=t12_from_decay(`WT`)), size = 2, col = palval_genotype[["WT"]]) +
         #geom_point(aes(y=t12_from_decay(`ELP1 KO`)), size = 2, col = palval_genotype[["ELP1 KO"]]) +
         
         geom_segment(aes(xend = as.integer(gloribin), y = t12_from_decay(`WT`), yend = t12_from_decay(`ELP1 KO`)), arrow = arrow(length = unit(2, "mm")) ) +
         
         #stat_poly_line(aes(y= t12_from_decay(`WT`)), col = palval_genotype[["WT"]], se = FALSE, linetype = 2) +
         #stat_poly_line(aes(y= t12_from_decay(`ELP1 KO`)), col = palval_genotype[["ELP1 KO"]], se = FALSE, linetype = 2) +
         geom_label_npc(label = "ELP1 KO", npcx = 0.98, npcy = 0.98) +
         scale_y_continuous(trans = "log10", breaks = c(6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
         ## style
         coord_cartesian(ylim = c(6, 60)) +
         scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
         theme_bw() +
         theme(legend.position = "none") +
         labs(x = expression(m^6*A~"level [A.U.]"),
              y = "Half-life [h]"
         )
       
       
     ]

pl.bar <-
plotdat[
       , 
       ggplot(.SD,
                aes(x =  as.integer(gloribin), y = log2(t12_from_decay(`ELP1 KO`) / t12_from_decay(`WT`)) )) +
     
          ## bar
          geom_point(aes(col = (gloribin) , fill = (gloribin))) +
         
         ## correlation
         stat_poly_line() +
         # stat_poly_eq(use_label("R2"), size = 4, coef.digits = 3, label.y = "top", label.x = "right", parse = TRUE) +
         
         ## scatter
         #geom_point(aes(col = (gloribin) , fill = (gloribin)), size = 0.75) +
         
         
         ## style
         #coord_cartesian(ylim = c(6, 60)) +
         scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
         scale_fill_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
         theme_bw() +
         theme(legend.position = "none") +
         labs(x = expression(m^6*A~"level [A.U.]"),
              y = expression(Log[2]*FC)
         )
       
       
     ]


ggarrange(pl.scatter +
            theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
            #theme(aspect.ratio=1) +
            NULL
          , 
          pl.bar
          , 
          ncol = 1, align = "v", heights = c(2,1)
          )
ggsave("~/exports/m6A_decay_ms_share/gloribin_t12_ELP1.pdf", width = 3, height = 4)

```

### Cancer Hallmark

```{r}
## geneset categories
gs_signaling = msigDT[grepl("HALLMARK|KEGG", gs_name) & grepl("SIGNALING|RESPONSE", gs_name) & !grepl("_UP$|_DN$|UNFOLDED", gs_name), unique(gs_name) ]
gs_metabolic = msigDT[grepl("HALLMARK|KEGG", gs_name) & grepl("METABOLISM|GLYCOLYSIS|SECRETION|OXYGEN|OXIDATIVE|CHOLESTEROL", gs_name), unique(gs_name) ]

gs_select = grep("TGF|TNF|NOTCH|MYC|OXIDATIVE|SECRETION|FATTY|INFLAMMATORY_RESPONSE|HEDGEHOG", c(gs_signaling, gs_metabolic), value = TRUE)

```

```{r hallmark_t12_m6A, fig.width = 3, fig.height = 3}

  with(mcm5s2U_SLAM,
       slam[
         region == "exonic"
       ][ 
         , merge(.SD
                 , exlenTabGene
                 , "gene_id")
       ][
         , merge(.SD
                 , msigDT[grepl("HALLMARK_", gs_name), .(gene, gs_name)]
                 , "gene"
                 , allow.cartesian = TRUE
         )
       ][
         , .(Readcount = sum(Readcount)
             , min2 = sum(min2)
             , exlen = mean(exlen)
         )
         , by = .(run, region, gs_name)
       ][
         , NTR := min2/Readcount
       ][
         , merge(.SD, meta, "run")
       ][
         , expression := ifelse(label_time_num == 0, 1, 1 - NTR)
       ][
         , exbin := cut2(exlen, g = exlenbins)
       ][
         , {
           fit = lm(formula = log2(expression) ~ label_time_num, data = .SD)
           coeff = summary(fit)$coefficients
           data.table(decay_rate = -1 * coeff["label_time_num", "Estimate"] )
         }
         , by = .(gs_name, region, exlen, exbin, genotype )
       ][
         , dcast(.SD, gs_name + region + exlen + exbin ~ genotype, value.var = "decay_rate")
       ][
         , merge(.SD
                 , merge(
                   glori$dt[, .(m6Alvl = sum(m6A_lvl)), by = .(gene = Gene)]
                   , msigDT, "gene")[, .(m6Alvl = mean(m6Alvl, na.rm = TRUE)), by = .(gs_name)]
                 
                 , "gs_name"
         )
       ])[
      ,
      ggplot(.SD,
             aes(x =  m6Alvl, y = t12_from_decay(`WT`) )) +
        ## correlation
        #geom_smooth(method = "lm") +
        #stat_cor(label.y = Inf, vjust = 1.3*2) +
        #stat_poly_line( ) +
        #stat_poly_eq() +
        ## scatter
        geom_point(aes(col = (exbin))) +
        #geom_point(aes(y=t12_from_decay(`WT`)), size = 2, col = palval_genotype[["WT"]]) +
        #geom_point(aes(y=t12_from_decay(`CTU2 KO`)), size = 2, col = palval_genotype[["CTU2 KO"]]) +
        
        
        geom_text_repel(data = function(x){x[gs_name %in% gs_select, ]},
                        aes(
                          label = gsub(
                            "HALLMARK_",
                            "",
                            gs_name
                          )
                          , col = exbin)
                        #,  position = position_nudge_center(y = -0.1, center_y = 18)
                        #, min.segment.length = 0
                        , size = 3
        ) +
        

        #stat_poly_line(aes(y= t12_from_decay(`CTU2 KO`)), col = palval_genotype[["CTU2 KO"]], se = FALSE, linetype = 2) +
        scale_y_continuous(trans = "log10", breaks = c(6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
        ## style
        coord_cartesian(ylim = c(3.8, 60)) +
        scale_color_manual(values = palval_exbin ) +
        theme_bw() +
        theme(legend.position = "none") +
        labs(x = expression(m^6*A~"level [A.U.]"),
             y = "Half-life [h]"
        ) +
        
        NULL
  ]
  
ggsave("~/exports/m6A_decay_ms_share/hallmark_t12_m6A.pdf", width = 3, height = 3)

```

```{r hallmark_t12_CTU2, fig.width = 3, fig.height = 4}

plotdat <-
  with(mcm5s2U_SLAM,
       slam[
         region == "exonic"
       ][ 
         , merge(.SD
                 , exlenTabGene
                 , "gene_id")
       ][
         , merge(.SD
                 , msigDT[grepl("HALLMARK_", gs_name), .(gene, gs_name)]
                 , "gene"
                 , allow.cartesian = TRUE
         )
       ][
         , .(Readcount = sum(Readcount)
             , min2 = sum(min2)
             , exlen = mean(exlen)
         )
         , by = .(run, region, gs_name)
       ][
         , NTR := min2/Readcount
       ][
         , merge(.SD, meta, "run")
       ][
         , expression := ifelse(label_time_num == 0, 1, 1 - NTR)
       ][
         , exbin := cut2(exlen, g = exlenbins)
       ][
         , {
           fit = lm(formula = log2(expression) ~ label_time_num, data = .SD)
           coeff = summary(fit)$coefficients
           data.table(decay_rate = -1 * coeff["label_time_num", "Estimate"] )
         }
         , by = .(gs_name, region, exlen, exbin, genotype )
       ][
         , dcast(.SD, gs_name + region + exlen + exbin ~ genotype, value.var = "decay_rate")
       ][
         , merge(.SD
                 , merge(
                   glori$dt[, .(m6Alvl = sum(m6A_lvl)), by = .(gene = Gene)]
                   , msigDT, "gene")[, .(m6Alvl = mean(m6Alvl, na.rm = TRUE)), by = .(gs_name)]
                 
                 , "gs_name"
         )
       ])

pl.scatter <- 
  plotdat[
      ,
      ggplot(.SD,
             aes(x =  m6Alvl, y = t12_from_decay(`WT`), col = (exbin) )) +
        ## correlation
        #geom_smooth(method = "lm") +
        #stat_cor(label.y = Inf, vjust = 1.3*2) +
        ## scatter
        geom_point(size = 0.75) +
        #geom_point(aes(y=t12_from_decay(`WT`)), size = 2, col = palval_genotype[["WT"]]) +
        #geom_point(aes(y=t12_from_decay(`CTU2 KO`)), size = 2, col = palval_genotype[["CTU2 KO"]]) +
        
        geom_segment(aes(xend = m6Alvl, y = t12_from_decay(`WT`), yend = t12_from_decay(`CTU2 KO`)), arrow = arrow(length = unit(2, "mm")) ) +
        
        geom_text_repel(data = function(x){x[gs_name %in% gs_select, ]},
                        aes(
                          label = gsub(
                            "HALLMARK_",
                            "",
                            gs_name
                          )
                          , col = exbin)
                        #,  position = position_nudge_center(y = -0.1, center_y = 18)
                        #, min.segment.length = 0
                        , size = 3
        ) +
        #stat_poly_line(aes(y= t12_from_decay(`WT`)), col = palval_genotype[["WT"]], se = FALSE, linetype = 2) +
        #stat_poly_line(aes(y= t12_from_decay(`CTU2 KO`)), col = palval_genotype[["CTU2 KO"]], se = FALSE, linetype = 2) +
        scale_y_continuous(trans = "log10", breaks = c(6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
        ## style
        coord_cartesian(ylim = c(3.8, 60)) +
        scale_color_manual(values = palval_exbin ) +
        theme_bw() +
        geom_label_npc(label = "CTU2 KO", npcx = 0.98, npcy = 0.98) +
        theme(legend.position = "none") +
        labs(x = expression(m^6*A~"level [A.U.]"),
             y = "Half-life [h]"
        ) +
        
        NULL
      
    ]

pl.bar <-
  plotdat[, ggplot(.SD,
                     aes(x =  m6Alvl, y = log2(t12_from_decay(`CTU2 KO`) / t12_from_decay(`WT`)) )) +
          ## bar
          geom_point(aes(col = (exbin) , fill = (exbin))) +
              
              ## correlation
              stat_poly_line() +
              # stat_poly_eq(use_label("R2"), size = 4, coef.digits = 3, label.y = "top", label.x = "right", parse = TRUE) +
              
              ## scatter
              #geom_point(aes(col = (gloribin) , fill = (gloribin)), size = 0.75) +
              
              
              ## style
              #coord_cartesian(ylim = c(6, 60)) +
              scale_color_manual(values = palval_exbin ) +
              scale_fill_manual(values = palval_exbin ) +
              theme_bw() +
              theme(legend.position = "none") +
              labs(x = expression(m^6*A~"level [A.U.]"),
                   y = expression(Log[2]*FC)
              )]
  


ggarrange(pl.scatter +
            theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
            #theme(aspect.ratio=1) +
            NULL
          , 
          pl.bar
          , 
          ncol = 1, align = "v", heights = c(2,1)
          )

ggsave("~/exports/m6A_decay_ms_share/hallmark_t12_CTU2.pdf", width = 3, height = 4)

```

```{r hallmark_t12_ELP1, fig.width = 3, fig.height = 4}

plotdat <-
  with(mcm5s2U_SLAM,
       slam[
         region == "exonic"
       ][ 
         , merge(.SD
                 , exlenTabGene
                 , "gene_id")
       ][
         , merge(.SD
                 , msigDT[grepl("HALLMARK_", gs_name), .(gene, gs_name)]
                 , "gene"
                 , allow.cartesian = TRUE
         )
       ][
         , .(Readcount = sum(Readcount)
             , min2 = sum(min2)
             , exlen = mean(exlen)
         )
         , by = .(run, region, gs_name)
       ][
         , NTR := min2/Readcount
       ][
         , merge(.SD, meta, "run")
       ][
         , expression := ifelse(label_time_num == 0, 1, 1 - NTR)
       ][
         , exbin := cut2(exlen, g = exlenbins)
       ][
         , {
           fit = lm(formula = log2(expression) ~ label_time_num, data = .SD)
           coeff = summary(fit)$coefficients
           data.table(decay_rate = -1 * coeff["label_time_num", "Estimate"] )
         }
         , by = .(gs_name, region, exlen, exbin, genotype )
       ][
         , dcast(.SD, gs_name + region + exlen + exbin ~ genotype, value.var = "decay_rate")
       ][
         , merge(.SD
                 , merge(
                   glori$dt[, .(m6Alvl = sum(m6A_lvl)), by = .(gene = Gene)]
                   , msigDT, "gene")[, .(m6Alvl = mean(m6Alvl, na.rm = TRUE)), by = .(gs_name)]
                 
                 , "gs_name"
         )
       ])

pl.scatter <- 
  plotdat[
      ,
      ggplot(.SD,
             aes(x =  m6Alvl, y = t12_from_decay(`WT`), col = (exbin) )) +
        ## correlation
        #geom_smooth(method = "lm") +
        #stat_cor(label.y = Inf, vjust = 1.3*2) +
        ## scatter
        geom_point(size = 0.75) +
        #geom_point(aes(y=t12_from_decay(`WT`)), size = 2, col = palval_genotype[["WT"]]) +
        #geom_point(aes(y=t12_from_decay(`ELP1 KO`)), size = 2, col = palval_genotype[["ELP1 KO"]]) +
        
        geom_segment(aes(xend = m6Alvl, y = t12_from_decay(`WT`), yend = t12_from_decay(`ELP1 KO`)), arrow = arrow(length = unit(2, "mm")) ) +
        
        geom_text_repel(data = function(x){x[gs_name %in% gs_select, ]},
                        aes(
                          label = gsub(
                            "HALLMARK_",
                            "",
                            gs_name
                          )
                          , col = exbin)
                        #,  position = position_nudge_center(y = -0.1, center_y = 18)
                        #, min.segment.length = 0
                        , size = 3
        ) +
        #stat_poly_line(aes(y= t12_from_decay(`WT`)), col = palval_genotype[["WT"]], se = FALSE, linetype = 2) +
        #stat_poly_line(aes(y= t12_from_decay(`ELP1 KO`)), col = palval_genotype[["ELP1 KO"]], se = FALSE, linetype = 2) +
        scale_y_continuous(trans = "log10", breaks = c(6, 12, 24, 48), expand = expansion(c(0,0)), minor_breaks = seq(0, 96, 2) )  +
        ## style
        coord_cartesian(ylim = c(3.8, 60)) +
        scale_color_manual(values = palval_exbin ) +
        theme_bw() +
        geom_label_npc(label = "ELP1 KO", npcx = 0.98, npcy = 0.98) +
        theme(legend.position = "none") +
        labs(x = expression(m^6*A~"level [A.U.]"),
             y = "Half-life [h]"
        ) +
        
        NULL
      
    ]

pl.bar <-
  plotdat[, ggplot(.SD,
                     aes(x =  m6Alvl, y = log2(t12_from_decay(`ELP1 KO`) / t12_from_decay(`WT`)) )) +
          ## bar
          geom_point(aes(col = (exbin) , fill = (exbin))) +
              
              ## correlation
              stat_poly_line() +
              # stat_poly_eq(use_label("R2"), size = 4, coef.digits = 3, label.y = "top", label.x = "right", parse = TRUE) +
              
              ## scatter
              #geom_point(aes(col = (gloribin) , fill = (gloribin)), size = 0.75) +
              
              
              ## style
              #coord_cartesian(ylim = c(6, 60)) +
              scale_color_manual(values = palval_exbin ) +
              scale_fill_manual(values = palval_exbin ) +
              theme_bw() +
              theme(legend.position = "none") +
              labs(x = expression(m^6*A~"level [A.U.]"),
                   y = expression(Log[2]*FC)
              )]
  


ggarrange(pl.scatter +
            theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
            #theme(aspect.ratio=1) +
            NULL
          , 
          pl.bar
          , 
          ncol = 1, align = "v", heights = c(2,1)
          )

ggsave("~/exports/m6A_decay_ms_share/hallmark_t12_ELP1.pdf", width = 3, height = 4)

```

## STM mcm5s2U

### GLORI bin

```{r, fig.width = 6, fig.height = 3}

as.data.table(drach.tx)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts, "ensembl_transcript_id")
][
  , .(glori_m6A_lvl = sum(glori_m6A_lvl))
  , by = .(ensembl_gene_id, region)
][
  glori_m6A_lvl < 15
][
  , gloribin := cut2(glori_m6A_lvl, cuts = seq(0, 20, 1), levels.mean = TRUE)
  #, by = .(region)
][
  , merge(.SD, mcm5s2U_STM$log2FC_STM, "ensembl_gene_id", allow.cartesian = TRUE )
][
  ,
  # normalize to stable genes
  {
    norm_fact = .SD[gloribin %in% levels(gloribin)[1] , .(nf = median(log2FC, na.rm = TRUE)), by = .(region, time)]
    merge(.SD, norm_fact, by = c("region", "time"))[,  log2FC := log2FC - nf ][, !"nf"][]
  }
][
  time == "0h" & !region == "utr5"
][
  , region2 := factor(sedit(as.character(region), c("cds", "utr3"), c("CDS~m^6*A", "'3UTR'~m^6*A") ), levels = c("CDS~m^6*A", "'3UTR'~m^6*A"))
][
  , 
  
    ggplot(.SD, aes(x = as.numeric(as.character(gloribin)), y = log2FC, col = genotype )) +
    stat_summary(aes(col = genotype), fun.data = mean_se) +
    stat_poly_line() +
    #stat_poly_eq(use_label("eq", "R2"), size = 4, vstep = 0.075, coef.digits = 1) +
    geom_hline(yintercept = 0, linetype = 2) +
    facet_nested(~ region2, labeller = label_parsed ) +
    scale_color_manual(values = palval_genotype ) +
    theme_bw() +
    theme(legend.position =  "none") +
          labs(x = expression("mRNA"~ m^6*A~"level"),
           y = expression("log"[2]*"FC (+/-STM)")
      ) +
    NULL
  
]

ggsave("~/exports/m6A_decay_ms_share/gloribin_mcm5s2U_STM.pdf", width = 6, height = 3)

```

## Queuosine

```{r queuosine_GLORI, fig.width = 4, fig.height = 3}

as.data.table(cods.gr)[
  codon %in% c("GAC", "AAC")
][
  , ensembl_transcript_id := gsub("\\..*", "", seqnames)
][
  , gene := setkeyv(canonical_transcripts[, .(ensembl_transcript_id, gene)], "ensembl_transcript_id")[ensembl_transcript_id, gene]
][
  , melt(.SD, id.vars=c("codon", "gloribin", "glori_m6A_lvl"), measure.vars = queuosine_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = .(gloribin, codon, run)
][
  , c("experiment", "genotype", "replicate")  := setkeyv(queuosine_riboseq$meta, "run")[run, .(experiment, genotype, replicate) ]
][
  
][
  , ggplot(.SD, aes(x = glori_m6A_lvl, y = occupancy, col = genotype)) +
    stat_summary(fun.data = mean_se) +
    stat_poly_line(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ], se = FALSE ) +
    stat_poly_eq(data = function(x)x[!gloribin %in% levels(gloribin)[c(length(levels(gloribin)))]  ],
                 aes(label = after_stat(
                   ifelse(
                     r.squared < 0.1,
                     paste0(
                       "m~`=`~", round(b_1, 2)
                       #, "*`,`~italic(R)^2<0.01"
                     )
                     ,
                     paste0(
                       "m~`=`~", round(b_1, 2)
                       #, "*`,`~italic(R)^2~`=`~", round(r.squared, 2)
                     ))
                 )),
                 output.type = "numeric",
                 parse = TRUE,
                 vstep = 0.1
    ) +
    scale_color_manual(values = palval_queosine ) +
    facet_nested( ~ codon ) +
    
    theme_bw() +
    labs(x = expression("Codon"~ m^6*A~"stoichiometry"), y="A-site occupancy") +
    theme(legend.position =  "none") +
    #scale_x_continuous(limits = c(0,1)) +
    #geom_vline(aes(xintercept = 0.05), linetype = 2) +
    NULL
]

ggsave("~/exports/m6A_decay_ms_share/queuosine_GLORI.pdf", width = 4, height = 3)

```




