---
title: "R Notebook"
output: html_notebook
---

Integrated analysis of multiple miCLIP experiments.
Requires the "prerequisites" objects of the the "analysis_human.qmd" notebook.

# Data

## Pileup BAMs

```{r bamList}

bamList = list.files(paste0("../results/",c("B116", "DRB", "NatMeth"),"/star/homo_sapiens_GRCh38_v35_nospike"), ".Aligned.sortedByCoord.out.dedup.bam$", full.names = TRUE)
names(bamList) = gsub(".Aligned.sortedByCoord.out.dedup.bam", "", basename(bamList))

names(bamList)

```

```{r pileup parameters}
require("Rsamtools")
require("data.table")

## choose read start position bins that are available for downstream position-dependent filtering 
#query_bins = c(0,1,2,3,4, Inf)
query_bins = c(1:100, Inf)


puparam <-
  PileupParam(
    max_depth = 1000000,
    min_base_quality = 24,
    min_mapq = 60,
    min_nucleotide_depth = 2,
    min_minor_allele_depth = 1, # return only positions that have at least one mismatch
    distinguish_strands = TRUE,
    distinguish_nucleotides = TRUE,
    ignore_query_Ns = TRUE,
    include_deletions = TRUE,
    include_insertions = TRUE,
    left_bins = NULL,
    query_bins = query_bins, # bin by position in read to allow filtering later
    cycle_bins = NULL
  )

bamparam <-
  ScanBamParam(flag = scanBamFlag(isSecondaryAlignment = FALSE,
                                  isDuplicate = FALSE))

levels(cut(1:100, query_bins))

```

```{r pileup function}

get_pileups <-
  function(bamFile = NULL,
           bamparam = NULL,
           puparam = NULL,
           bsg = NULL,
           tmpDir = NULL, 
           recalc = FALSE) {
    puFile = file.path(tmpDir,
                       paste0(basename(bamFile), ".pileup", digest::digest(c(bamparam, puparam)), ".tsv"))
    
    if (!file.exists(puFile) | recalc) {
      pu <-
        as.data.table(pileup(
          file = bamFile,
          scanBamParam = bamparam,
          pileupParam = puparam
        ))
      
      ## correct nucleotide for strandedness
      # pu[strand == "-", nucleotide := chartr("AGCT", "TCGA", nucleotide)]
      pu[, nuc_corr := ifelse(strand == "-",
                              chartr("AGCT", "TCGA", as.character(nucleotide)),
                              as.character(nucleotide))]
      ## rename insertions and deletions
      pu[nucleotide == "+", nuc_corr := "ins"]
      pu[nucleotide == "-", nuc_corr := "del"]
      
      pu[, nuc_corr := factor(nuc_corr, levels =  c("A", "C", "G", "T", "del", "ins"))]
      
      ## get reference nucleotide
      pu[seqnames %in% seqlevels(bsg) , ref := {
        gr = GRanges(
          seqnames = seqnames,
          ranges = IRanges(start = pos, width = 1),
          strand =  strand
        )
        seqlevels(gr, pruning.mode = "coarse") = seqlevels(bsg)
        sq = BSgenome::getSeq(bsg, gr)
        sq = as.character(sq)
        return(sq)
      }]
      ## write pu
      fwrite(pu, puFile, sep = "\t")
    } else {
      pu <-
        fread(puFile)
    }
    
    ## make sure query bins is a factor, as this is needed for downstream filtering
    pu[, query_bin := factor(query_bin)]
    
    return(pu)
    
  }

```


```{r puList}

registerDoParallel(cores = 24 )

system.time(
puList <-
  foreach(bam = names(bamList)) %dopar% {
    get_pileups(bamFile = bamList[[bam]], bamparam = bamparam, puparam = puparam, bsg = bsg, tmpDir = tmpDir)[, bam := bam] 
  }
)

## sum counts of all samples
pu <-
  rbindlist(puList)[, .(count = sum(count)), by = setdiff(names(puList[[1]]), "count")]

pu[, bootstrapFrac := 100] ## this is the full data

head(pu)

```

# Analysis

## Bootstrap
```{r get_fraction_pileup function}

#' A function to generate random bootstraps of BAM files
#'
#' @description
#' This function performs a system call to samtools to generate randomly subsampled BAM files containing a fraction of the original reads.
#' It then returns a data table with pileups for the individual fractions.
#'
#' @param
#' bamFile path to a BAM file
#' bamFrac vector of fractions to generate
#' tmpDir path to directory where fractional BAM files are stored.
#' samtoolsPath full path to samtools binary.
get_fraction_pileup <-
  function(bamFile = NULL, bamFrac = NULL, tmpDir = tmpDir, samtoolsPath = "samtools" ) {
    if(!dir.exists(tmpDir))dir.create(tmpDir, showWarnings = FALSE, recursive = TRUE)
    bamseed = 42
    ## modified fractions for bootstrap in samtools compatible format (see samtools docs)
    fracm <-
      gsub("^0\\.", paste0(bamseed, "."), as.character(bamFrac / 100))
    ## bootstrapped BAM file
    fracBamFile <-
      file.path(tmpDir, gsub(
        "\\.bam$",
        paste0(".fracm", gsub("\\.","_",fracm), ".bam"),
        basename(bamFile)
      ))
    
    if (!file.exists(fracBamFile)) {
      system(paste(samtoolsPath, "view -b -s", fracm, bamFile,">", fracBamFile))
      }
    if (!file.exists(paste0(fracBamFile, ".bai") )) {
       system(paste(samtoolsPath, "index", fracBamFile))
    }
    ## generate pileup
    pu <-
      get_pileups(bamFile = fracBamFile, bamparam = bamparam, puparam = puparam, bsg = bsg, tmpDir = tmpDir)
    pu[, bootstrapFrac := bamFrac]
    return(pu)
    }

```

```{r bsList}

#bootstrapFractions = setdiff((1:20)*5, 100)
bootstrapFractions = setdiff((1:10)*10, 100)

registerDoParallel(24)

system.time(
bsList <-
foreach(bam = names(bamList)) %:% foreach(bsFrac = bootstrapFractions, .combine = rbind) %dopar% {
  get_fraction_pileup(bamFile = bamList[[bam]],
                      bamFrac = bsFrac,
                      tmpDir = file.path(tmpDir, "bootstraps"),
                      samtoolsPath = "~/mambaforge/envs/samtools/bin/samtools")[, bam := bam]
}
)
names(bsList) = names(bamList)

## sum counts of all samples
bs <-
  rbindlist(bsList)[, .(count = sum(count)), by = setdiff(names(bsList[[1]]), c("count", "bam") )]

## append summed counts of full data
bs <-
  rbind(bs, pu[, .(count = sum(count)), by = setdiff(names(pu), c("count", "bam") )])

head(bs)

```

```{r}
# 
# get_fraction_pileup(bamFile = bamList[[names(bamList)[[2]] ]],
#                     bamFrac = bootstrapFractions[1],
#                     tmpDir = file.path(tmpDir, "bootstraps"),
#                     samtoolsPath = "~/mambaforge/envs/samtools/bin/samtools")#[, bam := names(bamList)]

```


```{r}
head(bs[bootstrapFrac == 100, ])
```


```{r puw_from_pu function}
## a function to generate, pre-filter and annotate pileup
puw_from_pu <- function(pu){
  ## cast to wide to get info about each covered position
puw.intern <-
  dcast(
    pu, ## here we could filter for position in read with usebin
    bootstrapFrac + seqnames + pos + strand + ref ~ nuc_corr ,
    value.var = "count",
    drop = c(TRUE, FALSE),
    fill = 0,
    fun.aggregate = sum
    )
## calculate total coverage
puw.intern[, cvg := as.integer(rowSums(.SD)), .SDcols = c("A", "C", "G", "T", "-", "+")[c("A", "C", "G", "T", "-", "+") %in% colnames(puw.intern)]]
require(matrixStats)
require(GenomicFeatures)
## only with defined reference base
puw.intern <- puw.intern[ !is.na(ref) & !ref=="N", ]

## only in reference transcripts
puw.intern <- puw.intern[suppressWarnings(overlapsAny(GRanges(seqnames = seqnames, IRanges(start=pos, width =1), strand=strand), ex_by_tx)), ]

## only where reference is major call
puw.intern[ref == "A", is_major :=  `A` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]
puw.intern[ref == "G", is_major :=  `G` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]
puw.intern[ref == "C", is_major :=  `C` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]
puw.intern[ref == "T", is_major :=  `T` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]

puw.intern <- puw.intern[ which(is_major), ]

## only reference C
puw.intern <- puw.intern[ref == "C", ]

## get c2t  ratio
puw.intern[, c2t := `T` / `C`]


## shift to (putative) A 
puw.intern[, posA := ifelse(strand == "+", pos-1, pos+1)]

## annotate
puw.intern.gr <- 
  GRanges(seqnames = puw.intern$seqnames, IRanges(start=puw.intern$posA, width =1), strand=puw.intern$strand, mcols = DataFrame(puw.intern[, !c("seqnames", "posA", "pos", "strand")]))

puw.intern[, kmer := suppressWarnings(as.character(getSeq(bsg, puw.intern.gr +2)))] ## genomic kmer

puw.intern[, kmer_DRACH := grepl("[ATG][AG]AC[ACT]",kmer)]
puw.intern[, kmer_A := grepl("..A..",kmer)]
puw.intern[, tx_DRACH := suppressWarnings(overlapsAny(puw.intern.gr, drach.gr)) ]
puw.intern[, in_tx := overlapsAny(puw.intern.gr, ex_by_tx) ]
puw.intern[, tx_A := kmer_A & in_tx]

return(puw.intern)
}

```

```{r define posfilter}

posfilterN  = 0 ## exclude first N positions in reads

posfilter = levels(bs$query_bin)[(1+posfilterN):length(levels(bs$query_bin))]

posfilter

```

```{r bsw posfilter}

## we filter for position in read using query_bin
system.time(
bsw <-
  puw_from_pu(bs[query_bin %in% posfilter,])
)

head(bsw)

```

## Evaluate calling parameters
Evaluate different calling parameters to find a balance between the number of called sites and their "purity" (i.e. the fraction of called A or DRACH). This is done only on the complete data, not the bootstraps.
```{r evaluate calling parameters}
require(ggplot2)
require(doParallel)
registerDoParallel(cores = 24 )

## parameter space to evaluate
eval_c2tNmin = 1:9
eval_c2tmax = unique(c(seq(0, 0.4, 0.05), seq(0.4, 0.5, 0.01)))
eval_c2tmin = unique(c(seq(0, 0.1, 0.01), seq(0.1, 0.5, 0.05)))

eval_calls <-
  function(puw = NULL, c2tNmin = NULL, c2tmin = NULL, c2tmax = NULL) {
    hits = puw[ref == "C" & `T`>= c2tNmin & c2t > c2tmin & c2t <= c2tmax, ]
    dt = data.table(
      c2tNmin = c2tNmin,
      c2tmin = c2tmin,
      c2tmax = c2tmax,
      nsites = nrow(hits),
      nsites_A = sum(hits$tx_A),
      nsites_DRACH = sum(hits$tx_DRACH)
    )
    dt[, frac_A := nsites_A / nsites]
    dt[, frac_DRACH := nsites_DRACH / nsites]
    dt[, A_in_DRACH := nsites_DRACH / nsites_A]
    return(dt)
  }

paramtest <- 
foreach(c2tNmin = eval_c2tNmin, .combine=rbind) %:% foreach(c2tmin = eval_c2tmin, .combine=rbind) %:% foreach(c2tmax = eval_c2tmax, .combine=rbind) %dopar% {
  eval_calls(puw = bsw[bootstrapFrac == 100 & grepl("^chr", seqnames), ], c2tNmin = c2tNmin, c2tmin = c2tmin, c2tmax = c2tmax)
  }

paramtest

```

Plot parameter space to get an idea of valid parameters:
```{r fig.width = 13, fig.height = 9}

ggplot(paramtest, 
       aes(x = frac_A, y = nsites)) +
  geom_point(alpha = 0.1, aes(size = factor(c2tNmin))) +
  geom_hline(yintercept = 30000, linetype = 2) +
  geom_vline(xintercept = 0.85, linetype = 2) +
  scale_y_continuous(trans="log10", limits = c(NA, NA)) + annotation_logticks(10, "l", TRUE) +
  xlim(0.5, NA) +
  #facet_wrap(~c2tNmin) +
  NULL

ggplot(paramtest, 
       aes(x = A_in_DRACH, y = nsites)) +
  geom_point(alpha = 0.1, aes(size = factor(c2tNmin))) +
  geom_hline(yintercept = 30000, linetype = 2) +
  geom_vline(xintercept = 0.85, linetype = 2) +
  scale_y_continuous(trans="log10", limits = c(NA, NA)) + annotation_logticks(10, "l", TRUE) +
  xlim(0.5, NA) +
  #facet_wrap(~c2tNmin) +
  NULL

```
So we could get ~ 30K called As of which more than 80% are in a DRACH.  

Next, we define optimized parameters for calling sites:
```{r}
## define target minimum site "purity"
target_frac_A = 0.85
target_A_in_DRACH = 0.75 

## define target minimum of called sites
target_nsites = 25000
target_nsites_A = 25000

## rank filtered parameters by product of site quality and number
paramtest[, eval_crit := NULL]
paramtest[, callrank := NULL]
paramtest[, eval_crit := nsites_DRACH * A_in_DRACH]
paramtest[frac_A >= target_frac_A &
            A_in_DRACH >= target_A_in_DRACH &
            nsites > target_nsites,
          callrank := rank(-eval_crit, na.last = TRUE) ]

## define parameter cutoffs
call_c2tNmin = paramtest[!is.na(callrank), ][order(callrank), ][1, c2tNmin]
call_c2tmin  = paramtest[!is.na(callrank), ][order(callrank), ][1, c2tmin]
call_c2tmax  = paramtest[!is.na(callrank), ][order(callrank), ][1, c2tmax]

paramtest[!is.na(callrank), ][order(callrank), ]

```

## Call sites
Now we can use de parameters defined above to call sites in all bootstraps.
```{r call}

# bsw[, call := ifelse(
#   test = ref == "C" &
#     (`T` >= call_c2tNmin &
#        c2t >= call_c2tmin &
#        c2t < call_c2tmax),
#   yes = "methylated",
#   no = ifelse(
#     test = ref == "C" &
#       `T` <= 0 & cvg >= 90,
#     yes = "unmethylated",
#     no = "intermediate"
#   )
# )]

bsw[, call := NULL]
bsw[cvg > 90, call := "unmethylated"]
bsw[ref == "C" & `T` > 0, call := "intermediate"]
bsw[ref == "C" &
    (`T` >= call_c2tNmin &
       c2t >= call_c2tmin &
       c2t < call_c2tmax), call := "methylated"]

bsw[, call := factor(call)]

bsw[bootstrapFrac == 100, .(N = .N, fracDRACH = sum(tx_DRACH)/length(tx_DRACH)*100), by = "call" ]

```


Overview plot:

```{r plot_miclip_summary function}

plot_miclip_summary <-
  function(gr = NULL) {
    p.hits.table <-
      ggtexttable(tribble(~"", ~"", 
                          "sites", as.character(length(gr)),
                          "A", as.character(sum(gr$tx_A)),
                          "DRACH", as.character(sum(gr$tx_DRACH)),
                          "% A", as.character(round(sum(gr$tx_A)/length(gr)*100, 2)), 
                          "% DRACH", as.character(round(sum(gr$tx_DRACH)/length(gr)*100, 2)),
                          "% A in DRACH", as.character(round(sum(gr$tx_DRACH)/sum(gr$tx_A)*100, 2))),
                  rows = NULL, cols = NULL)
    
    p.hits.logo <-
      ggplot() + geom_logo(as.character(gr$kmer),
                           seq_type = "dna",
                           method = "probability") +
      ylim(0, 1) +
      #labs(title = "Called C>T transitions", subtitle = p.hits.subtit) +
      NULL
    
    
    
    p.hits.meta <-
      ggplot(data.frame(relpos.cds = relpos.cds.gr(suppressWarnings(
        mapToTranscripts(gr, ex_by_tx)
      ))), aes(x = relpos.cds)) +
      geom_histogram(binwidth = 0.02) +
      geom_vline(
        xintercept = c(1, 2),
        linetype = 2,
        col = "blue",
        lwd = 0.3
      ) +
      xlim(0.5, 3)
    
    p <-
      ggarrange(
        p.hits.logo,
        p.hits.table,
        p.hits.meta,
        ncol = 1,
        nrow = 3,
        align = "v"
      )
    return(p)
  }

```


```{r ## miclip_summary ,fig.width=10, fig.height=8}
require(tibble)

hits <- 
  bsw[bootstrapFrac == 100 & call == "methylated"  ,]

hits.gr <- 
  GRanges(seqnames = hits$seqnames,
          IRanges(start=hits$posA, width =1),
          strand=hits$strand)
seqlevels(hits.gr) = seqlevels(ex_by_tx)

mcols(hits.gr) <-
  DataFrame(hits[, c("cvg", "c2t", "A", "C", "G", "T", "ins", "del", "kmer", "tx_DRACH", "tx_A")])


p.hits.paramspace <-
  ggplot(paramtest, 
       aes(x=A_in_DRACH, y= nsites_DRACH)) +
  geom_point(alpha = 0.1) +
  geom_point(data = function(x)x[callrank == 1, ], col = "red", size = 2) +
  geom_hline(yintercept = target_nsites_A, linetype =2) +
  #geom_vline(xintercept = target_frac_DRACH, linetype = 2) +
  scale_y_continuous(trans="log10", limits = c(NA, NA)) + annotation_logticks(10, "l", TRUE) +
  xlim(0.5, NA) +
  facet_wrap(~c2tNmin, labeller = label_both) +
  NULL

annotate_figure(ggarrange(p.hits.paramspace, plot_miclip_summary(hits.gr),  ncol = 2, widths = c(3,1) ),
                top = paste0("posfilterN=", posfilterN,
                              " | c2tNmin=",
                              call_c2tNmin,
                              " | c2tmin=",
                              call_c2tmin,
                              " | c2tmax=",
                              call_c2tmax
                              )
                )


# save_last_plot(file = file.path(plotDir, "region_histogram.pdf"), owncube_dir = "exports/plots", width = 25, height = 8, scale = 2)

```
```{r ## miclip_bootstrap,  fig.width = 9, fig.height = 4}

bckgrnd_DRACH = 3/4*2/4*1*1/4*3/4 * 100

annotate_figure(
  ggarrange(
    
    ggplot(bsw[, .(fracDRACH = sum(tx_DRACH) / length(tx_DRACH) * 100), by = c("bootstrapFrac", "call")],
           aes(
             x = bootstrapFrac, y = fracDRACH, col = call
           )) +
      geom_hline(yintercept = bckgrnd_DRACH, linetype = 2) +
      geom_text(x = -Inf, y = bckgrnd_DRACH, label = paste0("Background DRACH (",round(bckgrnd_DRACH, 1),"%)"), check_overlap = TRUE, col = "black", size = 3, hjust = 0, vjust = -0.2) +
      geom_point() +
      geom_smooth(aes(fill = call)) +
      ylim(0, 100) +
      xlim(0, 100) +
      #coord_fixed() +
      labs(x = "Bootstrap [%]", y = "Fraction in DRACH [%]") +
      theme_pubr(legend = "right")
    ,
    ggplot(bsw[call == "methylated", .N, by = c("bootstrapFrac", "call")],
           aes(
             x = bootstrapFrac, y = N, col = call
           )) +
      geom_point() +
      geom_smooth(method = "lm") +
      xlim(0, 100) +
      labs(x = "Bootstrap [%]", y = expr("Called" ~ m ^ 6 * A ~ "sites [N]")) +
      theme_pubr(legend = "right")
    ,
    ncol = 2,
    common.legend = TRUE,
    #align = "v",
    legend = "right"
  ),
  top = paste0(
    "c2tNmin=",
    call_c2tNmin,
    " | c2tmin=",
    call_c2tmin,
    " | c2tmax=",
    call_c2tmax
  )
)

```

## BED file export
We export all covered DRACH sites into a BED file and include the methylation status call in the site name.
```{r}

dir.create("../../results/analysis_miCLIP")

fwrite(
  bsw[bootstrapFrac == 100 & tx_DRACH == TRUE, ## all C2T transitions in DRACH (only from full dataset)
       .(
         chr    = seqnames,
         start  = format(posA -1 , scientific = FALSE), ## bed is 0-based
         end    = format(posA    , scientific = FALSE),
         name   = paste0(call,
                         "_cvg", cvg,
                         "_c", `C`,
                         "_t", `T`),
         score  = format(c2t *1000, digits = 3, scientific = FALSE), ## bed file format specifies score between 0 and 1000
         strand = strand)],
  file = "../../results/analysis_miCLIP/miclip_c2t.bed",
  sep = "\t",
  row.names = F,
  col.names = F
)

fread("../../results/analysis_miCLIP/miclip_c2t.bed")

```

### --- Experimental code ---

```{r ### stop ###}
stop()
```

```{r}

bsw[bootstrapFrac == 100 & tx_DRACH == TRUE,]

bed.gr <-
  import("../../results/analysis_miCLIP/miclip_c2t.bed")

bed.tx <-
  mapToTranscripts(bed.gr, ex_by_tx)

bed.gr

```


```{r}

test <-
relpos.cds.gr(suppressWarnings(
        mapToTranscripts(hits.gr, ex_by_tx)
      ))

table(is.na(test))

```


```{r}
paramtest[order(-eval_crit, na.last = TRUE), ]

paramtest[nsites_A > 30000, ][order(-A_in_DRACH), ]
paramtest[nsites_A > 30000, ][order(-frac_A), ]

paramtest[A_in_DRACH > 0.75, ][order(-nsites_A), ]

paramtest[frac_A > 0.8, ][order(-nsites_A), ]

```


```{r fig.width = 16, fig.height = 8}

plotdat <-
  pu[grepl("^chr", seqnames), ]

plotdat[, mut := paste0(ref, ">", nuc_corr)]
plotdat[grepl("ins", mut), mut := "ins"]
plotdat[grepl("del", mut), mut := "del"]
plotdat[, mut := factor(mut, levels = c("ins", "del", grep("ins|del", sort(unique(mut)), invert = T, value = T)))]

ggplot(plotdat[nuc_corr != ref & ref != "N", .(count = sum(count)), by = c("strand", "mut", "bam") ],
       aes(x = mut, y = count, fill = strand)) +
  geom_col(position = "dodge") +
  facet_wrap(~bam, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  NULL

```

```{r fig.width = 16, fig.height = 8}

plotdat <-
  pu[grepl("^chr", seqnames), ]

plotdat[, mut := paste0(ref, ">", nuc_corr)]
plotdat[grepl("ins", mut), mut := "ins"]
plotdat[grepl("del", mut), mut := "del"]
plotdat[, mut := factor(mut, levels = c("ins", "del", grep("ins|del", sort(unique(mut)), invert = T, value = T)))]

plotdat[, readpos := as.integer(query_bin)]

plotdat <-
  plotdat[ref != nuc_corr & readpos != max(readpos) & ref != "N", .(count = sum(count)), by = c("readpos", "mut", "bam") ]

ggplot(plotdat,
       aes(x=readpos,  y= count)) +
  geom_line(aes(col = mut)) +
  facet_wrap(~bam, scales = "free_y") +
  theme(legend.position = "right") +
  NULL

```

```{r}

plotdat <-
  pu[as.integer(query_bin) > 1 & nuc_corr != ref, .(count = sum(count)), by = setdiff(names(pu), c("count", "query_bin"))]

```


```{r, fig.width = 20, fig.height = 10}

bs[, mut := paste0(ref, ">", nuc_corr)]
bs[grepl("ins", mut), mut := "ins"]
bs[grepl("del", mut), mut := "del"]
bs[, mut := factor(mut, levels = c("ins", "del", grep("ins|del", sort(unique(mut)), invert = T, value = T)))]

ggplot(bs[! mut %in% c("G>G", "A>A", "C>C", "T>T") & !is.na(ref) & ref !="" & !ref=="N",.N, by=c("mut", "bootstrapFrac", "query_bin")],
       aes(x=mut, y= N)) + 
  geom_bar(stat="identity", aes(fill=mut), show.legend = F) +
  facet_wrap(query_bin~bootstrapFrac, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  NULL

```





```{r}


get_relpos <-
  function(dt, ex_by_tx){
    gr <-
    GRanges(seqnames = dt$seqnames,
          IRanges(start=dt$posA, width =1),
          strand=dt$strand)
    seqlevels(gr) = seqlevels(ex_by_tx)
    tx <-
       mapToTranscripts(gr, ex_by_tx)
    res <-
      data.table(relpos.cds = relpos.cds.gr(suppressWarnings(
        mapToTranscripts(gr, ex_by_tx)
      )))
    return(res)
  }

test <-
foreach(cl = levels(bsw$call), .combine = rbind) %dopar% {
  get_relpos(bsw[bootstrapFrac == 100 & call == cl  ,], ex_by_tx)[, call := cl]
}

ggplot(test, 
       aes(x=relpos.cds, col = call))+
  #geom_freqpoly(binwidth = 0.02) +
  geom_density() +
      geom_vline(
        xintercept = c(1, 2),
        linetype = 2,
        col = "blue",
        lwd = 0.3
      ) +
      xlim(0.5, 3) +
  NULL

```


## Benchmark

```{r}


# liftover to GRCh38
#chainFileURL = "http://hgdownload.cse.ucsc.edu/goldenpath/hg19/liftOver/hg19ToHg38.over.chain.gz"
chainFileURL = "http://hgdownload.cse.ucsc.edu/goldenpath/hg38/liftOver/hg38ToHg19.over.chain.gz"
chainFileGZ = file.path(tmpDir, basename(chainFileURL))
chainFile = gsub("*\\.gz$", "", chainFileGZ)
if(!file.exists(chainFile)){
  if(!file.exists(chainFileGZ))download.file(chainFileURL, chainFileGZ)
  GEOquery::gunzip(chainFileGZ, chainFile)
}
c2t <-
  unlist(rtracklayer::liftOver(c2t, import.chain(chainFile)))

```


### MasterList
```{r}

dataURL = "https://linder.owncube.com/index.php/s/ML43zTddqi7wXoE/download"
dataFile = file.path(tmpDir, "master.cims.unique.bed")
if(!file.exists(dataFile))download.file(dataURL, dataFile)

bed.gr <- 
  rtracklayer::import(dataFile)  

bed.gr$kmer <-
  suppressWarnings(as.character(getSeq(bsg, bed.gr +2)))

## determine DRACH by genomic kmer
bed.gr$kmer_DRACH = grepl("[ATG][AG]AC[ACT]",bed.gr$kmer)
bed.gr$kmer_A = grepl("..A..",bed.gr$kmer)

## the A can be safely determined from the genomic kmer. However, we need to filter for only those that are in ex_by_tx, otherwise ratio to DRACH will be skewed by intergenic sites
bed.gr$in_tx = overlapsAny(bed.gr, ex_by_tx)
bed.gr$tx_A  = bed.gr$kmer_A & bed.gr$in_tx

bed.gr$tx_DRACH = suppressWarnings(overlapsAny(bed.gr, drach.gr))

ggarrange(
  annotate_figure(plot_miclip_summary(hits.gr), top = paste0("Nc2tmin=", call_c2tNmin, " | c2tmin=", call_c2tmin, " | c2tmax=", call_c2tmax)),
  annotate_figure(plot_miclip_summary(bed.gr ), top = "miCLIP master list"), ncol = 2
)

```

### Nature Methods
Comapare to original miCLIP dataset:
```{r}

dataURL = "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE63nnn/GSE63753/suppl/GSE63753_hek293.abcam.CIMS.m6A.9536.bed.txt.gz"
dataFile = file.path(tmpDir, basename(dataURL))
if(!file.exists(dataFile))download.file(dataURL, dataFile)

bed.gr <-
makeGRangesFromDataFrame(
setnames(fread(dataFile), paste0("V", 1:6), c('chr', 'start', 'end', 'id', 'score', 'strand')),
keep.extra.columns =T,
starts.in.df.are.0based = T)

seqlevels(bed.gr) = seqlevels(ex_by_tx)

bed.gr$kmer <-
  suppressWarnings(as.character(getSeq(bsg, bed.gr +2)))

## determine DRACH by genomic kmer
bed.gr$kmer_DRACH = grepl("[ATG][AG]AC[ACT]",bed.gr$kmer)
bed.gr$kmer_A = grepl("..A..",bed.gr$kmer)

## the A can be safely determined from the genomic kmer. However, we need to filter for only those that are in ex_by_tx, otherwise ratio to DRACH will be skewed by intergenic sites
bed.gr$in_tx = overlapsAny(bed.gr, ex_by_tx)
bed.gr$tx_A  = bed.gr$kmer_A & bed.gr$in_tx
bed.gr$tx_DRACH = suppressWarnings(overlapsAny(bed.gr, drach.gr))


ggarrange(
  annotate_figure(plot_miclip_summary(hits.gr), top = paste0("Nc2tmin=", call_c2tNmin, " | c2tmin=", call_c2tmin, " | c2tmax=", call_c2tmax)),
  annotate_figure(plot_miclip_summary(bed.gr ), top = "Linder et al. 2018"), ncol = 2
)

```



## Calculations of DRACH abundance
The distribution of Cs in DRACHes in the pileups is way higher than the theoretical mean (~7%)
### theoretical
```{r}

## theoretical DRACH fraction for all Cs (all 4 nuc equally abundant)
DRACH_frac_theo = (3/4)*(2/4)*(1/4)*(1)*(3/4) * 100 
print(DRACH_frac_theo)

## DRACH fraction for all Cs in txmodel
DRACH_frac_txmodel = length(drach.tx) / length(unlist(vmatchPattern("C", tx.seqs, fixed = TRUE))) * 100
print(DRACH_frac_txmodel)

## DRACH fraction for Cs with IPed reads coverage but without transition (C == cvg)
DRACH_frac_notrans = tdt(bsw[bootstrapFrac == 100 &  ref=="C" & C == cvg,  .N , by = .(tx_DRACH)])[, `TRUE`/(`TRUE`+`FALSE`)*100 ]
print(DRACH_frac_notrans)

## DRACH fraction for Cs with exactly one transition (T == 1)
DRACH_frac_onetrans = tdt(bsw[bootstrapFrac == 100 &  ref=="C" & `T` == 1,  .N , by = .(tx_DRACH)])[, `TRUE`/(`TRUE`+`FALSE`)*100 ]
print(DRACH_frac_onetrans)

## DRACH fraction for Cs with at least one transition (T >= 1)
DRACH_frac_alltrans = tdt(bsw[bootstrapFrac == 100 &  ref=="C" & `T` >= 1,  .N , by = .(tx_DRACH)])[, `TRUE`/(`TRUE`+`FALSE`)*100 ]
print(DRACH_frac_alltrans)

## DRACH fraction in called hits
DRACH_frac_called = sum(hits.gr$tx_DRACH) / length(hits.gr$tx_DRACH)  *100
print(DRACH_frac_called)

```




## plot rariant
```{r, fig.width=12, fig.height = 4}
require(Rariant)
require(ggbio)
require(Biostrings)
require(ggbio)


gr.actb <- GRanges(seqnames = "chr7", IRanges(5567374, width = 1), strand = "-") ## actb site
gr.eef2 <- GRanges(seqnames = "chr19", IRanges(3976466, width = 1), strand = "-") ## eef2 site
#gr.eef2 <- GRanges(seqnames = "chr19", IRanges(3976161, width = 1), strand = "-") ## eef2 site

gr.atf4 <- GRanges(seqnames = "chr22", IRanges(39916568, width = 1), strand = "+") ## atf4 m6Am
gr.hspa1b <- GRanges(seqnames = "chr6", IRanges(31795649, width = 1), strand = "+") ## hsp70 utr5

gr = gr.actb

bsw[bootstrapFrac == 100  & seqnames == as.character(seqnames(gr)) & strand == as.character(strand(gr))  & pos %in% (start(gr+2):end(gr+2)) ,  ][]

winsize = 25 

## we need to generate an new GRanges and drop the strand, as tallyPlot is not strand aware
gr2 = GRanges(seqnames(gr), IRanges(start(gr), end(gr))) + winsize

p.bf  <- 
  tallyPlot(file = bamFile , region = gr2, ref = Hsapiens, nCycles = 5) +
    geom_vline(xintercept = start(gr), linetype=2)
## reference sequence
p.ref <- autoplot(Hsapiens, gr2) + theme_clear() + theme(legend.position="none")
## todo: add DRACHes p.drach.pos <- autoplot(drach.pos)

tracks(bf = p.bf,  ref = p.ref,  heights = c(5, 0.5) )

```

```{r}
stop()
```


look for recalls
```{r}



get_recall <-
  function(dt, bs){
    dt <-
      dt[call == "methylated", ]
    dt[, id := paste(seqnames, pos, strand, sep="_")]
    current <- 
      dt[bootstrapFrac == bs, ]
    previous <-
      dt[bootstrapFrac <  bs, ]
    current[, is_recall := id %in% unique(previous$id)]
    return(current)
  }



recall <-
foreach(bs = unique(bsw$bootstrapFrac), .combine = rbind ) %dopar% {
  get_recall(dt = bsw, bs = bs)
 }

recall_agg <-
  dcast(recall[, .N, by = c("bootstrapFrac", "is_recall")],bootstrapFrac ~ ifelse(is_recall, "recall", "firstcall"), value.var = "N", fill = NA)

recall_agg[, totalcall := firstcall + recall]

ggplot(recall_agg, 
       aes(x=bootstrapFrac, y = totalcall)) +
  geom_point() +
  geom_smooth()


recall_agg$recall_binsize = foreach(i = 1:nrow(recall_agg), .combine = c) %do% {recall_agg[i,  recall_agg[i+1, bootstrapFrac] - recall_agg[i, bootstrapFrac] ] }


recall_agg$recall_binsize = (c(recall_agg$bootstrapFrac, 100) - c(0, recall_agg$bootstrapFrac))[1:nrow(recall_agg)]

ggplot(recall_agg, 
       aes(x=bootstrapFrac, y = firstcall)) +
  geom_point() +
  geom_smooth()

ggplot(recall_agg, 
       aes(x=totalcall, y = firstcall)) +
  geom_point() +
  geom_smooth() +
  geom_smooth(data = function(x)x[bootstrapFrac > 30, ], method = "lm", fullrange = TRUE) +
  ylim(0, NA) +
  xlim(0, 5e4) +
  NULL

```




