---
title: "R Notebook"
output: html_notebook
---

Integrated analysis of multiple miCLIP experiments.

# Prerequisites

Requires the objects from the "Prerequisites" and "Transcript Model" sections of the the "analysis_human.qmd" notebook.

# Data

## Pileup BAMs

```{r bamList}

bamList = list.files(paste0("../results/",c("miCLIP", "DRB", "NatMeth", "Mauer_et_al"),"/star/homo_sapiens_GRCh38_v35"), ".Aligned.sortedByCoord.out.dedup.bam$", full.names = TRUE)
names(bamList) = gsub(".Aligned.sortedByCoord.out.dedup.bam", "", basename(bamList))

names(bamList)

```

```{r pileup parameters}
require("Rsamtools")
require("data.table")

## choose read start position bins that are available for downstream position-dependent filtering 
#query_bins = c(0,1,2,3,4, Inf)
query_bins = c(1:100, Inf)


puparam <-
  PileupParam(
    # max_depth = 1000000,
    min_base_quality = 24,
    min_mapq = 60,
    min_nucleotide_depth = 2,
    min_minor_allele_depth = 1, # return only positions that have at least one mismatch
    distinguish_strands = TRUE,
    distinguish_nucleotides = TRUE,
    ignore_query_Ns = TRUE,
    include_deletions = TRUE,
    include_insertions = TRUE,
    left_bins = NULL,
    query_bins = query_bins, # bin by position in read to allow filtering later
    cycle_bins = NULL
  )

bamparam <-
  ScanBamParam(flag = scanBamFlag(isSecondaryAlignment = FALSE,
                                  isDuplicate = FALSE))

levels(cut(1:100, query_bins))

```

```{r pileup function}

get_pileups <-
  function(bamFile = NULL,
           bamparam = NULL,
           puparam = NULL,
           bsg = NULL,
           tmpDir = NULL, 
           recalc = FALSE) {
    puFile = file.path(tmpDir,
                       paste0(basename(bamFile), ".pileup", digest::digest(c(bamparam, puparam)), ".tsv"))
    
    if (!file.exists(puFile) | recalc) {
      cat(paste("Calculating pileup file..."))
      pu <-
        as.data.table(pileup(
          file = bamFile,
          scanBamParam = bamparam,
          pileupParam = puparam
        ))
      
      ## correct nucleotide for strandedness
      # pu[strand == "-", nucleotide := chartr("AGCT", "TCGA", nucleotide)]
      pu[, nuc_corr := ifelse(strand == "-",
                              chartr("AGCT", "TCGA", as.character(nucleotide)),
                              as.character(nucleotide))]
      ## rename insertions and deletions
      pu[nucleotide == "+", nuc_corr := "ins"]
      pu[nucleotide == "-", nuc_corr := "del"]
      
      pu[, nuc_corr := factor(nuc_corr, levels =  c("A", "C", "G", "T", "del", "ins"))]
      
      ## get reference nucleotide
      pu[seqnames %in% seqlevels(bsg) , ref := {
        gr = GRanges(
          seqnames = seqnames,
          ranges = IRanges(start = pos, width = 1),
          strand =  strand
        )
        seqlevels(gr, pruning.mode = "coarse") = seqlevels(bsg)
        sq = BSgenome::getSeq(bsg, gr)
        sq = as.character(sq)
        return(sq)
      }]
      ## write pu
      fwrite(pu, puFile, sep = "\t")
      cat(paste("...done!\n"))
    } else {
      cat(paste("reading existing pileup file.\n"))
      pu <-
        fread(puFile)
    }
    
    ## make sure query bins is a factor, as this is needed for downstream filtering
    pu[, query_bin := factor(query_bin)]
    pu[, query_bin := factor(query_bin, levels = gtools::mixedsort(levels(query_bin)))]
    return(pu)
    
  }

```


```{r puList}

registerDoParallel(cores = 10 )

system.time(
puList <-
  foreach(bam = names(bamList)) %dopar% {
    get_pileups(bamFile = bamList[[bam]], bamparam = bamparam, puparam = puparam, bsg = bsg, tmpDir = tmpDir)[, bam := bam] 
  }
)

## sum counts of all samples
pu <-
  rbindlist(puList)[, .(count = sum(count)), by = setdiff(names(puList[[1]]), "count")]

pu[, query_bin := factor(query_bin, levels = gtools::mixedsort(levels(query_bin)))]

head(pu)

```

## Basic QC
```{r}

pu[
  , .N, by = bam
]

pu[
  nuc_corr != ref
][
  ,  ts := paste0(ref, ">", nuc_corr)
][
  !is.na(ts)
  , .(N = .N)
  , by = .(query_bin, ts, bam) 
][
  , frac := N / sum(N)
  , by = .(ts, bam)
][
  ts == "C>T"
][
  , ggplot(.SD, aes(x = query_bin, y = frac )) +
    geom_point(aes(col = bam)) +
    facet_wrap(~ts) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 4))
]

```

```{r, fig.width = 24, fig.height = 24}

pu[
  nuc_corr != ref
][
  bam == "miCLIP"
][
  ,  ts := paste0(ref, ">", nuc_corr)
][
  !is.na(ts)
  , .(N = .N, frac = .N / sum(.N))
  , by = .(query_bin, ts, bam) 
][
  , ggplot(.SD, aes(x = query_bin, y = N)) +
    geom_point(aes(col = bam)) +
    facet_wrap(~ts) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 4))
]

```


# Analysis

```{r puw_from_pu function}
## a function to generate, pre-filter and annotate pileup
puw_from_pu <- function(pu){
  ## cast to wide to get info about each covered position
puw.intern <-
  dcast(
    pu, ## here we could filter for position in read with usebin
    seqnames + pos + strand + ref ~ nuc_corr ,
    value.var = "count",
    drop = c(TRUE, FALSE),
    fill = 0,
    fun.aggregate = sum
    )
## calculate total coverage
puw.intern[, cvg := as.integer(rowSums(.SD)), .SDcols = c("A", "C", "G", "T", "-", "+")[c("A", "C", "G", "T", "-", "+") %in% colnames(puw.intern)]]
require(matrixStats)
require(GenomicFeatures)
## only with defined reference base
puw.intern <- puw.intern[ !is.na(ref) & !ref=="N", ]

## only in reference transcripts
puw.intern <- puw.intern[suppressWarnings(overlapsAny(GRanges(seqnames = seqnames, IRanges(start=pos, width =1), strand=strand), ex_by_tx)), ]

## only where reference is major call
puw.intern[ref == "A", is_major :=  `A` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]
puw.intern[ref == "G", is_major :=  `G` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]
puw.intern[ref == "C", is_major :=  `C` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]
puw.intern[ref == "T", is_major :=  `T` >= rowMaxs(as.matrix(.SD))  , .SDcols=c("A","C", "G", "T") ]

puw.intern <- puw.intern[ which(is_major), ]

## only reference C
puw.intern <- puw.intern[ref == "C", ]

## get c2t  ratio
puw.intern[, c2t := `T` / `C`]


## shift to (putative) A 
puw.intern[, posA := ifelse(strand == "+", pos-1, pos+1)]

## annotate
puw.intern.gr <- 
  GRanges(seqnames = puw.intern$seqnames, IRanges(start=puw.intern$posA, width =1), strand=puw.intern$strand, mcols = DataFrame(puw.intern[, !c("seqnames", "posA", "pos", "strand")]))

puw.intern[, kmer := suppressWarnings(as.character(getSeq(bsg, puw.intern.gr +2)))] ## genomic kmer

puw.intern[, kmer_DRACH := grepl("[ATG][AG]AC[ACT]",kmer)]
puw.intern[, kmer_A := grepl("..A..",kmer)]
puw.intern[, tx_DRACH := suppressWarnings(overlapsAny(puw.intern.gr, drach.gr)) ]
puw.intern[, in_tx := overlapsAny(puw.intern.gr, ex_by_tx) ]
puw.intern[, tx_A := kmer_A & in_tx]

return(puw.intern)
}

```

```{r define posfilter}

posfilterN  = 0 ## exclude first N positions in reads

posfilter = levels(pu$query_bin)[(1+posfilterN):length(levels(pu$query_bin))]

posfilter

```

```{r drach.gr}
## DRACH sites in genome space
drach.gr <- unique(mapFromTranscripts(drach.tx, ex_by_tx))

drach.gr

```

## Evaluate calling parameters

```{r puw}
puw <-
  puw_from_pu(pu)
```

## Call sites

```{r manual call parameters}

## define parameter cutoffs
call_c2tNmin = 3
call_c2tmin  = 0.02
call_c2tmax  = 0.49

```


```{r call}

puw[, call := NULL]
puw[cvg > 90 & ref == "C" & `T` == 0, call := "unmethylated"]
puw[ref == "C" & `T` > 0, call := "intermediate"]
puw[ref == "C" &
    (`T` >= call_c2tNmin &
       c2t >= call_c2tmin &
       c2t < call_c2tmax), call := "methylated"]

puw[, call := factor(call)]

puw[, .(N = .N, fracDRACH = sum(tx_DRACH)/length(tx_DRACH)*100), by = "call" ]

```

Overview plot:
```{r plot_miclip_summary function}

plot_miclip_summary <-
  function(gr = NULL) {
    p.hits.table <-
      ggtexttable(tribble(~"", ~"", 
                          "sites", as.character(length(gr)),
                          "A", as.character(sum(gr$tx_A)),
                          "DRACH", as.character(sum(gr$tx_DRACH)),
                          "% A", as.character(round(sum(gr$tx_A)/length(gr)*100, 2)), 
                          "% DRACH", as.character(round(sum(gr$tx_DRACH)/length(gr)*100, 2)),
                          "% A in DRACH", as.character(round(sum(gr$tx_DRACH)/sum(gr$tx_A)*100, 2))),
                  rows = NULL, cols = NULL)
    
    p.hits.logo <-
      ggplot() + geom_logo(as.character(gr$kmer),
                           seq_type = "dna",
                           method = "probability") +
      ylim(0, 1) +
      #labs(title = "Called C>T transitions", subtitle = p.hits.subtit) +
      NULL
    
    
    
    p.hits.meta <-
      ggplot(data.frame(relpos.cds = relpos.cds.gr(suppressWarnings(
        mapToTranscripts(gr, ex_by_tx)
      ))), aes(x = relpos.cds)) +
      geom_histogram(binwidth = 0.02) +
      geom_vline(
        xintercept = c(1, 2),
        linetype = 2,
        col = "blue",
        lwd = 0.3
      ) +
      xlim(0.5, 3)
    
    p <-
      ggarrange(
        p.hits.logo,
        p.hits.table,
        p.hits.meta,
        ncol = 1,
        nrow = 3,
        align = "v"
      )
    return(p)
  }

```


```{r ## miclip_summary ,fig.width=10, fig.height=8}
require(tibble)
require(ggseqlogo)

hits <- 
  puw[ call == "methylated"  ,]

hits.gr <- 
  GRanges(seqnames = hits$seqnames,
          IRanges(start=hits$posA, width =1),
          strand=hits$strand)
seqlevels(hits.gr) = seqlevels(ex_by_tx)

mcols(hits.gr) <-
  DataFrame(hits[, c("cvg", "c2t", "A", "C", "G", "T", "ins", "del", "kmer", "tx_DRACH", "tx_A")])


# p.hits.paramspace <-
#   ggplot(paramtest,
#        aes(x=A_in_DRACH, y= nsites_DRACH)) +
#   geom_point(alpha = 0.1) +
#   geom_point(data = function(x)x[callrank == 1, ], col = "red", size = 2) +
#   geom_hline(yintercept = target_nsites_A, linetype =2) +
#   #geom_vline(xintercept = target_frac_DRACH, linetype = 2) +
#   scale_y_continuous(trans="log10", limits = c(NA, NA)) + annotation_logticks(10, "l", TRUE) +
#   xlim(0.5, NA) +
#   facet_wrap(~c2tNmin, labeller = label_both) +
#   NULL

annotate_figure(ggarrange(NA, plot_miclip_summary(hits.gr),  ncol = 2, widths = c(3,1) ),
                top = paste0("posfilterN=", posfilterN,
                              " | c2tNmin=",
                              call_c2tNmin,
                              " | c2tmin=",
                              call_c2tmin,
                              " | c2tmax=",
                              call_c2tmax
                              )
                )


# save_last_plot(file = file.path(plotDir, "region_histogram.pdf"), owncube_dir = "exports/plots", width = 25, height = 8, scale = 2)

```

## BED file export
We export all covered DRACH sites into a BED file and include the methylation status call in the site name.
```{r}

dir.create("../results/analysis_miCLIP")

fwrite(
  puw[tx_DRACH == TRUE, ## all C2T transitions in DRACH (only from full dataset)
       .(
         chr    = seqnames,
         start  = format(posA -1 , scientific = FALSE), ## bed is 0-based
         end    = format(posA    , scientific = FALSE),
         name   = paste0(call,
                         "_cvg", cvg,
                         "_c", `C`,
                         "_t", `T`),
         score  = format(c2t *1000, digits = 3, scientific = FALSE), ## bed file format specifies score between 0 and 1000
         strand = strand)],
  file = "../results/analysis_miCLIP/miclip_c2t.bed",
  sep = "\t",
  row.names = F,
  col.names = F
)

fread("../results/analysis_miCLIP/miclip_c2t.bed")

```

