---
title: "R Notebook"
output: html_notebook
---

# Setup

```{r codon_type}

palval_codon_type = c("other" = "grey40", "mcm5" = palval_genotype[["ELP1 KO"]], "mcm5s2U" = palval_genotype[["CTU2 KO"]])


get_codon_type <-
  function(codon){
    mcm5_codons = c("ACA", "CCA", "GCA", "GGA", "GTA", "TCA", "TTA")
    mcm5s2U_codons = c("CAA", "AAA", "AGA", "GAA")
    codon_type = ifelse(codon %in% mcm5_codons, "mcm5", ifelse(codon %in% mcm5s2U_codons, "mcm5s2U", "other"))
    codon_type = factor(codon_type, levels = names(palval_codon_type))
    return(codon_type)
  }


palval_codon_type

```

# Data
```{r functions}

## function for reading the coverage files generated by Jie's pipeline
read_tx_coverage <-
  function(coverage_file, file_suffix){
    data <-
      fread(coverage_file, sep = "\t", sep2 = ",", header = FALSE)[
        , .(codon_rank = as.integer(1:(V2+1)), coverage = as.numeric(str_split_1(V3, ",")) )
        , by = .(ensembl_transcript_id = str_split_i(V1, "\\|", 2), cds_len_cods = V2 )]
    setnames(data, "coverage", gsub(file_suffix, "", basename(coverage_file)))
    setnames(data, gsub("^PSLib[0-9]+_RPF_", "", names(data)))
  }

## wrapper to get coverage objects
get_tx_coverage <-
  function(experiment_name, file_path, file_suffix){
    
    coverage_files = list.files(path = file_path, pattern = file_suffix, full.names = TRUE)
    if(length(coverage_files) == 0){stop("No files found!")}else(cat(paste("Reading", length(coverage_files), "files.")))
    
    ## read in data
    coverage <-
      foreach(coverage_file = coverage_files,
              .combine = function(x,y){merge(x,y, c("ensembl_transcript_id", "cds_len_cods", "codon_rank"), all = TRUE)}) %do% {
                read_tx_coverage(coverage_file, file_suffix)
              }
    
    meta <-
      data.table(experiment = experiment_name, run = setdiff(names(coverage), c("ensembl_transcript_id", "cds_len_cods", "codon_rank")) )
      
    
    return(list(file_suffix = file_suffix, coverage = coverage, meta = meta))
    
  }
 
```

```{r mcm5s2U_riboseq}

# mcm5s2U_riboseq experiment 1
## read in data
mcm5s2U_riboseq_1 <-
  get_tx_coverage(experiment_name = "exp1",
                  file_path = "/mnt/c/Users/linde/Nextcloud/inbox/exports/m6A_decay_ms_share/RP_coverage/old_experiment/RPM_coverage/",
                  file_suffix = "_clipped_tr_nohrRNA_un_codon_Asite_offset_RPM_coverage.txt")
## normalize
with(mcm5s2U_riboseq_1, coverage[,meta$run := lapply(.SD, function(x){x/(sum(x[is.finite(x)])/cds_len_cods) }), .SDcols = meta$run, by = .(ensembl_transcript_id, cds_len_cods)])

mcm5s2U_riboseq_1$coverage <-
  copy(mcm5s2U_riboseq_1$coverage)

## metadata
mcm5s2U_riboseq_1$meta[, title := run]
mcm5s2U_riboseq_1$meta[, genotype  := factor(sedit(str_extract(title, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO")) ]
mcm5s2U_riboseq_1$meta[, treatment := factor("CTRL", levels = c("CTRL", "STM")) ]
mcm5s2U_riboseq_1$meta[, replicate := str_extract(title, "[0-9]$") ]


# mcm5s2U_riboseq experiment 2 (STM treatment)
## read in data
mcm5s2U_riboseq_2 <-
  get_tx_coverage(experiment_name = "exp2",
                  file_path = "/mnt/c/Users/linde/Nextcloud/inbox/exports/m6A_decay_ms_share/RP_coverage/new_experiment/RPM_coverage/",
                  file_suffix = "_Mono_clpd_tr_no_r_t_sno_sn_RNA_codon_Asite_offset_RPM_coverage.txt")
## normalize
with(mcm5s2U_riboseq_2, coverage[,meta$run := lapply(.SD, function(x){x/(sum(x[is.finite(x)])/cds_len_cods) }), .SDcols = meta$run, by = .(ensembl_transcript_id, cds_len_cods)])

## metadata
mcm5s2U_riboseq_2$meta[, title := run]
mcm5s2U_riboseq_2$meta[, genotype  := factor(sedit(str_extract(title, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO")) ]
mcm5s2U_riboseq_2$meta[, treatment := factor(str_extract(title, "CTRL|STM"), levels = c("CTRL", "STM")) ]
mcm5s2U_riboseq_2$meta[, replicate := str_extract(title, "[0-9]$") ]


# mcm5s2U_riboseq experiment 3 (disomes)
## read in data
mcm5s2U_riboseq_3 <-
  get_tx_coverage(experiment_name = "exp3",
                  file_path = "/mnt/c/Users/linde/Nextcloud/inbox/exports/m6A_decay_ms_share/RP_coverage/new_experiment/disome_coverage/disome_mergerep_RPM_coverage/",
                  file_suffix = "_allrep_codon_Asite_offset_RPM_coverage.txt")

setnames(mcm5s2U_riboseq_3$coverage, mcm5s2U_riboseq_3$meta$run, paste0("disome_", mcm5s2U_riboseq_3$meta$run) )
mcm5s2U_riboseq_3$meta[, run := paste0("disome_", run)]

## normalize
with(mcm5s2U_riboseq_3, coverage[,meta$run := lapply(.SD, function(x){x/(sum(x[is.finite(x)])/cds_len_cods) }), .SDcols = meta$run, by = .(ensembl_transcript_id, cds_len_cods)])

## metadata
mcm5s2U_riboseq_3$meta[, title := run]
mcm5s2U_riboseq_3$meta[, genotype  := factor(sedit(str_extract(title, "WT|elp|ctu"), from = c("ctu", "elp"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO")) ]
mcm5s2U_riboseq_3$meta[, treatment := factor(toupper(str_extract(title, "Ctrl|STM")), levels = c("CTRL", "STM")) ]
mcm5s2U_riboseq_3$meta[, replicate := 0]

#
# combine experiments
#

mcm5s2U_riboseq <-
  list(
    coverage = Reduce(f = function(x,y){merge(x,y,c("ensembl_transcript_id", "cds_len_cods", "codon_rank"), all = TRUE)},
       x = list(mcm5s2U_riboseq_1$coverage, mcm5s2U_riboseq_2$coverage, mcm5s2U_riboseq_3$coverage)),
    meta = rbind(mcm5s2U_riboseq_1$meta, mcm5s2U_riboseq_2$meta, mcm5s2U_riboseq_3$meta)
  )

setkeyv(mcm5s2U_riboseq$meta, "run")

rm("mcm5s2U_riboseq_1", "mcm5s2U_riboseq_2", "mcm5s2U_riboseq_3")
gc()

head(mcm5s2U_riboseq$coverage)

mcm5s2U_riboseq$meta

```

```{r lift coverage to codons}

mcols(cods.gr) = mcols(cods.gr)[! names(mcols(cods.gr)) %in% mcm5s2U_riboseq$meta$run]

mcols(cods.gr) <-
  cbind(mcols(cods.gr), as.data.table(cods.gr)[
    , codonID := paste0(gsub("\\..*$", "", seqnames), "_",codon_rank)
  ][][
    , merge(.SD, mcm5s2U_riboseq$coverage[, codonID := paste0(ensembl_transcript_id, "_", codon_rank)][
      , c("codonID", mcm5s2U_riboseq$meta$run)
      , with = FALSE
    ], "codonID", all.x = TRUE, sort = FALSE) 
  ][
    , mcm5s2U_riboseq$meta$run
    , with = FALSE
  ])

## set NA coverage to 0
for(run in mcm5s2U_riboseq$meta$run) {
  mcols(cods.gr)[run][which(is.na(mcols(cods.gr)[run])),] = 0
}

cods.gr

## remove coverage data to free up RAM
mcm5s2U_riboseq$coverage = NULL
gc()

```

```{r, fig.width = 12}

as.data.table(cods.gr)[
  , region := ifelse(codon_rank < 20, "start", ifelse(codon_rank_from_stop > -20, "stop", NA) )
][
  !is.na(region)
][
  , melt(.SD, id.vars=c( c("region", "codon_rank", "codon_rank_from_stop"), "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = c( c("region", "codon_rank", "codon_rank_from_stop"), "run")
][
  , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  genotype == "WT"
][
  , ggplot(.SD, 
           aes(x = ifelse(region == "start", codon_rank, codon_rank_from_stop), y = occupancy, col = treatment )) +
    stat_summary(fun.data = mean_se, size = 0.5) +
    stat_summary(fun.data = mean_se, geom = "line") +
    scale_color_manual(values = palval_m6A_inh) +
    facet_nested(experiment~region, scales = "free")
]

```


```{r, fig.width = 12}

as.data.table(cods.gr)[
  , region := ifelse(codon_rank < 20, "start", ifelse(codon_rank_from_stop > -20, "stop", NA) )
][
  !is.na(region)
][
  , melt(.SD, id.vars=c( c("region", "codon_rank", "codon_rank_from_stop"), "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = c( c("region", "codon_rank", "codon_rank_from_stop"), "run")
][
  , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  , ggplot(.SD, 
           aes(x = ifelse(region == "start", codon_rank, codon_rank_from_stop), y = occupancy, col = genotype )) +
    stat_summary(fun.data = mean_se, size = 0.5) +
    stat_summary(fun.data = mean_se, geom = "line") +
    scale_color_manual(values = palval_genotype) +
    facet_nested(experiment + treatment ~ region, scales = "free")
]

```

```{r}

tx_has_m6A <-
  as.data.table(cods.gr)[
    (codon_rank_from_stop > -20) & glori_m6A_lvl > 0.5, unique(seqnames)
  ]


tx_has_no_m6A <-
as.data.table(cods.gr)[
  , has_no_m6A := (codon_rank_from_stop > -20) & sum(glori_m6A_lvl) < 0.1,  by = .(seqnames)
][(has_no_m6A), unique(seqnames)]


```


```{r, fig.width = 12, fig.height = 9}

as.data.table(cods.gr)[
  , prestop_m6A_status := ifelse(seqnames %in% tx_has_m6A, "has m6A", ifelse(seqnames %in% tx_has_no_m6A, "has no m6A", NA))
][
  !is.na(prestop_m6A_status)
][
  , region := ifelse(codon_rank < 30, "start", ifelse(codon_rank_from_stop > -30, "stop", NA) )
][
  region == "stop"
][
  , melt(.SD, id.vars=c( c("region", "prestop_m6A_status", "codon_rank", "codon_rank_from_stop"), "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = c( c("region", "prestop_m6A_status", "codon_rank", "codon_rank_from_stop"), "run")
][
  , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  
][
  , ggplot(.SD, 
           aes(x = ifelse(region == "start", codon_rank, codon_rank_from_stop), y = occupancy, col = genotype )) +
    stat_summary(fun.data = mean_se, size = 0.5) +
    stat_summary(fun.data = mean_se, geom = "line") +
    scale_color_manual(values = palval_genotype) +
    facet_nested(experiment  ~  prestop_m6A_status + treatment, scales = "free")
]

```


```{r, fig.width = 12, fig.height = 9}

as.data.table(cods.gr)[
  , prestop_m6A_status := ifelse(seqnames %in% tx_has_m6A, "has m6A", ifelse(seqnames %in% tx_has_no_m6A, "has no m6A", NA))
][
  !is.na(prestop_m6A_status)
][
  , region := ifelse(codon_rank < 30, "start", ifelse(codon_rank_from_stop > -30, "stop", NA) )
][
  region == "stop"
][
  , melt(.SD, id.vars=c( c("region", "prestop_m6A_status", "codon_rank", "codon_rank_from_stop"), "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = c( c("region", "prestop_m6A_status", "codon_rank", "codon_rank_from_stop"), "run")
][
  , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  codon_rank_from_stop < 0 
][
  , ggplot(.SD, 
           aes(x = ifelse(region == "start", codon_rank, codon_rank_from_stop), y = occupancy, col = genotype )) +
    stat_summary(fun.data = mean_se, size = 0.5) +
    stat_summary(fun.data = mean_se, geom = "line") +
    scale_color_manual(values = palval_genotype) +
    facet_nested(experiment  ~  prestop_m6A_status + treatment, scales = "free")
]

```


```{r, fig.width = 12, fig.height = 9}

as.data.table(cods.gr)[
 , is_expressed := sum(cvg, na.rm = TRUE) > 1
 , by = .(seqnames)
][
 (is_expressed)
][
 , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
 , merge(.SD, canonical_transcripts )
][
  , prestop_m6A_status := ifelse(seqnames %in% tx_has_m6A, "has m6A", ifelse(seqnames %in% tx_has_no_m6A, "has no m6A", NA))
][
  !is.na(prestop_m6A_status)
][
  , region := ifelse(codon_rank < 30, "start", ifelse(codon_rank_from_stop > -30, "stop", NA) )
][
  region == "stop"
][
  , melt(.SD, id.vars=c( c("region", "prestop_m6A_status", "codon_rank", "codon_rank_from_stop"), "glori_m6A_lvl"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
  , c(setNames(as.list(mean_se(occupancy)), c("occupancy", "occ_min", "occ_max")), N = .N, glori_m6A_lvl = mean(glori_m6A_lvl))
  , by = c( c("region", "prestop_m6A_status", "codon_rank", "codon_rank_from_stop"), "run")
][
  , c("experiment", "genotype", "treatment", "replicate")  := mcm5s2U_riboseq$meta[run, .(experiment, genotype, treatment, replicate) ]
][
  codon_rank_from_stop < 0 
][
  , ggplot(.SD, 
           aes(x = ifelse(region == "start", codon_rank, codon_rank_from_stop), y = occupancy, col = genotype )) +
    geom_smooth() +
    stat_summary(fun.data = mean_se, size = 1, geom = "point") +
    #stat_summary(fun.data = mean_se, geom = "line") +
    scale_color_manual(values = palval_genotype) +
    facet_nested(experiment  ~  prestop_m6A_status + treatment)
]

```

```{r}
as.data.table(cods.gr)[
 , .(cvg = sum(cvg, na.rm = TRUE))
 , by = .(seqnames)
][
  , prestop_m6A_status := ifelse(seqnames %in% tx_has_m6A, "has m6A", ifelse(seqnames %in% tx_has_no_m6A, "has no m6A", NA))
][
 , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
 , merge(.SD, canonical_transcripts )
][
  , ggplot(.SD, aes(x = prestop_m6A_status, y = cvg )) +
    geom_boxplot() +
    scale_y_continuous(trans = "log10") 
]
```

Check for differing CDS length annotations:
```{r}

non_matching_tx <-
  as.data.table(cods.gr)[
    , .(cds_len_cods  = max(codon_rank))
    , by = .(ensembl_transcript_id = gsub("\\..*$", "", seqnames))
  ][
    , merge(.SD, unique(mcm5s2U_riboseq$coverage[, .(ensembl_transcript_id,cds_len_cods)]), "ensembl_transcript_id")
  ][
    cds_len_cods.x != cds_len_cods.y
  ]

non_matching_tx 

```
So 84 transcripts have different lengths... However, for the rest it should be fine to transfer the coverages to the GRanges object



# Analysis

## Codon averages
```{r}
## codon strength according to Grosjean and Westhoff 2016


## weak
paste0("TT", c("T","C","G","A"))
paste0("TA", c("T","C","G","A"))
paste0("AT", c("T","C","G","A"))
paste0("AA", c("T","C","G","A"))

## intermediate
paste0("AG", c("T","C","G","A"))
paste0("AC", c("T","C","G","A"))
paste0("TG", c("T","C","G","A"))
paste0("TC", c("T","C","G","A"))
paste0("GA", c("T","C","G","A"))
paste0("GT", c("T","C","G","A"))
paste0("CA", c("T","C","G","A"))
paste0("CT", c("T","C","G","A"))

## strong
paste0("GG", c("T","C","G","A"))
paste0("GC", c("T","C","G","A"))
paste0("CG", c("T","C","G","A"))
paste0("CC", c("T","C","G","A"))



```

```{r}

as.data.table(cods.gr)[
  , gloribin := cut2(glori_m6A_lvl,cuts = c(0.1, 0.25, 0.5, 0.750))
][
  , .(occupancy = mean(WT_CTRL_rep1, na.rm = TRUE), .N)
  , by = .(codon, gloribin) 
][
  order(N)
]

```


```{r, fig.width = 8, fig.height = 2}

as.data.table(cods.gr)[
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , .(occupancy = mean(WT_CTRL_rep1, na.rm = TRUE), .N)
  , by = .(codon, gloribin) 
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
  , ggplot(.SD, aes(x = codon, y = occupancy)) +
    geom_point(aes(col = gloribin)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    NULL
]

```

```{r, fig.width = 28, fig.height = 2}

as.data.table(cods.gr)[
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon, gloribin, run) 
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  
][
  , ggplot(.SD, aes(x = codon, y = occupancy)) +
    geom_point(aes(col = gloribin, shape = replicate)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    facet_nested(.~ genotype + treatment) +
    NULL
]

```


```{r, fig.width = 8, fig.height = 2}

as.data.table(cods.gr)[
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon, gloribin, run) 
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  genotype == "WT" & treatment == "CTRL"
][
  , ggplot(.SD, aes(x = codon, y = occupancy)) +
    geom_point(aes(col = gloribin, shape = replicate)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    facet_nested(.~ genotype + treatment) +
    NULL
]

```


```{r, fig.width = 8, fig.height = 2}

as.data.table(cods.gr)[
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon, gloribin, run) 
][
  ! codon %in% stop_codons &
    codon %in% drach_codons &
   rep(T,.N)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  genotype == "WT" & treatment == "CTRL"
][
  , ggplot(.SD, aes(x = gloribin, y = occupancy)) +
    geom_point(aes(col = gloribin, shape = replicate)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    # geom_text(data = function(x)x,
    #           aes(label=N), inherit.aes=FALSE,
    #           x=-Inf,y=Inf, check_overlap = TRUE, hjust = 0, vjust = 1.3) +
    facet_nested(.~ genotype + treatment + codon) +
    NULL
]

```

```{r, fig.width = 8, fig.height = 2}

as.data.table(cods.gr)[
  codon %in% drach_codons
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon, gloribin, run) 
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  
][
  , ggplot(.SD, aes(x = codon, y = occupancy)) +
    geom_point(aes(col = gloribin, shape = replicate)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    facet_nested(.~ genotype + treatment) +
    NULL
]

```


```{r}

as.data.table(cods.gr)[
  , codon_type := get_codon_type(codon)
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.2)) )
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
  , melt(.SD, id.vars=c("codon_type", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon_type, gloribin, run) 
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , ggplot(.SD, aes(x = codon_type, y = occupancy)) +
    geom_point(aes(col = gloribin, shape = replicate)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    facet_nested(.~ genotype + treatment) +
    NULL
]

```

```{r}
as.data.table(cods.gr)[
  , codon_type := get_codon_type(codon)
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.2)) )
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
  , melt(.SD, id.vars=c("codon_type", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon_type, gloribin, run) 
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , ggplot(.SD, aes(x = codon_type, y = occupancy)) +
    geom_point(aes(col = gloribin, shape = replicate)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    facet_nested(.~ genotype + treatment) +
    NULL
]
```


```{r, fig.width = 12, fig.height = 2}

as.data.table(cods.gr)[
  codon %in% c(drach_codons, "ACT", "CAA")
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value, na.rm = TRUE), .N)
  , by = .(codon, gloribin, run) 
][
  ! codon %in% stop_codons &
   rep(T,.N)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  
][
  , ggplot(.SD, aes(x = gloribin, y = occupancy)) +
    #geom_point(aes(col = genotype, shape = replicate)) +
    stat_summary(fun.data = mean_se, aes(col = genotype)) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = palval_genotype ) +
    facet_nested(.~ codon + treatment, scales = "free_y" ) +
    NULL
]

```



```{r, fig.width = 18, fig.height = 2}

require("ggpmisc")

as.data.table(cods.gr)[
  , gloribin := cut2(glori_m6A_lvl,cuts = seq(0,1,0.1))
][
  codon %in% drach_codons & 
    ! codon %in% stop_codons &
    rep(T,.N)
][
  , melt(.SD, id.vars=c("codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run", value.name = "occupancy")
][
   #! (glori_m6A_lvl == 0 & occupancy > 10) & 
    occupancy > 0 & occupancy < 100 &
    rep(T,.N)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  genotype == "WT" & treatment == "CTRL"
][
  , ggplot(.SD, aes(x = glori_m6A_lvl , y = occupancy )) +
    geom_point(aes(col = gloribin), alpha = 0.5) +
    #geom_quasirandom(aes(col = gloribin), alpha = 0.2) +
    #stat_summary(fun.data = mean_se, aes(col = gloribin)) +
    #geom_smooth(method = "lm") +
    #stat_cor(size = 4, label.y = Inf, vjust = 1.3) +
    stat_poly_line() +
    stat_poly_eq(use_label("eq"), size = 3, label.y = Inf, vjust = 1.3) +
    stat_poly_eq(use_label("R2"), size = 3, label.y = Inf, vjust = 2.6) +
    theme_bw() +
    #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(levels(.SD$gloribin)))) +
    facet_nested_wrap(~codon, nrow = 1) +
    coord_cartesian(ylim = c(0, 50)) +
    theme(legend.position = "none") +
    labs(x = "m6A stoichiometry", y = "A-site occupancy") +
    NULL
]


```

```{r, fig.width = 8, fig.height = 2}

as.data.table(cods.gr)[
  , gloribin := cut2(glori_m6A_lvl,cuts = seq(0,1,0.2))
][
  , occupancy := rowMeans(.SD), .SDcols = c("WT_CTRL_rep1", "WT_CTRL_rep2")
][
  codon %in% c("AAA", "AGA", "GAA", "GGA", "CAA") & 
    #glori_m6A_lvl > 0 &
    #occupancy > 0 &
    occupancy < 20 &
    rep(T,.N)
][
  , ggplot(.SD, aes(x = gloribin, y = occupancy )) +
    stat_summary(fun.data = mean_se, aes(col = gloribin)) +
    #geom_smooth(method = "lm") +
    #stat_cor() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, family = "Courier") ) +
    scale_color_manual(values = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(unique(.SD$gloribin)))) +
    #geom_text(data = .SD[, .N, by = .(codon)], aes(label = paste0("N = ", N) ), x = Inf, y = Inf, vjust = 3.3, hjust = 1) +
    facet_nested(~codon) +
    NULL
]

```

## Pausing by cumsum

Idea: show pausing across genes by cumulative coverage plots

1. Morscher, R.J., Ducker, G.S., Li, S.H.-J., Mayer, J.A., Gitai, Z., Sperl, W., and Rabinowitz, J.D. (2018). Mitochondrial translation requires folate-dependent tRNA methylation. Nature 554, 128–132. https://doi.org/10.1038/nature25460.

Figure 3 b


```{r, fig.width = 16, fig.height = 18}

as.data.table(cods.gr)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
  hgnc_symbol %in% c("MYC", "YTHDF2", "ACTB", "JUN", "EEF2", "ZNF217")
][
#  ensembl_transcript_id %in% c("ENST00000322157", "ENST00000366955", "ENST00000368232")
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("hgnc_symbol", "codon_rank", "codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  order(run, hgnc_symbol, codon_rank)
  , cumocc :=  cumsum(value)
  , by = .(hgnc_symbol, run)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , ggplot(.SD, aes(x = codon_rank, y = cumocc )) +
    
    ## m6A sites
    geom_vline(data = function(x)x[ glori_m6A_lvl > 0.25, ], aes(xintercept = codon_rank, col = "m6A"), linetype = 2, linewidth = 0.25) +
    scale_color_manual(name="Codon m6A", values = c("m6A" = "salmon")) +
    new_scale_color() +
    
    ## Codons
    geom_point(data = function(x)x[ codon %in% c("GAA"), ], aes(col = codon), y = 0) +
    geom_point(data = function(x)x[ glori_m6A_lvl > 0.25, ], aes(col = codon), y = 0) +
    scale_color_discrete(name = "Codon type") +
    new_scale_color() +
    
    ## Occupancy
    geom_line(aes(col = treatment)) +
    #ylim(0,1) +
    scale_color_manual(name = "Treatment", values = c("CTRL" = "grey40", palval_m6A_inh) ) +
    new_scale_color() +
    
    ## GAA
    
    
    ## Style
    #geom_point(data = function(x)x[ codon %in% c("AGA", "GAA", "AAA") & glori_m6A_lvl > 0.5], y = 0, aes(col = codon)) +
    #geom_point(data = function(x)x[ glori_m6A_lvl > 0.5], y = 0, col = "red") +
    theme_bw() +
    facet_nested_wrap(hgnc_symbol~genotype + replicate, scales = "free", nrow = length(unique(.SD$hgnc_symbol))) +
    scale_y_continuous(name = "Cumulative ribosome occupancy", expand = expansion(c(0.025, 0.05))) +
    labs(x = "Codon position") +
    NULL
]

```


Targets from Rachel Green's emetine paper:
```{r, fig.width = 16, fig.height = 18}

as.data.table(cods.gr)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
  hgnc_symbol %in% c("JUN", "DDIT4", "DUSP1", "TXNIP", "ATF3", "ZNF574")
][
#  ensembl_transcript_id %in% c("ENST00000322157", "ENST00000366955", "ENST00000368232")
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("hgnc_symbol", "codon_rank", "codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  order(run, hgnc_symbol, codon_rank)
  , cumocc :=  cumsum(value)
  , by = .(hgnc_symbol, run)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , ggplot(.SD, aes(x = codon_rank, y = cumocc )) +
    
    ## m6A sites
    geom_vline(data = function(x)x[ glori_m6A_lvl > 0.25, ], aes(xintercept = codon_rank, col = "m6A"), linetype = 2, linewidth = 0.25) +
    scale_color_manual(name="Codon m6A", values = c("m6A" = "salmon")) +
    new_scale_color() +
    
    ## Codons
    geom_point(data = function(x)x[ codon %in% c("GAA") | glori_m6A_lvl > 0.25, ], aes(col = codon), y = 0) +
    #geom_point(data = function(x)x[ codon %in% c("GAA") | glori_m6A_lvl > 0.25 & codon %in% mcm5s2U_codons, ], col = "red", y = 0, shape = 21) +
    scale_color_discrete(name = "Codon type") +
    new_scale_color() +
    
    ## Occupancy
    geom_line(aes(col = treatment)) +
    #ylim(0,1) +
    scale_color_manual(name = "Treatment", values = c("CTRL" = "grey40", palval_m6A_inh) ) +
    new_scale_color() +
    
    ## GAA
    
    
    ## Style
    #geom_point(data = function(x)x[ codon %in% c("AGA", "GAA", "AAA") & glori_m6A_lvl > 0.5], y = 0, aes(col = codon)) +
    #geom_point(data = function(x)x[ glori_m6A_lvl > 0.5], y = 0, col = "red") +
    theme_bw() +
    facet_nested_wrap(hgnc_symbol~genotype + replicate, scales = "free", nrow = length(unique(.SD$hgnc_symbol))) +
    scale_y_continuous(name = "Cumulative ribosome occupancy", expand = expansion(c(0.025, 0.05))) +
    labs(x = "Codon position") +
    NULL
]

```



```{r, fig.width = 16, fig.height = 18}

as.data.table(cods.gr)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
  hgnc_symbol %in% c("JUN", "DDIT4", "DUSP1", "TXNIP", "ATF3", "ZNF574")
][
#  ensembl_transcript_id %in% c("ENST00000322157", "ENST00000366955", "ENST00000368232")
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("hgnc_symbol", "codon_rank", "codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  order(run, hgnc_symbol, codon_rank)
  , cumocc :=  cumsum(value)
  , by = .(hgnc_symbol, run)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , codon_type := get_codon_type(codon)
][
  , ggplot(.SD, aes(x = codon_rank, y = cumocc )) +
    
    ## m6A sites
    geom_vline(data = function(x)x[ glori_m6A_lvl > 0.25, ], aes(xintercept = codon_rank, col = "m6A"), linetype = 2, linewidth = 0.25) +
    scale_color_manual(name="Codon m6A", values = c("m6A" = "salmon")) +
    new_scale_color() +
    
    ## Codons
    geom_point(data = function(x)x[ codon %in% c("GAA") | glori_m6A_lvl > 0.25, ], aes(col = codon_type), y = 0) +
    #geom_point(data = function(x)x[ codon %in% c("GAA") | glori_m6A_lvl > 0.25 & codon %in% mcm5s2U_codons, ], col = "red", y = 0, shape = 21) +
    scale_color_manual(name = "Codon type", values = palval_codon_type ) +
    new_scale_color() +
    
    ## Occupancy
    geom_line(aes(col = treatment)) +
    #ylim(0,1) +
    scale_color_manual(name = "Treatment", values = c("CTRL" = "grey40", palval_m6A_inh) ) +
    new_scale_color() +
    
    ## GAA
    
    
    ## Style
    #geom_point(data = function(x)x[ codon %in% c("AGA", "GAA", "AAA") & glori_m6A_lvl > 0.5], y = 0, aes(col = codon)) +
    #geom_point(data = function(x)x[ glori_m6A_lvl > 0.5], y = 0, col = "red") +
    theme_bw() +
    facet_nested_wrap(hgnc_symbol~genotype + replicate, scales = "free", nrow = length(unique(.SD$hgnc_symbol))) +
    scale_y_continuous(name = "Cumulative ribosome occupancy", expand = expansion(c(0.025, 0.05))) +
    labs(x = "Codon position") +
    NULL
]

```



```{r}
### todo: window around m6A codons
as.data.table(cods.gr)[
  codonID %in% as.data.table(cods.gr)[glori_m6A_lvl > 0.5, paste(seqnames, codon_rank + c(-20:20), sep = "_" )]
][
  , melt(.SD, id.vars=c("codon_rank"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value))
  , by = .(codon_rank, run)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  
]

```


```{r}
as.data.table(cods.gr)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
  hgnc_symbol %in% c("MYC", "YTHDF2", "ACTB", "JUN", "EEF2", "ZNF217")
][
#  ensembl_transcript_id %in% c("ENST00000322157", "ENST00000366955", "ENST00000368232")
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("hgnc_symbol", "codon_rank", "codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  order(run, hgnc_symbol, codon_rank)
  , .(sumocc =  sum(value))
  , by = .(gene = hgnc_symbol, run)
][
  , dcast(.SD, gene ~ run, value.var = "sumocc")
]
```


```{r}

as.data.table(cods.gr)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
  hgnc_symbol %in% c("JUN", "DDIT4", "DUSP1", "TXNIP", "ATF3", "ZNF574")
][
#  ensembl_transcript_id %in% c("ENST00000322157", "ENST00000366955", "ENST00000368232")
][
  , gloribin := ifelse(glori_m6A_lvl == 0, "0", cut2(glori_m6A_lvl,cuts = seq(0,1,0.25)) )
][
  , melt(.SD, id.vars=c("hgnc_symbol", "codon_rank", "codon", "glori_m6A_lvl", "gloribin"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , codon_type := ifelse(codon %in% mcm5_codons, "mcm5", ifelse(codon %in% mcm5s2U_codons, "mcm5s2U", "other"))
][
  , ggplot(.SD, aes(x = codon_rank, y = cumocc )) +
    
    ## m6A sites
    geom_vline(data = function(x)x[ glori_m6A_lvl > 0.25, ], aes(xintercept = codon_rank, col = "m6A"), linetype = 2, linewidth = 0.25) +
    scale_color_manual(name="Codon m6A", values = c("m6A" = "salmon")) +
    new_scale_color() +
    
    ## Codons
    geom_point(data = function(x)x[ codon %in% c("GAA") | glori_m6A_lvl > 0.25, ], aes(col = codon_type), y = 0) +
    #geom_point(data = function(x)x[ codon %in% c("GAA") | glori_m6A_lvl > 0.25 & codon %in% mcm5s2U_codons, ], col = "red", y = 0, shape = 21) +
    scale_color_manual(name = "Codon type", values = palval_codon_type ) +
    new_scale_color() +
    
    ## Occupancy
    geom_line(aes(col = treatment)) +
    #ylim(0,1) +
    scale_color_manual(name = "Treatment", values = c("CTRL" = "grey40", palval_m6A_inh) ) +
    new_scale_color() +
    
    ## GAA
    
    
    ## Style
    #geom_point(data = function(x)x[ codon %in% c("AGA", "GAA", "AAA") & glori_m6A_lvl > 0.5], y = 0, aes(col = codon)) +
    #geom_point(data = function(x)x[ glori_m6A_lvl > 0.5], y = 0, col = "red") +
    theme_bw() +
    facet_nested_wrap(hgnc_symbol~genotype + replicate, scales = "free", nrow = length(unique(.SD$hgnc_symbol))) +
    scale_y_continuous(name = "Cumulative ribosome occupancy", expand = expansion(c(0.025, 0.05))) +
    labs(x = "Codon position") +
    NULL
]
```





```{r, fig.width = 18, fig.height = 3}
as.data.table(cods.gr)[
  , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
 # hgnc_symbol %in% c("MYC", "YTHDF2", "ACTB", "JUN")
][
  , melt(.SD, id.vars=c("hgnc_symbol"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  value > 0
][
  , .(occupancy = mean(value))
  , by = .(gene = hgnc_symbol, run)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , merge(.SD, exlenTabGene)
][
  , ggplot(.SD, aes(x = exbin, y = occupancy)) +
    geom_boxplot(aes(fill = treatment)) +
    facet_nested(~genotype + replicate) +
    ylim(NA, 300)
]
```


```{r, fig.width = 18, fig.height = 3}
as.data.table(cods.gr)[
  codon_rank != max(codon_rank) | codon_rank != 1
  , exclude_codon := TRUE
  , by = .(seqnames)
][
  (exclude_codon) , ensembl_transcript_id := gsub("\\..*$", "", seqnames)
][
  , merge(.SD, canonical_transcripts )
][
 # hgnc_symbol %in% c("MYC", "YTHDF2", "ACTB", "JUN")
][
  , melt(.SD, id.vars=c("relpos.cds"), measure.vars = mcm5s2U_riboseq$meta$run, variable.name = "run")
][
  , .(occupancy = mean(value))
  , by = .(relpos.cds = cut2(relpos.cds, seq(1,2,0.1)), run)
][
 , genotype  := factor(sedit(str_extract(run, "WT|ELP1|CTU2"), from = c("CTU2", "ELP1"), to = c("CTU2 KO", "ELP1 KO")), levels = c("WT", "ELP1 KO", "CTU2 KO"))
][
  , treatment := factor(str_extract(run, "CTRL|STM"), levels = c("CTRL", "STM"))
][
  , replicate := str_extract(run, "rep[0-9]$")
][
  , ggplot(.SD, aes(x = relpos.cds, y = occupancy)) +
    geom_point(aes(col = treatment)) +
    facet_nested(~genotype + replicate) +
    NULL
  
]
```

```{r}
mcm5s2U_riboseq$meta
```



